### 2019-04-30 doc/bilevel/CommonAbstract.tex
Adding draft documentation for pao.bilevel
I haven't yet confirmed that this compiles.  I'm not expecting we'll
use all of this text here.

We describe new capabilities for modeling bilevel programs within
the Pyomo modeling software.  These capabilities include new modeling
components that represent subproblems, modeling transformations for
re-expressing models with bilevel structure in other forms, and
optimize bilevel programs with meta-solvers that apply transformations
and then perform optimization on the resulting model.  We illustrate
the breadth of Pyomo's modeling capabilities for bilevel programs,
and we describe how Pyomo's meta-solvers can perform local and
global optimization of bilevel programs.

%\keywords{Python \and algebraic modeling language \and bilevel program}

### 2019-04-30 doc/bilevel/main.tex
Adding draft documentation for pao.bilevel
I haven't yet confirmed that this compiles.  I'm not expecting we'll
use all of this text here.

%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
% 	DO NOT MODIFY HERE, SCROLL DOWN
%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -

\documentclass[a4paper,11pt]{article}
\usepackage[latin1]{inputenc}
\usepackage[english]{babel}
\usepackage{graphicx,amsmath,amsfonts,amssymb,a4wide}

\usepackage{listings}
\usepackage{color}
\include{macros}

\newenvironment{vuoto}{}{}
\newenvironment{abst}{
\textbf{Abstract}\begin{vuoto}}{
\end{vuoto}}

\newenvironment{keyw}{
\textbf{Keywords:}\begin{vuoto}}{
\end{vuoto}}

\newcommand{\papertitle}[1]{%
\begin{center}
\LARGE \textbf{#1} \\
\vspace*{0.5cm}
}
\newcommand{\session}[1]{%
\newpage
\begin{center}
%\Large 
\textit{#1} \\
\end{center}
\vspace*{0.5cm}
}
\newcommand{\presentername}[1]{%
\normalsize
\textbf{#1}$^*$ \\
}
\newcommand{\presenteraffiliation}[2]{%
\normalsize \emph{#1}, \texttt{#2} \\
\vspace*{1ex} 
}
\newcommand{\coauthorname}[1]{%
\normalsize
\textbf{#1} \\
}
\newcommand{\coauthoraffiliation}[2]{%
\normalsize \emph{#1}, \texttt{#2} \\
\vspace*{1ex}
}
\newcommand{\paperabstract}[1]{%
\setcounter{figure}{0}
\end{center}
\begin{abst}
\normalsize #1
\end{abst}
\vspace*{1ex}
}

\newcommand{\paperkeywords}[1]{%
\begin{keyw}
\normalsize \emph{#1}
\end{keyw}
\vspace*{1ex}
}

\newenvironment{papertext}{}

\newcommand{\vref}[1]{
\item #1
}
\newcommand{\paperreferences}[1]{%
~\\
\noindent \large \textbf{References}
\renewcommand{\theenumi}{[\arabic{enumi}]}
\renewcommand{\labelenumi}{\theenumi}
\begin{enumerate}
\normalsize
#1
\end{enumerate}
\newpage
}

\begin{document}

%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
% 		MODIFY HERE
%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -

%------- Session of the talk (invited or contributed)
\session{Contributed session}
%------- Title of the talk
\papertitle{Modeling Multilevel Programs in Pyomo}
%------- Speaker
\presentername{William E. Hart\index{Hart, William E.|pp}}
\presenteraffiliation{Center for Computer Research, Sandia National Laboratories}{wehart@sandia.gov}
%------- Coauthor 1
\coauthorname{Bryan Arguello\index{Arguello, Bryan|ca}}
\presenteraffiliation{Energy and Geoscience Center, Sandia National Laboratories}{barguel@sandia.gov}
\coauthorname{John Siirola\index{Siirola, John|ca}}
\presenteraffiliation{Center for Computer Research, Sandia National Laboratories}{jdsiiro@sandia.gov}
\coauthorname{Bismark Singh\index{Singh, Bismark|ca}}
\presenteraffiliation{Center for Computer Research, Sandia National Laboratories}{bsingh@sandia.gov}
\coauthorname{Jean-Paul Watson\index{Watson, Jean-Pual|ca}}
\presenteraffiliation{Center for Computer Research, Sandia National Laboratories}{jwatson@sandia.gov}

%------- Abstract (max 3 pages)
\paperabstract{
We describe capabilities for modeling bilevel and multilevel programs
within the Pyomo modeling software.  These capabilities include
modeling components that represent subproblems, modeling transformations
for re-expressing models with bilevel structure in other forms, and
optimize bilevel programs with meta-solvers that apply transformations
and then perform optimization on the resulting model.  We illustrate
the application of Pyomo to model powergrid reliability problems.
}
%This is an example of the short abstract template (max 3 pages) and how to cite (max 3) references \ref{CY59}, \ref{CP61}.

\paperkeywords{algebraic modeling languages, reformulations, powergrid security}

\section{Introduction}

Although multilevel problems arise in many applications, few algebraic
modeling languages (AML) have integrated capabilities for expressing
these problems.  AMLs are high-level programming languages for
describing and solving mathematical problems, particularly
optimization-related problems.  AMLs provide a mechanism
for defining variables and generating constraints with a concise
mathematical representation, which is essential for large-scale,
real-world problems that involve thousands or millions of constraints
and variables.  GAMS, YALMIP and Pyomo
provide explicit support for modeling bilevel programs.  A variety
of other AMLs support the solution of bilevel programs through the
expression of Karush-Kuhn-Tucker conditions and associated
reformulations using mixed-complementarity conditions, but these
reformulations must be expressed by the user in these AMLs.

The main point of this paper is to describe modeling techniques in
Pyomo that can express complex multilevel optimization problems.
Pyomo is an open-source software package that supports the definition
and solution of optimization applications using the Python
language~\ref{PyomoBook17}.  Python is a powerful programming
language that has a clear, readable syntax and intuitive object
orientation.  Pyomo uses an object-oriented approach for defining
models that contain decision variables, objectives, and constraints.

Multilevel models can be easily expressed with Pyomo modeling
components for submodels, which can be nested in a general manner.
Further, Pyomo's object-oriented design naturally supports the
ability to automate the reformulation of multilevel models into
other forms.  In particular, we describe Pyomo's capabilities for
transforming bilevel models for several broad classes of problems,
and we illustrate this capability on a power grid security application.

\section{Modeling Bilevel Programs}

The \code{pyomo.bilevel} package extends Pyomo by defining a new
modeling component: \code{SubModel}.  The \code{SubModel} component
defines a subproblem that represents the lower level decisions in
a bilevel program.  This component is like Pyomo's \code{Block} component;
any components can be added to a \code{SubModel} object.  In general,
a submodel is expected to have an objective, one or more variables
and it may define constraints.

The \code{SubModel} class generalizes the \code{Block} component
by including constructor arguments that denote which variables in
the submodel should be considered fixed or variable.  When expressions
in a submodel refer to variables defined outside of the submodel,
the user needs to indicate whether these are fixed values defined
by an upper-level problem.  Fixed variables are treated as constants
within the submodel, but non-fixed variables are defined by the
current submodel or by a lower-level problem.

Consider the following example:
\begin{equation}
\begin{array}{lll}
\min_{x, y, v}      & x + y + v & \\
\st                 & x + v \geq 1.5 \\
                    & 1 \leq x \leq 2 &\\
                    & 1 \leq v \leq 2 & \\
                    & \max_{y,w}  & x + w\\
                    &           & y + w \leq 2.5 \\
                    &           & 1 \leq y \leq 2\\
                    &           & 1 \leq w \leq 2
\end{array}
\end{equation}\\
The following Pyomo model defines four variables, \code{x}, \code{v}, \code{sub.y} and \code{sub.w}:
\listing{simple.py}{pyomo}{2}{15}

\mbox{}\\
Variables \code{x} and \code{v} are declared in the upper-level
problem, and \code{v} only appears in the upper-level problem.
Variables \code{sub.y} and \code{sub.w} are declared in the submodel.
However, note that the \code{sub.y} variable appears in the upper-level
problem, while the \code{sub.w} variable only appears in the
lower-level problem.

\section{Bilevel Optimization for Power Grid Security and Resiliency}

Expressions in
submodels can be linear or nonlinear, convex or nonconvex, continuous
or discontinuous, and more.  Additionally, submodels can be nested
to an arbitrary degree.  Thus, the range of bilevel programs that
can be expressed with Pyomo is quite broad.  However, the real
challenge is solving these models.

Pyomo's supports the structured transformation of models, and this
capability can be applied to transform bilevel and multi-level
problems into standard forms.  Pyomo's object-oriented design allows
the transformation of sub-models within a formulation.  Additionally,
blocks of constraints can be activated and deactivated, which
facilitates \textit{in place} transformations that do not require
the creation of a separate copy of the original model.  Thus, bilevel
models can be transformed and solved, and the resulting solution
can be easily mapped back into the original model.

We describe how these transformations can be applied to reformulate
a bilevel formulation that is used identify worst-case security
risks for power grid disruptions (e.g. terrorist attacks or vandals).
The bilevel formulation is used to identify critical system components
by finding the most disruptive attacks given limited attacker
resources.  The impact of a disruption is derived from a power flow
model, which could be weighted (e.g. to balance equities in different
regions with different network toplogies).  Power flow is modeled
with a linear DC-OPF formulation, and transformations can be applied to 
reformulate this formulation as an integer program.

We describe how these results are used to support an investment
model that prioritizes mitigation investments (e.g. to mitigate
consequences or reduce likelihood of disruptions).  The critical
components identified in the bilevel model are prioritized for
investments.  Finally, we discuss whether trilevel formulations are
suitable for the investment planning problem.  We describe ``nested
defense'' strategies that can be used to maximize the utility of
migitation over a range of possible investment budgets, and discuss
the need for multi-year investment planning.

%------- References
\paperreferences{
\vref{\label{PyomoBook17} Hart, W. E., C. D. Laird, J.-P. Watson, D. L. Woodruff, G. A. Hackebeil, B. L. Nicholson, and J. D. Siirola, \emph{Pyomo: Optimization Modeling in Python}, Springer, 2017.}

}

\end{document}

### 2019-04-30 doc/bilevel/paper-mpec.tex
Adding draft documentation for pao.bilevel
I haven't yet confirmed that this compiles.  I'm not expecting we'll
use all of this text here.

\lstset{language=Python}
\definecolor{light-gray}{gray}{0.9}
\lstset{backgroundcolor=\color{light-gray}}
\lstset{aboveskip=1em,belowskip=1em,showspaces=false,showstringspaces=false}

\SANDsection{Introduction}

Mathematical Programs with Equilibrium Constraint (MPEC) problems
arise in a large number of applications in engineering and economic
systems~\cite{FerPan97,LouPanRal96,OutKocZow98}.  An MPEC is an
optimization problem that includes equilibrium contraints in the
form of complementarity conditions.  Equilibrium constraints naturally
arise as the solution to an optimization subproblem (e.g. for bilevel
programs), variational inequalities, and
complementarity problems~\cite{HarPan90}.

Since MPEC problems frequently arise in practice, many algebraic
modeling languages (AML) have integrated capabilities for expressing
complementarity conditions~\cite{Mun00}, including AMLs like\linebreak
AIMMS~\citep{AIMMS}, AMPL~\citep{AMPL,FouGay03}, GAMS~\citep{GAMS},
MATLAB~\cite{MATLAB} and YALMIP~\cite{Lof04}.  AMLs are high-level
programming languages for describing
and solving mathematical problems, particularly optimization-related
problems~\citep{Kal04}.  AMLs provide a mechanism for defining
variables and generating constraints with a concise mathematical
representation, which is essential for large-scale, real-world
problems that involve thousands or millions of constraints and
variables.

In this paper, we describe new functionality in Pyomo 4.1 for expressing and
optimizing MPEC models in the Pyomo modeling environment.  Pyomo is an
open-source software package that supports
the definition and solution of optimization applications using the
Python language~\cite{pyomotrac,pyomoweb,HarWatWoo11,HarLaiWatWoo12}.  Python is a powerful 
programming language that has a clear, readable syntax and
intuitive object orientation.  Pyomo uses an object-oriented approach
for defining models that contain decision variables, objectives,
and constraints.  MPEC models can be easily expressed with Pyomo
modeling components for complementarity conditions.  Further, Pyomo's
object-oriented design naturally supports the ability to automate
the reformulation of MPEC models into other forms (e.g. disjunctive
programs).  We describe Pyomo meta-solvers that transform MPECs as
MIP or NLP problems, which are then optimized with standard solvers.  Further,
we describe interfaces to specialized mixed complementarity problem
solvers, which
solve MPEC problems expressed without an optimization objective.

The remainder of this paper is organized as follows.
Section~\ref{sec:design} describes how Pyomo supports modeling of
equilibrium constraints as mixed-complementarity conditions.
Section~\ref{sec:transformations} describes transformation capabilities
that automate the reformulation of MPEC models.  Section~\ref{sec:solvers}
describes meta-solvers in Pyomo that leverage these transformations
to support global and local optimization of MPEC problems.
Section~\ref{sec:discussion}
describes future work that is planned for Pyomo.

\SANDsection{Modeling Equilibrium Conditions}
\label{sec:design}

\subsection{Complementarity Conditions}

\citet{FerFouGay99} note that there are a few fundamental forms
that account for a wide range of complementarity conditions that
arise in practice.  Consider a variable $x$ and function $g(x)$.  The classical form of 
complementarity condition can be expressed as
\[
x \geq 0  \;\bot\;  g(x) \geq 0, 
\]
which expresses the complementarity restriction that at least one of these 
must hold with equality.
When the variable $x$ is bounded such that $x \in [l, u]$, then 
a \textit{mixed complementarity condition} can be expressed as
\[
l \leq x \leq u  \;\bot\;  g(x),
\]
which expresses the complementarity restriction that at least one of the following
must hold:
\[
\begin{array}{llll}
& x = l & \text{and} & g(x) \geq 0,\\
& x = u & \text{and} & g(x) \leq 0,\\
\text{or} & l < x < u & \text{and} & g(x) = 0.
\end{array}
\]

These forms can be generalized by substituting a function $f(x)$ for the variable $x$.  Thus,
a \textit{generalized mixed complementarity condition} can be expressed as
\[
l \leq f(x) \leq u  \;\bot\;  g(x),
\]
which expresses the complementarity restriction that at least one of the following
must hold:
\[
\begin{array}{llll}
& f(x) = l & \text{and} & g(x) \geq 0,\\
& f(x) = u & \text{and} & g(x) \leq 0,\\
\text{or} & l < f(x) < u & \text{and} & g(x) = 0.
\end{array}
\]
For completeness, note that the complementarity condition
\[
f(x)  \;\bot\;  g(x) = 0
\]
is a special case where the function $f(x)$ is unbounded.

\subsection{Complementarity Expressions}

The design of complementarity conditions in Pyomo relies on the
specification of Pyomo constraint expressions.  A Pyomo constraint
expression defines an equality, a simple inequality, or a pair of
inequalities.  For example:
\[
\begin{array}{l}
{expr}_1 = {expr}_2\\
{expr}_1 <= {expr}_2\\
{const}_1 <= {expr}_2 <= {const}_2
\end{array}
\]
where ${const}_i$ are constant arithmetic expressions that may only
contain variables that are fixed, and ${expr}_i$ are arithmetic expressions
that contain unfixed variables.

A complementarity condition is defined with a pair of constraint expressions 
\[
{l}_1 <= {expr}_1 <= {u}_1 \;\;\bot\;\; {l}_2 <= {expr}_2 <= {u}_2,
\]
where exactly two of the constant bounds $l_1$, $u_1$, $l_2$ and
$u_2$ are finite.  The non-finite bounds values are omitted in
practice, so this condition directly describes a classical or mixed
complementarity condition.
Additionally, a complementarity condition can be expressed with a simple inequality:
\[
{l}_1 <= {expr}_1 <= {u}_1 \;\;\bot\;\; {expr}_2 <= {expr}_3.
\]
This complementarity condition is implicitly transformed to a form with constant bounds:
\[
{l}_1 <= {expr}_1 <= {u}_1 \;\;\bot\;\; {expr}_2 - {expr}_3 <= 0.
\]

\subsection{Modeling Mixed-Complementarity Conditions}

Pyomo employs an object-oriented strategy for representing models.
A Pyomo model object contains modeling components that define
standard elements of algebraic models (e.g. parameters, sets,
variables, constraints, and objectives).  This allows Pyomo to
automatically manage the naming of AML components, and multiple
Pyomo models can be simultaneously defined.

Additionally, Pyomo's modeling capabilities can be extended by
simply defining new modeling components.  Pyomo's \texttt{pyomo.mpec}
package defines the \texttt{Complementarity} component that is used
to declare complementarity conditions.

For example, consider the \texttt{ralph1} problem in MacMPEC~\cite{MacMPEC}:
\[
\begin{array}{ll}
\min & 2 x - y\\
     & 0 \leq y \;\bot\; y \geq x\\
     & x,y \geq 0
\end{array}
\]

\if 0
The following script defines a Pyomo model for \texttt{ralph1}:
\listing{examples/ralph1_strip.py}{}{1}{14}
The first lines in this script import Pyomo packages:
\listing{examples/ralph1.py}{imports}{4}{5}
The first line imports \code{pyomo.environ} to initialize Pyomo's environment, and it
imports the core modeling components from Pyomo.  The 
second line imports modeling components for complementarity
conditions.
The subsequent lines in this script create a model, declare variables
\code{x} and \code{y}, declare an objective \code{f1}, and
declare a complementarity condition \code{compl}.  
\fi

The complementarity condition is declared with the \code{Complementarity}
component.  In the simplest case, this Python class takes a keyword
argument \code{expr} that contains the value of the \code{complements}
function.  This function accepts two Pyomo constraint expressions
that are used to declare a complementarity condition.

\if 0
Pyomo also supports indexed components, where a set of components
are initialized over an index set using a construction rule.
Thus, the \texttt{Complementarity} component can be declared
with an index set.  For example, consider the following model, \code{indexed}:
\[
\begin{array}{lll}
\min & \sum_{i=1}^n i (x_i - 1)^2 &\\
     & 0 \leq x_i \;\bot\; 0 \leq x_{i+1} & \;\; i=1,\ldots,n-1
\end{array}
\]
The following script defines a Pyomo model for \code{indexed} with $n=5$:
\listing{examples/ex1a.py}{}{1}{16}
The complementarity conditions are defined with a single
\code{Complementarity} component that is indexed over the set $1,
\ldots, n-1$ and initialized with a construction rule \code{compl\_}.
This rule is a function that accepts a model instance and an index,
and returns the $i$-th complementarity
condition.

The declared set of indexes may be a superset of the indices that define
complementarity conditions.
For example, if the construction rule returns \code{Complementarity.Skip}, then the corresponding
index is skipped.  For example:
\listing{examples/ex1d.py}{}{1}{18}

This example can also be expressed with the 
\code{ComplementarityList} component:
\listing{examples/ex1b.py}{}{1}{18}
This component defines a list of complementarity conditions.  The list index can be 
used in Pyomo, but this component simplifies the declaration of models for which 
the index values are not important.  The \code{ComplementarityList} 
component can also be defined with a rule that iteratively returns complementarity conditions:
\listing{examples/ex1c.py}{}{1}{19}
%The value \code{ComplementarityList.End} is used to denote the end
%of the list.

Similarly, the construction rule may be a list expression that
generates a sequence of complementarity conditions:
\listing{examples/ex1e.py}{}{1}{15}
\fi

\SANDsection{MPEC Transformations}
\label{sec:transformations}

Pyomo's object-oriented design supports the structured transformation
of models.  Pyomo can iterate through model components as well as
nested model blocks.  Thus, model components can be easily transformed
locally, and global data can be collected to support global
transformations.  Further, Pyomo components and blocks can be
activated and deactivated, which facilitates \textit{in place}
transformations that do not require the creation of a separate
copy of the original model.

\if 0
Pyomo's \texttt{pyomo.mpec} package defines several model transformations
that can be easily applied.  For example, if \code{model} defines
an MPEC model (as in our previous examples), then the following
example illustrates how to apply a model transformation:
\listing{examples/ex2.py}{transform}{19}{19}
In this case, the \code{mpec.simple\_nonlinear} transformation is applied.
The following sections describe the transformations currently
supported in \code{pyomo.mpec}.
\fi

\subsection{Standard Form}

In Pyomo, a complementarity condition is expressed as a pair of
constraint expressions
\[
{l}_1 <= {expr}_1 <= {u}_1 \;\;\bot\;\; {l}_2 <= {expr}_2 <= {u}_2,
\]
where exactly two of the constant bounds $l_1$, $u_1$, $l_2$ and
$u_2$ are finite.  The non-finite bounds are typically omitted, but
the value \code{None} can be used to express infinite bounds.
Additionally, each constraint expression can be expressed with a
simple inequality of the form
\[
{expr}_1 <= {expr}_2.
\]

The \code{mpec.simple\_nonlinear} transformation reformulates each
complementarity condition in a model into a standard form:
\[
{l}_1 <= expr <= {u}_1 \;\;\bot\;\; {l}_2 <= var <= {u}_2,
\]
where exactly two of the constant bounds $l_1$, $u_1$, $l_2$ and
$u_2$ are finite, and either $l_2$ is zero or both $l_2$ or $u_2$ are finite.

Note that this transformation creates new variables and constraints
as part of this transformation.  For example, the complementarity
condition
\[
1 \leq x + y \;\;\bot\;\; 1 \leq 2x - y,
\]
get re-expressed as the following:
\[
\begin{array}{l}
1 \leq x + y\\
v = 2x - y - 1\\
v \in \mathbb{R}, v \ge 0
\end{array}
\]

For each complementary condition object, the new variable and
constraints are added as additional components within the
complementarity object.  Thus, the overall structure of the MPEC model
is not changed by this transformation.

\subsection{Simple Nonlinear}

The \code{mpec.simple\_nonlinear} transformation begins by applying
the \code{mpec.standard\_form} transformation.  Subsequently, a
nonlinear constraint is created that defines the complementarity
condition.  This is a simple nonlinear transformation adapted from
\citet{FerDirMee05}, which can be described by three different
cases:
\begin{itemize}

\item If $l_1$ is finite, then the following constraint is defined:
\[
    (expr - l_1) * v \leq \epsilon
\]

\item If $u_1$ is finite, then the following constraint is defined:
\[
    (u_1 - expr) * v \leq \epsilon
\]

\item If $l_2$ and $u_2$ are both finite, then the following constraints are defined:
\[
\begin{array}{l}
(var - l_2) * expr \leq \epsilon\\
(var - u_2) * expr \leq \epsilon\\
\end{array}
\]

\end{itemize}
Each of these cases ensure that the complementarity condition is
met when $\epsilon$ is zero.  For example, in the first case, we
know that $0 \leq v$ and $0 \leq expr - l_1$.  When $\epsilon$ is
zero, this constraint ensures that either $v$ is zero or $expr -
l_1$ is zero.

This transformation uses the parameter \code{mpec\_bound},
which defines the value for $\epsilon$ for every complementarity
condition.  This allows for the specification of a relaxed nonlinear
problem, which may be easier to optimize with some nonlinear
programming solvers.
The default value of \code{mpec\_bound} is zero.

\subsection{Simple Disjunction}

The \code{mpec.simple\_disjunction}
transformation expresses a complementarity condition as a disjunctive
program.  
We are given a complementarity condition defined with a pair of constraint expressions 
\[
{l}_1 <= {expr}_1 <= {u}_1 \;\;\bot\;\; {l}_2 <= {expr}_2 <= {u}_2,
\]
where exactly two of the constant bounds $l_1$, $u_1$, $l_2$ and
$u_2$ are finite.  Without loss of generality, we assume that either $l_1$ or $u_1$ is finite.

This transformation can be described by
three different cases:
\begin{itemize}

\item If the first constraint is an equality, then the complementarity condition is trivially replaced by that equality constraint.

\item If both bounds on the first constraint are finite but different, then the disjunction has the form:
\[
\left[\begin{array}{c}
Y_1\\
l_1 = {expr}_1\\
{expr}_2 \geq 0
\end{array}\right]
\bigvee
\left[\begin{array}{c}
Y_2\\
{expr}_1 = u_1\\
{expr}_2 \leq 0
\end{array}\right]
\bigvee
\left[\begin{array}{c}
Y_3\\
l_1 \leq {expr}_1 \leq u_1\\
{expr}_2 = 0
\end{array}\right]
\]\[
% You need a logic constraint so that only one of the disjuncts is
% selected.  I don't like this notation (and would rather express the
% XOR between the disjuncts, but this is the notation that Ignacio has
% consistently used.
Y_1 \veebar Y_2 \veebar Y_3 = True
\]
% WEH: This seems redundant
%\[
%Y_1 \land Y_2 \land Y_3 = False
%\]\[
% I am being pedantic here: you need to set the domain of the logical
% indicator variables.
\[
Y_1, Y_2, Y_3 \in \{True, False\}
\]

\item Otherwise, each constraint is a simple inequality.  The complementarity condition is reformulated as
\[
0 <= \overline{expr}_1  \;\;\bot\;\; 0 <= \overline{expr}_2,
\]
and the disjunction has the form:
\[
\left[\begin{array}{c}
Y\\
0 = \overline{expr}_1\\
0 \leq \overline{expr}_2
\end{array}\right]
\bigvee
\left[\begin{array}{c}
\lnot Y\\
0 \leq \overline{expr}_1\\
0 = \overline{expr}_2
\end{array}\right]
\]\[
Y \in \{ True, False \}
\]

\end{itemize}

This transformation makes use of modeling components and transformations
from Pyomo's \code{pyomo.gdp} package~\cite{PyomoGDP}.  The
transformation expresses each of the disjunctive terms explicitly using
\code{Disjunct} components and the \emph{select exactly one} logical
condition using the \code{Disjunction} component.  The transformation
adds the \code{Disjunct} and \code{Disjunction} components within the
objects that represent the complementarity conditions.  It then recasts
the modified complementarity components into simple \code{Block}
components.  This localizes all changes to the model to the individual
complementarity components.  Subsequent transformation of the
disjunctive expressions to algebraic constraints can be effected through
either Big-M (\code{gdp.bigm}) or Convex Hull (\code{gdp.chull})
transformations.

\subsection{AMPL Solver Interface}

Solvers like PATH~\cite{FerMun98} have been tailored to work with
the AMPL Solver Library (ASL).  AMPL uses \code{nl} files to
communicate with solvers, which read \code{nl} files with the ASL.
Pyomo can also create \code{nl} files, and the \code{mpec.nl}
transformation processes \code{Complementarity} components into a
canonical form that is suitable for this format~\cite{FerFouGay99}.

\SANDsection{Solver Interfaces and Meta-Solvers}
\label{sec:solvers}

Pyomo supports interfaces to third-party solvers as well as meta-solvers that 
apply transformations and third-party solvers, perhaps in an iterative manner.
The \code{pyomo.mpec} package includes an interface to the \code{PATH} solver, as well as several 
meta-solvers.  These are described in this section, and examples are provided that 
employ the \code{pyomo} command-line interface.

\subsection{Nonlinear Reformulations}

The \code{mpec.simple\_nonlinear} transformation provides a generic
way for transforming an MPEC into a nonlinear program.  When the
MPEC only has continuous decision variables, the resulting model
can be optimized by a wide range of solvers.

\if 0
For example, the \code{pyomo} command-line interface allows the
user to specify a nonlinear solver and a model transformation that
is applied to a model:
\listing{examples/nlp1.sh}{pyomo}{4}{4}
This example illustrates the use of the \code{ipopt} interior-point
solver with the\linebreak \code{mpec.simple\_nonlinear} transformation.  
When a transformation is used directly like this, the results
that are returned to the user include decision variables for the transformed
model.  Pyomo does not have general capabilites for mapping a
solution back into the space from the original model.  In this
example, the results object includes values for the $x$ variables
as well as the variables $v$ introduced when applying the transformation
to the standard form (see above).

Pyomo includes a meta-solver, \code{mpec\_nlp} that applies the
nonlinear transformation, performs optimization, and then returns
results for the original decision variables.  For example, \code{mpec\_nlp} 
executes the same logic as the previous \code{pyomo} example:
\listing{examples/nlp2.sh}{pyomo}{4}{4}
Additionally, this meta-solver can also manipulate the $\epsilon$
values in the model, starting with larger values and iteratively
tightening them to generate a more accurate model.
\listing{examples/nlp3.sh}{pyomo}{4}{5}
This approach may be useful when using a nonlinear solver that has difficulty
optimizing with equality constraints.
\fi

\subsection{Disjunctive Reformulations}

The \code{mpec.simple\_disjunction} transformation provides a generic
way for transforming an MPEC into a disjunctive program.  The
\code{mpec\_minlp} solver applies this transformation to create a
nonlinear disjunctive program, and then further reformulates the
disjunctive model using a ``Big-M'' transformation that is provided
by the \code{pyomo.gdp} package.  The resulting transformation is
similar the reformulation of bilevel models described by~\citet{ForMcC81}.
If the original model was nonlinear, then the resulting model is a
mixed-integer nonlinear program (MINLP).  Pyomo includes interfaces
to solvers that use the AMPL Solver Library (ASL), so \code{mpec\_minlp} can 
optimize nonlinear MPECs with a solver like Couenne~\cite{COUENNE}.

\if 0
If the original model was a linear MPEC, then the resulting model
is a mixed-integer linear program that can be globally optimized
(e.g. see~\citet{HuMitPanBenKun08,Jud11}).  For example, the
\code{pyomo} command can be used to execute the \code{mpec\_minlp}
solver using a specified MIP solver:
\listing{examples/mip1.sh}{pyomo}{4}{4}
Note that Pyomo includes interfaces to a variety of commonly used
MIP solvers, including CPLEX, Gurobi, CBC, and GLPK.
\fi

\subsection{PATH and the ASL Solver Interface}

Pyomo's solver interface for the AMPL Solver Library (ASL) applies
the \code{mpec.nl} transformation, writes an AMPL \code{.nl} file, executes
an ASL solver, and then loads the solution into the original model.
Pyomo provides a custom interface to the PATH solver~\cite{FerMun98},
which simply allows the solver to be specified as \code{path} while
the solver executable is named \code{pathamp}.

\if 0
The \code{pyomo} command can execute the PATH solver by simply specifying the 
\code{path} solver name.  For example, consider the \code{munson1} problem 
from MCPLIB:
\listing{examples/munson1.py}{}{1}{23}
%\pagebreak
This problem can be solved with the following command:
\listing{examples/path1.sh}{pyomo}{4}{4}
\fi

\SANDsection{Discussion}
\label{sec:discussion}

Pyomo supports the ability to model complementarity conditions in a manner that is
similar to other AMLs.  For example, Pyomo's \code{pyomo.data}
package~\cite{pyomo.data} includes Pyomo formulations for many of
the \mbox{MacMPEC}~\cite{MacMPEC} and MCPLIB~\cite{MCPLIB} models, which were originally formulated in GAMS and AMPL.  However,
Pyomo does not currently support related modeling capabilities for
equilibrium models, variational inequalities and embedded models,
which are supported by the GAMS extended mathematical programming
framework~\cite{FerDirJagMee09}.

The tranformations and meta-solvers currently included in Pyomo
illustrate how Pyomo's MPEC modeling capability can be leveraged.
We expect these capabilities to mature and expand in response to
application needs.  For example, the \code{mpec.simple\_nonlinear}
transformation could be expanded to support reformulations that
are well-suited for sequential quadratic programming solvers~\cite{Ley06}.
Similarly, current meta-solvers could be extended to directly
support the communication of suffix information from the solver
back to the original model.

### 2019-04-30 doc/bilevel/paper.tex
Adding draft documentation for pao.bilevel
I haven't yet confirmed that this compiles.  I'm not expecting we'll
use all of this text here.

\lstset{language=Python}
\definecolor{light-gray}{gray}{0.9}
\lstset{backgroundcolor=\color{light-gray}}
\lstset{aboveskip=1em,belowskip=1em,showspaces=false,showstringspaces=false}

\SANDsection{Introduction}

% TODO: More references here

Many planning situtations involve the analysis of several objectives
that reflect a hierarchy of decision-makers.  For example, policy
decisions are made at different levels of a government, each of
which has a different objective and decision space.  Similarly, robust planning
against adversaries is often modeled with a 2-level hierarchy, where
the defensive planner makes decisions that account for adversarial
response.

\textit{Multilevel optimization} techniques partition control over
decision variables amongst the levels.  Decisions at each level of
the hierarchy may be constrained by decisions at other levels, and
the objectives for each level may account for decisions made at
other levels.  In practice, multilevel problems have proven difficult
to solve, and most of the literature has focused on \textit{bilevel
programs}, which model a 2-level hierarchy~\citep{ColMarSav07}.

Although multilevel problems arise in many applications, few algebraic
modeling languages (AML) have integrated capabilities for expressing
these problems.  AMLs are high-level programming languages for
describing and solving mathematical problems, particularly
optimization-related problems~\citep{Kal04}.  AMLs provide a mechanism
for defining variables and generating constraints with a concise
mathematical representation, which is essential for large-scale,
real-world problems that involve thousands or millions of constraints
and variables.  GAMS~\citep{GAMS}, YALMIP~\cite{Lof04} and Pyomo
provide explicit support for modeling bilevel programs.  A variety
of other AMLs support the solution of bilevel programs through the
expression of Karush-Kuhn-Tucker conditions and associated
reformulations using mixed-complementarity conditions, but these
reformulations must be expressed by the user in these AMLs.

In this paper, we describe new functionality in Pyomo 4.3 for
expressing and optimizing multilevel models in the Pyomo modeling
environment.  Pyomo is an open-source software package that supports
the definition and solution of optimization applications using the
Python language~\cite{pyomotrac,pyomoweb,HarWatWoo11,HarLaiWatWoo12}.
Python is a powerful programming language that has a clear, readable
syntax and intuitive object orientation.  Pyomo uses an object-oriented
approach for defining models that contain decision variables,
objectives, and constraints.

The main point of this paper is to demonstrate that Pyomo provides
an intuitive syntax for expressing multilevel optimization problems.
Multilevel models can be easily expressed with Pyomo modeling
components for submodels, which can be nested in a general manner.
Further, Pyomo's object-oriented design naturally supports the
ability to automate the reformulation of multilevel models into
other forms.  In particular, we describe Pyomo's capabilities for
transforming bilevel models for several broad classes of problems.
We describe Pyomo meta-solvers that transform bilevel programs into
mixed integer programs (MIP) or nonlinear programs (NLP), which are
then optimized with standard solvers.

The remainder of this paper is organized as follows.
Section~\ref{sec:problems} describes motivating classes of bilevel
programming problems.  Section~\ref{sec:design} describes how Pyomo
supports modeling of submodels.  Sections~\ref{sec:BLP} and~\ref{sec:QMM}
describe transformations that automate the reformulation of bilevel
programs and meta-solvers in Pyomo that leverage these transformations
to support global and local optimization of bilevel programs.
Section~\ref{sec:discussion} discusses Pyomo's ability to model multilevel problems, and possible extensions for non-cooperative bilevel programs.

\SANDsection{Motivating Problems}
\label{sec:problems}

In multilevel optimization problems, a subset of decision variables
at each level is constrained to take values associated with an
optimal solution of a distinct, lower level optimization problem.
For example, a general formulation for bilevel programs is
\begin{equation}
\label{eqn:bilevel}
\begin{array}{ll}
\min_{x \in X, y}   & F(x,y)\\
\st                 & G(x,y) \leq 0\\
                    & y \in P(x)
\end{array}
\end{equation}
where
\[
\begin{array}{llll}
P(x) & = & \argmin_{y \in Y}    & f(x,y)\\
 & &                            & g(x,y) \leq 0
\end{array}
\]
$P(x)$ defines a \textit{lower-level problem}, which may have multiple solutions.
Here $x$ is the primary upper-level decision, and $y$ is the anticipated lower-level 
decision.

When $P(x)$ contains multiple solutions, this formulation ensures
that the selected value for $y$ will minimize $F(x,y)$.  Consequently,
this formulation has been called \textit{optimistic} or
\textit{cooperative}, since the selection of the lower-level decision
variables minimized the upper-level objective.  Most research on
bilevel programming has considered this problem, since this formulation
has an optimal solution under reasonable assumptions.

The following subsections describe specializations of
Equation~(\ref{eqn:bilevel}) that will be explored in greater detail
throughout this paper.  These are well-studied classes of bilevel
programs that that can be reformulated into other canonical
optimization forms.

\subsection{Linear Bilevel Programs with Continuous Variables}

Multilevel linear programming considers the case where decision
variables are continuous, and both objectives and contraints are
linear.  In the 2-level case, we have the following linear bilevel program:
\begin{equation}
\label{eqn:bilevel_A}
\begin{array}{lll}
\min_{x, y}         & c^T_1 x + d^T_1 y & \\
\st                 & A_1 x + B_1 y \leq b_1 & \\
                    & x \geq 0 & \\
                    & \min_{y}  & c^T_2 x + d^T_2 y\\
                    &           & A_2 x + B_2 y \leq b_2\\
                    &           & y \geq 0
\end{array}
\end{equation}

\subsection{Quadratic Min/Max}

Consider the case where the lower-level decisions $y$ do not constrain the upper-level decisions.  Let
\[
X = \{ x \mid A_1 x \leq b, x \geq 0\}
\]
Then we have
\begin{equation}
\label{eqn:bilevel_B}
\begin{array}{lll}
\min_{x \in X} & \max_{y \geq 0} & c^T_1 x + d^T_1 y + x^T Q y\\
& & A_2 x + B_2 y \leq b_2
\end{array}
\end{equation}
In our discussion below, we allow the $x_i$ to be binary without loss of generality.

\SANDsection{Modeling Bilevel Programs}
\label{sec:design}

\subsection{Pyomo Models}

Pyomo supports an object-oriented design for the definition of
optimization models.  The basic steps of a simple modeling process are as follows:
\begin{quote}
\begin{enumerate}
\item Create model object and declare components
\item Instantiate the model
\item Apply solver
\item Interrogate solver results
\end{enumerate}
\end{quote}
In practice, these steps may be applied repeatedly with different
data or with different constraints applied to the model.  However,
we focus on this simple modeling process throughout this paper.

A Pyomo {\em model} consists of a collection of modeling {\em
components} that define different aspects of the model.  Pyomo
includes the modeling components that are commonly supported by
modern AMLs:  components for decision variables, objectives and
constraints are used to express the model, and components for index
sets and symbolic parameters are used to manage data used in model
expressions.  These modeling components are defined in Pyomo through
the following Python classes:
\begin{center}
\begin{tabular}{ll}
\code{Var} & decision variables in a model\\
\code{Objective} & expressions that are minimized or maximized in a model\\
\code{Constraint} \hspace{0.2in} & constraint expressions in a model\\
\code{Set} & set data that is used to define a model instance\\
\code{Param} & parameter data that is used to define a model instance
\end{tabular}
\end{center}\mbox{}

Modeling components are directly added to a Pyomo model object as an attribute of the object.
For example:
\begin{qlisting}
model = ConcreteModel()
model.x = Var()
\end{qlisting}
Pyomo defines two model classes: \code{ConcreteModel} and \code{AbstractModel}.  
In concrete models, components are immediately initialized when
they are added to the model.  In abstract models, components are
explicitly initialized later, which allows the creation of multiple
model instances from a single abstract model.

Another Pyomo component is the \code{Block} class, which defines a
named collection of model components.  Blocks provide a convenient
mechanism for segregating different elements of a model within
separate namespaces, and they can be used to define repetitive model
structure in a concise manner.  For example, the following model
defines two variables \code{x}:
\begin{qlisting}
model = ConcreteModel()
model.x = Var()
model.b = Block()
model.b.x = Var()
\end{qlisting}
Block \code{b} defines a namespace that allows these two variables to be
uniquely referenced.

\subsection{Modeling Bilevel Programs}

The \code{pyomo.bilevel} package extends Pyomo by defining a new
modeling component: \code{SubModel}.  The \code{SubModel} component
defines a subproblem that represents the lower level decisions in
a bilevel program.  This component is like a \code{Block} component;
any components can be added to a \code{SubModel} object.  In general,
a submodel is expected to have an objective, one or more variables
and it may define constraints.

The \code{SubModel} class generalizes the \code{Block} component
by including constructor arguments that denote which variables in
the submodel should be considered fixed or variable.  When expressions
in a submodel refer to variables defined outside of the submodel,
the user needs to indicate whether these are fixed values defined
by an upper-level problem.  Fixed variables are treated as constants
within the submodel, but non-fixed variables are defined by the
current submodel or by a lower-level problem.

Consider the following example:
\begin{equation}
\begin{array}{lll}
\min_{x, y, v}      & x + y + v & \\
\st                 & x + v \geq 1.5 \\
                    & 1 \leq x \leq 2 &\\
                    & 1 \leq v \leq 2 & \\
                    & \max_{y,w}  & x + w\\
                    &           & y + w \leq 2.5 \\
                    &           & 1 \leq y \leq 2\\
                    &           & 1 \leq w \leq 2
\end{array}
\end{equation}

\newpage
The following Pyomo model defines four variables, \code{x}, \code{v}, \code{sub.y} and \code{sub.w}:
\listing{examples/simple.py}{pyomo}{2}{15}
Variables \code{x} and \code{v} are declared in the upper-level
problem, and \code{v} only appears in the upper-level problem.
Variables \code{sub.y} and \code{sub.w} are declared in the submodel.
However, note that the \code{sub.y} variable appears in the upper-level
problem, while the \code{sub.w} variable only appears in the
lower-level problem.

\if 0
 {examples/simple_implicit.py}{pyomo}{2}{16}
The submodel object declares that the \code{x} variable is fixed,
so it is treated as an upper-level variable.  
\fi

These modeling capabilities are quite general.  Expressions in
submodels can be linear or nonlinear, convex or nonconvex, continuous
or discontinuous, and more.  Additionally, submodels can be nested
to an arbitrary degree.  Thus, the range of bilevel programs that
can be expressed with Pyomo is quite broad.  However, the real
challenge is solving these models.  The following sections describe
two general strategies for solving the two classes of bilevel
programs described in Section~\ref{sec:problems}.

In each section we describe model transformations and their use in
meta-solvers.  Pyomo's object-oriented design supports the structured
transformation of models.  Pyomo can iterate through model components
as well as nested model blocks.  Thus, model components can be
easily transformed locally, and global data can be collected to
support global transformations.  Further, Pyomo components and
blocks can be activated and deactivated, which facilitates \textit{in
place} transformations that do not require the creation of a separate
copy of the original model.

\if 0
Similarly, the submodel could have been declared as:
\begin{qlisting}
model.sub = SubModel(var=model.y)
\end{qlisting}
The \code{fixed} and \code{var} arguments complement each other,
and they will likely be used in different contexts.
\fi

\if 0
\subsection{Comparison with GAMS}
\label{gams}

To our knowledge, GAMS is the only modern algebraic modeling language
(AML) that supports the definition and automated analysis of bilevel
programs.  Thus, it is worth comparing Pyomo's bilevel modeling
capability with GAMS; we later contrast the solution strategies
supported by GAMS and Pyomo.

Consider the following bilevel program described by Bard~\cite{Bard98} (example 5.1.1):
\begin{equation*}
\begin{array}{lll}
\min_{x \geq 0} & x - 4 y & \\
\st & \min_{y \geq 0} & y\\
 & \st & -x -y \leq -3\\
& & -2 x + y \leq 0\\
& & 2 x + y \leq 12\\
& & 3 x - 2 y \leq 4
\end{array}
\end{equation*}

\if 0
The following GAMS model defines a bilevel program for this
model:\footnote{ This is the \code{bard511.gms} model from the
online GAMS EMPLIB.}
\listing{bard511.gms}
GAMS allows modelers to declare equations and inequalities, which
can subsequently be composed into a model.  The \code{emp} solver
exploits this flexibility through its use of a configuration file,
\code{emp.info}, that defines the structure of the bilevel program.
In this example, the \code{emp.info} file is generated automatically, and the equations
\code{defin}, \code{e1}, \code{e2}, \code{e3} and \code{e4} are included in the
submodel.
\fi

The following Pyomo model defines a bilevel program for this model:
\listing{examples/bard511.py}{}{1}{19}
This model defines the bilevel structure in an object-oriented
manner.  Consequently, the structure is an inherent part of the
model, thereby avoiding the use of auxilliary files in the specification of
the model.

Note that the Pyomo example does not describe how the model is
optimized.  The following section describes a specific model
reformulations strategy for a broad class of bilevel programs.
Further, this section describes a generic solver strategy that leverages
this reformulation to provide an automated solution process.
\fi

\SANDsection{Solving Linear Bilevel Programs}
\label{sec:BLP}

We consider the formulation for linear bilevel programs described
in Equation~(\ref{eqn:bilevel_A}):
\begin{equation}
\begin{array}{lll}
\min_{x, y}         & c^T_1 x + d^T_1 y & \\
\st                 & A_1 x + B_1 y \leq b_1 & \\
                    & x \geq 0 & \\
                    & \min_{y}  & c^T_2 x + d^T_2 y\\
                    &           & A_2 x + B_2 y \leq b_2\\
                    &           & y \geq 0
\end{array}
\tag{\ref{eqn:bilevel_A}}
\end{equation}
Following \citet{Bard98}, we can
replace the lower-level problem with corresponding optimality
conditions.  This transformation gives the following model:
\begin{equation}
\label{eqn:bilevel_A_xfrm}
\begin{array}{ll}
\min & c^T_1 x + d^T_1 y\\
\st  & A_1 x + B_1 y \leq b_1\\
     & d_2 + B^T_2 u - v = 0\\
     & b_2 - A_2 x - B_2 y \geq 0 \perp u \geq 0\\
     & y \geq 0 \perp v \geq 0\\
     & x \geq 0, y \geq 0
\end{array}
\end{equation}
This transformation results in a mathematical program with equillibrium constraints (MPEC).

Pyomo's \texttt{pyomo.bilevel} package automates the application
of this model transformation.  For example, if \code{model} defines
an linear bilevel program,  then the code applies this transformation
to change the model in place:
\listing{examples/ex2.py}{transform}{6}{7}
The \code{bilevel.linear\_mpec} transformation modifies the model
by creating a new block of variables and constraints for the
optimality conditions in the lower-level problem.

A variety of solution strategies can leverage this transformation
to support optimization.  Pyomo defines two meta-solvers, which
apply the \code{bilevel.linear\_mpec} transformation and then perform
optimization with a third-party solver.

Consider the following problem, which is Example 5.1.1. in Bard~\citep{Bard98}:
\begin{equation}
\begin{array}{llrrrrrr}
\min_{x, y}     & x - 4 y & &&&&\\
                & x \geq 0 & &&&&\\
\st             & \min_{y} y & &&&&\\
                & \st   & - x &- &y &\leq &-3\\
                &       & -2 x &+ &y &\leq &0\\
                &       & 2 x &+ &y &\leq &12\\
                &       & 3 x &- &2 y &\leq &4
\end{array}
\end{equation}
Note that the last constraint in this example is negated from the text in \citet{Bard98};  this corrects an error in the example, which is reflected in Bard's discussion of the solution to this example.  The Pyomo model for this problem is:
\listing{examples/bard511.py}{}{1}{19}

\subsection{Global Optimization}

Following Fortuny-amat and McCarl~\cite{ForMcC81}, Pyomo's
\code{bilevel\_blp\_global} meta-solver chains together reformulations
to generate the following sequence of models:
\[
\textrm{BLP} \Rightarrow \textrm{MPEC} \Rightarrow \textrm{GDP} \Rightarrow \textrm{MIP},
\]
where GDP refers to generalized disjunctive programs.  Note that
this leverages advanced modeling capabilities in Pyomo for
MPEC~\cite{PyomoMPEC} and GDP~\cite{PyomoGDP}.

Pyomo provides general support for solving MIPs with commercial and
open-source solvers.  The \code{bilevel\_blp\_global} meta-solver
applies these transformations, solves the resulting MIP, and
translates the MIP solution back into the original linear bilevel
program.  The result is a globally optimal solution.  For example,
the \code{pyomo} command can be used to execute the
\code{mpec\_blp\_global} solver using a specified MIP solver:
\datalisting{examples/mip1a.sh}{pyomo}{4}{5}
Note that a ``Big-M'' transformation is used to convert the GDP
model into a MIP.  The default M value is very large, which may
make it difficult to solve the resulting MIP problem.  Hence, this solver includes
a \code{bigM} option that can be used to specify a problem-specific value:
\datalisting{examples/mip2a.sh}{pyomo}{4}{5}

\subsection{Local Optimization}

Pyomo's \code{bilevel\_blp\_local} meta-solver chains together reformulations to generate the
following sequence of models:
\[
\textrm{BLP} \Rightarrow \textrm{MPEC} \Rightarrow \textrm{NLP}.
\]
This leverages model transformations in \code{pyomo.gdp} to transform an MPEC into an NLP through a simple nonlinear transformation adapted from \citet{FerDirMee05}.  For example, the complementarity condition
\[
y \geq 0 \perp w \geq 0
\]
is transformed to the constraints
\[
\begin{array}{l}
y \geq 0\\
w \geq 0\\
y w \leq \epsilon .
\end{array}
\]

Pyomo provides general support for solving NLPs with commercial and
open-source solvers.  The \code{bilevel\_blp\_local} meta-solver
applies these transformations, solves the resulting NLP, and
translates the solution back into the original linear bilevel
program.  In general, the result is a locally optimal solution.
For example, the \code{pyomo} command can be used to execute the
\code{mpec\_blp\_local} solver using a specified NLP solver:
\datalisting{examples/nlp1a.sh}{pyomo}{4}{5}
Note that the tolerance value $\epsilon$ is initially set to a small
value, which some solvers may have difficulty with.  This value can
be explicitly set with the \code{mpec\_bound} option:
\datalisting{examples/nlp2a.sh}{pyomo}{4}{5}

\SANDsection{Solving Quadratic Min-Max Bilevel Programs}
\label{sec:QMM}

We consider the formulation for quadratic min-max bilevel programs
described in Equation~(\ref{eqn:bilevel_B}):
\begin{equation}
\begin{array}{lll}
\min_{x \in X} & \max_{y \geq 0} & c^T_1 x + d^T_1 y + x^T Q y\\
& & A_2 x + B_2 y \leq b_2
\end{array}
\tag{\ref{eqn:bilevel_B}}
\end{equation}
where
\[
X = \{ x \mid A_1 x \leq b, x \geq 0\}.
\]
The lower-level problem
is linear, so we can replace the lower-level problem with corresponding
optimality conditions.  Since, the objectives are opposite and the
upper-level constraints do not constrain the lower level decisions,
we get a single minimization problem.  This transformation gives
the following model:
\begin{equation}
\label{eqn:bilevel_B_xfrm}
\begin{array}{ll}
\min & c^T_1 x + (b_2 - A_2 x)^T \nu\\
\st  & B^T_2 \nu \geq d_1 + Q^T x\\
     & A_1 x \leq b_1\\
     & x \geq 0, \nu \geq 0
\end{array}
\end{equation}
In Pyomo, this is implemented as the \code{bilevel.linear\_dual} transformation.

If $A_2 \equiv 0$, then the lower-level problem does
not constrain the upper-level decisions.  This is a simple case, where the transformation
generates a linear program if the upper-level decision variables
$x$ are continuous and a MIP if some or all of the $x$ are binary.

More generally, suppose that the upper-level decision variables $x$ are
binary and $A_2 \not \equiv 0$.  We can linearize the quadratic terms
in the objective using \code{gdp.bilinear} transformation, which
creates the following disjunctive representation:
\begin{equation}
\begin{array}{ll}
\min & c^T_1 x + b^T_2 \nu - 1^T z\\
\st  & B^T_2 \nu \geq d_1 + Q^T x\\
     & A_1 x \leq b_1\\
     & \left(\begin{array}{l}
       x_i = 0\\
       z_i = 0
       \end{array}\right) \wedge \left(\begin{array}{l}
       x_i = 1\\
       z_i = A^T_2(*,i) \nu
       \end{array}\right)\\
     & x_i \geq \{0,1\},  \nu \geq 0
\end{array}
\end{equation}
Subsequently, this GDP can be transformed into a MIP using a Big-M transformation.

Consider the following network interdiction problem, for which an
attacker eliminates links in a network to minimize the maximum flow
through the network from a fixed source $s$ to a fixed destination
$t$.  Let $N$ be the nodes in the network through which flow occurs.
Let $y_{ij}$ be a variable that indicates flow from node $i$
to node $j$, and let $c_{ij}$ be the maximum capacity on that arc.  The attacker selects 
$b$ variables $x_{ij}$; if $x_{ij}$ is one then the arc is removed.  This network interdiction problem can be
written as:
\begin{equation}
\begin{array}{lllll}
\min_{x \in X} & \max_{y} & \eta & & \\
    & & \sum_{i \in N} y_{in} = \sum_{j \in N} y_{nj} & \forall n \in N & \textrm{(flow balance constraint)}\\
    & & \eta \leq \sum_{j \in N} y_{sj} & & \textrm{(node s flow)}\\
    & & \eta \leq \sum_{i \in N} y_{it} & & \textrm{(node t flow)}\\
    & & 0 \leq y_{ij} \leq t_{ij} (1-x_{ij}) & \forall \;\textrm{arcs}\; (i,j) & \textrm{(capacity constraint)}
\end{array}
\end{equation}
where
\[
X = \left\{x \mid x_{ij} \in \{0,1\}, \sum_{i,j} x_{ij} \leq b\right\}.
\]

\newpage
The following Pyomo model describes this bilevel program:
\datalisting{examples/interdiction.py}{}{1}{42}
In this example, the file \code{interdiction\_data.py} defines a
simple network with 6 nodes (including $s$ and $t$), adapted from
an example by Will Traves:
\begin{center}
\includegraphics[scale=0.4]{network.png}
\end{center}

Pyomo's \code{bilevel\_ld} meta-solver applies the
\code{bilevel.linear\_dual} transformation and then applies subsequent
transformations when $A_2 \not \equiv 0$.
\datalisting{examples/ld1a.sh}{pyomo}{4}{5}
Note that this solver includes a \code{bigM} option that can be
used to specify a problem-specific value when a MIP is generated
from a GDP.

\SANDsection{Discussion}
\label{sec:discussion}

Note that Pyomo's ability to model multilevel optimization problem
extends far beyond the bilevel programs that are currently supported.
For example, declarations of \code{SubModel} can be arbitrarily
nested with clear semantics.  For example, consider the following trilevel model~\cite{ZhaLuMonZen10}:
\[
\begin{array}{ll}
\min_{x, y, z}  & x - 4y + 2z\\
\st             & -x -y \leq -3\\
                & -3x +2y -z \geq -10\\
                & \begin{array}{ll}\\
                    \min_{y,z} & x + y - z\\
                    \st & -2x + y -2z \leq -1\\
                        & 2x + y + 4z \leq 14\\
                        & \begin{array}{ll}\\
                            \min_{z} & x - 2y - 2z\\
                            \st & 2x - y - z \leq 2
                          \end{array}
                    \end{array}
\end{array}
\]
The following model illustrates how this trilevel model could be implemented in Pyomo:
\listing{examples/trilevel.py}{pyomo}{2}{21}

Pyomo use of object-oriented model specification makes it fundamentally
different from the specification of bilevel models in GAMS and
YALMIP.  Both GAMS and YALMIP allow users to specify expressions
for variables, objectives and constraints, and then the users
specifies which of these are associated with an upper-level or
lower-level problem.  This design allows users to mix-and-match
different modeling components in a flexible manner.  However, it
is limited to a strictly bilevel form.  By contrast, Pyomo submodels
can be nested in an arbitrary manner.  This includes multilevel
models, as was just illustrated.  But it also allows for the
specification of a tree of nested submodels.  For example, Pyomo
supports the specification of independent submodels at the same
level, which can be used to model a single agent cooperating with
decisions for two independent agents that make subsequent decisions..

Earlier, we noted that Pyomo supports optimistic or cooperative
bilevel models.  It seems relatively straightforward to extend
Pyomo's current semantics to support pessimistic bilevel models.
For example, a \code{pessimistic} constructor option could be added
to the \code{SubModel} component.  This capability will be added
when transformations and/or solvers are added to Pyomo for these
bilevel models.

### 2019-04-30 doc/bilevel/lstanchors.pl

@@ -1,0 +1,91 @@
+#!/usr/bin/env perl
+
+$re = '^(.*?)(\\listing|\\spacedlisting|\\autobreaklisting|\\scriptsizelisting|\\tinylisting)\{([^\}]*)\}\{([^\}]*)\}(\{[^\}]*\}\{[^\}]*\})?(.*)';
+
+unless ( @ARGV )
+  { die "Usage: $0 target.tex ...\n"; }
+
+foreach $f ( @ARGV )
+  {
+  unless ( -f $f and -r $f and -w $f )
+    {
+    print "File '$f' does not appear to be a valid target.  Skipping.\n";
+    next;
+    }
+  print "Updating $f\n";
+  &update_file($f);
+  }
+
+exit 0;
+
+
+sub update_file($)
+  {
+  my($fname) = @_;
+
+  open FILE, "<$fname" or die "Error opening TeX file $fname\n";
+
+  my(@lines) = ();
+  foreach $line ( <FILE> )
+    {
+    unless ( $line =~ /$re/so )
+      { push @lines, $line; }
+    else
+      {
+      print "Updating reference: $line";
+      my($front) = $1;
+      my($cmd)   = $2;
+      my($file)  = $3;
+      my($anchor)= $4;
+      my($firstlast) = $5;
+      my($end)   = $6;
+
+      my($firstline) = -1;
+      my($lastline) = -1;
+      open REF, "<$file" or die("Error opening referenced file $file ".
+                                "while processing $fname\n");
+      my($lineno) = 0;
+      # scan file for the anchor tags (start = '@anchor:', end = '@:anchor')
+      while ( $_ = <REF> )
+        {
+        ++$lineno;
+        if ( /\@${anchor}:/ )
+          {
+          if ( $firstline != -1 )
+            { die("Error: multiple starting tags found for anchor $anchor ".
+                  " in file $file while processing $fname"); }
+          $firstline = $lineno+1;
+          }
+        if ( /\@:${anchor}$/ )
+          {
+          if ( $lastline != -1 )
+            { die("Error: multiple ending tags found for anchor $anchor ".
+                  " in file $file while processing $fname"); }
+          $lastline = $lineno-1;
+          }
+        }
+      close REF;
+
+      if ( $firstline == -1 )
+        { $firstline = 1; }
+    
+      if ( $lastline == -1 )
+        { $lastline = $lineno-1; }
+      
+      # form the modified line
[…]

### 2019-05-07 doc/bilevel/main.tex
notes in Abstract on what we need to do next

@@
 \paperabstract{
-We describe capabilities for modeling bilevel and multilevel programs
-within the Pyomo modeling software.  These capabilities include
-modeling components that represent subproblems, modeling transformations
-for re-expressing models with bilevel structure in other forms, and
-optimize bilevel programs with meta-solvers that apply transformations
-and then perform optimization on the resulting model.  We illustrate
-the application of Pyomo to model powergrid reliability problems.
+We will use this file to document the following:\\
+1. Bilevel model forms\\
+\indent	a. Canonical Model Form (e.g., generic affine or convex constraints, discrete variables?)\\
+\indent	b. Numerical Example\\
+\indent	c. Current Pyomo Code for (a)\\
+\indent	d. How would this be solved?\\
+2. Uncertainty model forms\\
+
 }

### 2019-05-08 doc/bilevel/CommonAbstract.tex
Fixing Anya's last edits
The main.tex file should not have been used here.  I moved the
edits to the abstract into the CommonAbstract.tex file.

@@
+We will use this file to document the following:\\
+1. Bilevel model forms\\
+\indent a. Canonical Model Form (e.g., generic affine or convex constraints, discrete variables?)\\
+\indent b. Numerical Example\\
+\indent c. Current Pyomo Code for (a)\\
+\indent d. How would this be solved?\\
+2. Uncertainty model forms\\
+
+\if 0
 We describe new capabilities for modeling bilevel programs within
@@
 global optimization of bilevel programs.
+\fi
 

### 2019-05-13 doc/bilevel/paper.tex
Slight reorg.

@@
 \lstset{aboveskip=1em,belowskip=1em,showspaces=false,showstringspaces=false}
+
+\if 0
+
+% WEH - For now, let's omit this introduction since it needs to be 
+%       reworked anyway.
 
@@
 
-\SANDsection{Motivating Problems}
-\label{sec:problems}
+\fi
+
+\SANDsection{Expressing Multilevel Programs}
+\label{sec:modeling}
 
@@
 \begin{array}{ll}
-\min_{x \in X, y}   & F(x,y)\\
+\min_{x \in X, y \in Y}   & F(x,y)\\
 \st                 & G(x,y) \leq 0\\
@@
 has an optimal solution under reasonable assumptions.
+
+\textbf{TODO: Describe pessimistic problems here.}
+
+Many authors use a more compact form to express (\ref{eqn:bilevel}):
+\begin{equation}
+\label{eqn:bilevel-compact}
+\begin{array}{lll}
+\min_{x \in X, y \in Y}   & F(x,y) & \\
+\st                       & G(x,y) \leq 0 & \\
+                          & \min_{y \in Y}    & f(x,y)\\
+                          &                   & g(x,y) \leq 0
+\end{array}
+\end{equation}
+This notation defines an implied constraint by not explicitly referencing the solution to the lower-level problem.
+
+The \code{pyomo.bilevel} package uses the \code{SubModel} component to declare a block that defines a lower-level problem. 
+For example, consider:
+\begin{qlisting}
+model = ConcreteModel()
+model.x = Var(bounds=(1,2))
+model.v = Var(bounds=(1,2))
+model.sub = SubModel()
+model.sub.y = Var(bounds=(1,2))
+model.sub.w = Var(bounds=(-1,1))
+
+model.o = Objective(expr=model.x + model.sub.y + model.v)
+model.c = Constraint(expr=model.x + model.v >= 1.5)
+model.sub.o = Objective(expr=model.x+model.sub.w, sense=maximize)
+model.sub.c = Constraint(expr=model.sub.y + model.sub.w <= 2.5)
+\end{qlisting}
+This Pyomo model defines four variables, \code{x}, \code{v}, \code{sub.y} and \code{sub.w}.
+Variables \code{x} and \code{v} are declared in the upper-level
+problem, and \code{v} only appears in the upper-level problem.
+Variables \code{sub.y} and \code{sub.w} are declared in the submodel.
+However, note that the \code{sub.y} variable appears in the upper-level
+problem, while the \code{sub.w} variable only appears in the
+lower-level problem.
+
+While this syntax mimics the structure of (\ref{eqn:bilevel-compact}), it adds additional abstractions that make it difficult to interpret:
+\begin{itemize}
+
+\item It can be difficult to easily discern which are the upper and lower level variables.
+In general, Pyomo does not place strong restrictions on the location of variable declarations relative to the location where those variables are used in expressions.
+Thus, while a hierarchical block structure suggests nested model representations, this is not strongly enforced when objective and constraint expressions are generated.\\
+\mbox{}\\
+IDEA: Enforce a stronger semantics on PAO block representations for bilevel programs.\\
+IDEA: Use model annotations to denote upper- and lower- level variables.\\
+
+\item This syntax does not indicate how the solution of the lower level relates to constraints in the upper level problem.
+Specifically, there is no explicit indication how solutions to the lower level problem relate to variables in the upper-level problem.
+In simple applications, where there is a direct correspondence of variables (using the same names), this simple syntax may be sufficient.
+But in more complex applications the variables need not have the same naming in the upper- and lower-level problems.
+Thus, PAO needs a mechanism to associate these variables.
+
+\end{itemize}
+
+There are also practical issues that reflect the needs of users with
+respect to the solution of bilevel programs.  Generally, it is assumed
+that the solution of~(\ref{eqn:bilevel}) consists of the optimal upper-
+and lower-level variables.  However in practice a user may only need
+the optimal upper-level variables.  If true, this may allow for a
+faster response.  Additionally, there's another solution response that
+may be of interest to analysts, where solution consists of the optimal
+upper-level variables along with other diagnostic information related
+to the solution of the lower-level problem.  For example, in some cases
+the dual of the lower-level problem can be used, so the solution to the
+problem could include the dual lower-level variables.  Note that these
+considerations also generalize to multi-level problems, where a user may
+wish to specify the requirements for a solution value (e.g. needing only
+the first- and second-level variables in a tri-level problem).
+
+The following subsections describe ideas for expressing and solving multi-level programs, with the
+goal of defining some alternatives that the PAO developers can discuss while refining the 
+problem representations in PAO.
+
+\subsection{TODO}
+
+TODO: Discuss a new modeling idea here.
+
+\subsection{TODO}
+
+TODO: Discuss a new modeling idea here.
+
+\SANDsection{Special Cases}
+\label{sec:problems}
 

### 2019-05-13 doc/bilevel/paper.tex
Drafing some modeling ideas

@@
 
-\subsection{TODO}
+\subsection{Using Set Notation}
 
-TODO: Discuss a new modeling idea here.
+IDEA: Declare the SubModel using arguments that define the lower-level variables whose
+optimal value constrains the upper-level variables.
 
-\subsection{TODO}
+\begin{qlisting}
+model = ConcreteModel()
 
-TODO: Discuss a new modeling idea here.
+model.x = Var(bounds=(1,2))
+model.v = Var(bounds=(1,2))
+model.y = Var(bounds=(1,2))
+
+model.o = Objective(expr=model.x + model.v + model.y)
+model.c = Constraint(expr=model.x + model.v >= 1.5)
+
+model.sub = SubModel(fixed=[model.x, model.v], y=model.y)
+
+# NOTE - model.sub.y is implicitly defined
+model.sub.w = Var(bounds=(-1,1))
+
+model.sub.o = Objective(expr=model.x + model.sub.w, sense=maximize)
+model.sub.c = Constraint(expr=model.sub.y + model.sub.w <= 2.5)
+\end{qlisting}
+
+\noindent PROS
+\begin{itemize}
+\item More explicit than pyomo.bilevel
+\end{itemize}
+CONS
+\begin{itemize}
+\item Users may not appreciate this difference and use model.y instead of model.sub.y.  Can we robustly catch errors when that happens?
+\item May not be able to capture complex relationships between upper- and lower-level variables
+\item May not be useful for defining pessimistic relationships.
+\end{itemize}
+
+\subsection{Using Set Notation}
+
+IDEA:  What if the SubModel declaration defined a set, which can be explicitly used to
+define a constraint (similar to~(\ref{eqn:bilevel})?
+
+TODO ...
 

### 2019-05-16 doc/bilevel/paper.tex
Updated first section to include pessmistic bilevel

@@
 P(x) & = & \argmin_{y \in Y}    & f(x,y)\\
- & &                            & g(x,y) \leq 0
+ & &                            & g(x,y) \leq 0.
 \end{array}
@@
 
-\textbf{TODO: Describe pessimistic problems here.}
+Alternatively, a \textit{pessimistic} bilevel problem is when \eqref{eqn:bilevel} is expressed as
+
+\begin{equation}
+\label{eqn:bilevel_pess}
+\begin{array}{ll}
+\min_{x \in X, y \in Y}   & F(x,y)\\
+\st                 & G(x,y) \leq 0\\
+                    & \forall y \in P(x)
+\end{array}
+\end{equation}
+where 
+\[
+\begin{array}{llll}
+P(x) & = & \argmin_{y \in Y}    & f(x,y)\\
+ & &                            & g(x,y) \leq 0.
+\end{array}
+\]
+This formulation ensures that the selection value for $y$ will in fact maximize $F(x,y)$; in other words, the upper-level problem must be feasible for all values $y$ including the least favorable outcome of the lower-level problem $P(x)$.
+Unless otherwise stated, the canonical bilevel problem in the subsequent examples will be derived from the \textit{optimistic} approach presented in \eqref{eqn:bilevel}.
 

### 2019-05-16 doc/bilevel/paper.tex
updates to bilevel

@@
 
-The \code{pyomo.bilevel} package uses the \code{SubModel} component to declare a block that defines a lower-level problem. 
-For example, consider:
+\subsection{Introduction to \code{pao}}
+
+The Python package \code{pao} is a bilevel mathematical programming library that has dependencies on \code{pyomo} for syntactical representation of bilevel problems.
+Similar to \code{pyomo} mathematical problems, \code{pao} problems can be optimized with solvers that are written either in Python or in compiled, low-level languages.
+
+The \code{pao} package uses the \code{SubModel} component in \code{pyomo} to declare a block that defines a lower-level problem. 
+For example, consider the bilevel problem:
+\begin{equation}
+\label{eqn:bilevel_ex1}
+\begin{array}{ll}
+\min_{v \in [1,2], x \in [1,2]}   & v + x + y\\
+\st                 & x + v \geq 1.5\\
+                    & y \in P(v,x)
+\end{array}
+\end{equation}
+where 
+\[
+\begin{array}{llll}
+P(v,x) & = & \argmin_{w \in [-1,1], y \in [1,2]}    & x + w\\
+ & &                            &  y + w \leq 2.5
+\end{array}
+\]
+which can be represented in \code{pao} as:
+
 \begin{qlisting}
@@
 
-model.o = Objective(expr=model.x + model.sub.y + model.v)
+model.o = Objective(expr=model.x + model.sub.y + model.v, sense=minimize)
 model.c = Constraint(expr=model.x + model.v >= 1.5)
@@
 
-While this syntax mimics the structure of (\ref{eqn:bilevel-compact}), it adds additional abstractions that make it difficult to interpret:
+While this syntax mimics the abstract structure of \eqref{eqn:bilevel-compact} -- as represented directly in \eqref{eqn:bilevel_ex1} --  it adds additional abstractions that make it difficult to interpret:
 \begin{itemize}

### 2019-05-17 doc/bilevel/paper.tex
updates to bilevel
Updates to include a new pyomo example for bilevel

@@
 \min_{v \in [1,2], x \in [1,2]}   & v + x + y\\
-\st                 & x + v \geq 1.5\\
+\st                 & v + x\geq 1.5\\
                     & y \in P(v,x)
@@
 \end{qlisting}
-This Pyomo model defines four variables, \code{x}, \code{v}, \code{sub.y} and \code{sub.w}.
-Variables \code{x} and \code{v} are declared in the upper-level
+This Pyomo model defines four variables,  \code{v}, \code{x}, \code{sub.w} and \code{sub.y}.
+Variables \code{v}  and \code{x} are declared in the upper-level
 problem, and \code{v} only appears in the upper-level problem.
-Variables \code{sub.y} and \code{sub.w} are declared in the submodel.
+Variables \code{sub.w} and \code{sub.y} are declared in the submodel.
 However, note that the \code{sub.y} variable appears in the upper-level
@@
 \end{qlisting}
+QUESTION: Is the last line correct here???\\
 
@@
 TODO ...
+
+\subsection{Using New Pyomo Notation}
+
+For example, consider the bilevel problem:
+\begin{equation}
+\label{eqn:bilevel_ex1}
+\begin{array}{ll}
+\min_{v \in [1,2], x \in [1,2]}   & v + x + y\\
+\st                 & v + x\geq 1.5\\
+                    & y \in P(v,x)
+\end{array}
+\end{equation}
+where 
+\[
+\begin{array}{llll}
+P(v,x) & = & \argmin_{w \in [-1,1], y \in [1,2]}    & x + w\\
+ & &                            &  y + w \leq 2.5
+\end{array}
+\]
+which can be represented in \code{pao} pseudo-code as:
+
+\begin{qlisting}
+from pyomo.environ import *
+import pyomo.kernel as pmo
+from pao.bilevel import *
+model = ConcreteModel()
+model.x = Var(bounds=(1,2))
+model.v = Var(bounds=(1,2))
+model.y = Var(bounds=(1,2))
+model.w = Var(bounds=(-1,1))
+model.upper = SubModel(vars=[model.x,model.v])
+model.lower = SubModel(vars = [model.y,model.w], parent = upper, 
+		type = optimistic, argmin = model.y)
+
+model.upper = Objective(expr=model.x + model.y + model.v, sense=minimize)
+
+model.o = Objective(expr=model.x + model.y + model.v, sense=minimize)
+model.c = Constraint(expr=model.x + model.v >= 1.5)
+model.lower.o = Objective(expr=model.x+model.w, sense=maximize)
+model.lower.c = Constraint(expr=model.y + model.w <= 2.5)
+\end{qlisting}
+
+\noindent PROS
+\begin{itemize}
+\item Requiring that each level of the problem be defined in a SubModel() block, which enables each submodel to be solved independently
+\item The user must specify important aspects about that level of the problem when instantiating the SubModel()
+\end{itemize}
+CONS
+\begin{itemize}
+\item Implicitly fixes the vars not specified in the vars argument to the SubModel() to their current value
+\item Variables have to be defined prior to instantiating the submodel; the user may want to define these on the SubModel() block
+\end{itemize}
 

### 2019-05-17 doc/bilevel/paper.tex
Updates to include a new pyomo example for bilevel

@@
 from pyomo.environ import *
-import pyomo.kernel as pmo
 from pao.bilevel import *

### 2019-06-18 doc/bilevel/paper.tex
Editing pessimistic formulation to include both the general version and the min max version, with the Wiesemann et al 2013 comparison of the two

@@
 
-Alternatively, a \textit{pessimistic} bilevel problem is when \eqref{eqn:bilevel} is expressed as
-
-\begin{equation}
-\label{eqn:bilevel_pess}
-\begin{array}{ll}
-\min_{x \in X, y \in Y}   & F(x,y)\\
-\st                 & G(x,y) \leq 0\\
-                    & \forall y \in P(x)
-\end{array}
-\end{equation}
-where 
-\[
-\begin{array}{llll}
-P(x) & = & \argmin_{y \in Y}    & f(x,y)\\
- & &                            & g(x,y) \leq 0.
-\end{array}
-\]
-This formulation ensures that the selection value for $y$ will in fact maximize $F(x,y)$; in other words, the upper-level problem must be feasible for all values $y$ including the least favorable outcome of the lower-level problem $P(x)$.
-Unless otherwise stated, the canonical bilevel problem in the subsequent examples will be derived from the \textit{optimistic} approach presented in \eqref{eqn:bilevel}.
-
 Many authors use a more compact form to express (\ref{eqn:bilevel}):
@@
 This notation defines an implied constraint by not explicitly referencing the solution to the lower-level problem.
+
+Alternatively, a \textit{pessimistic} bilevel problem is when \eqref{eqn:bilevel} is expressed as
+
+\begin{equation}
+\label{eqn:bilevel_pess}
+\begin{array}{lll}
+\min_{x \in X}   & F(x)\\
+\st                 & G(x,y) \leq 0 &\forall y \in P(x)
+\end{array}
+\end{equation}
+where 
+\[
+\begin{array}{llll}
+P(x) & = & \argmin_{y \in Y}    & f(x,y)\\
+ & &                            & g(x,y) \leq 0.
+\end{array}
+\]
+This formulation ensures that the upper-level problem is feasible for all values $y$ including the least favorable outcome of the lower-level problem $P(x)$. 
+It is common to see the pessimistic problem written as
+\begin{equation}
+\label{eqn:bilevel_pess_minmax}
+\min_{x \in X}\max_{y \in P(x)} H(x,y).
+\end{equation}
+Though formulation (\ref{eqn:bilevel_pess_minmax}) makes clear the selected value for $y$ will in fact maximize $H(x,y)$, note that it is actually less general than (\ref{eqn:bilevel_pess}) since we can rewrite it in the form of (\ref{eqn:bilevel_pess}), as
+\[
+\begin{array}{lll}
+\min_{x, \tau} &\tau \\
+\st & \tau \geq H(x, y) & \forall y \in P(x) \\
+& x \in X,
+\end{array}
+\]
+but it is not always possible to tranform problems in the form of (\ref{eqn:bilevel_pess_minmax}) to (\ref{eqn:bilevel_pess}) without definting functions $H(x,y)$ and $f(x,y)$ on the extended real line \cite{WiesemannTKR2013}.
+Unless otherwise stated, the canonical bilevel problem in the subsequent examples will be derived from the \textit{optimistic} approach presented in \eqref{eqn:bilevel}.
 

### 2021-04-13 doc/sand/abstract.tex
Adding SAND document

PAO is a Python\sphinxhyphen{}based package for Adversarial Optimization.  The goal of
this package is to provide a general modeling and analysis capability for
bilevel, trilevel and other multilevel optimization forms that express
adversarial dynamics.  PAO integrates two different modeling abstractions:
\begin{enumerate}
\sphinxsetlistlabels{\arabic}{enumi}{enumii}{}{.}%
\item {}
\sphinxstylestrong{Algebraic models} extend the modeling concepts in the
\sphinxhref{https://github.com/Pyomo/pyomo}{Pyomo} algebraic modeling language
to express problems with an intuitive algebraic syntax.  Thus, we
expect that this modeling abstraction will commonly be used by PAO
end\sphinxhyphen{}users.

\item {}
\sphinxstylestrong{Compact models} express objective and constraints in a manner
that is typically used to express the mathematical form of these
problems (e.g. using vector and matrix data types).  PAO defines
custom \sphinxstyleemphasis{Multilevel Problem Representations} (MPRs) that simplify the
implementation of solvers for bilevel, trilevel and other multilevel
optimization problems.

\end{enumerate}

### 2021-04-13 doc/sand/acknow.tex
Adding SAND document

We thank Richard Chen, John Siirola and Jean-Paul Watson for collaborating
on the development of \textbf{pyomo.bilevel}.  The PAO software is
derived from ideas we prototyped in \textbf{pyomo.bilevel}.  We also
thank members of the SECURE Grand Challenge LDRD who provided input on
design and application of PAO, especially Bryan Arguello, Jared Gearhart,
Ali Pinar, and Cindy Phillips .

This work was funded by the U.S. Department of Energy National Nuclear
Security Administration's Office of Defense Nuclear Nonproliferation
Research and Development (NA-22). Sandia National Laboratories is a
multimission laboratory managed and operated by National Technology
and Engineering Solutions of Sandia, LLC., a wholly owned subsidiary
of Honeywell International, Inc., for the U.S. Department of Energy's
National Nuclear Security Administration under contract DE-NA0003525.
This paper describes objective technical results and analysis.
Any subjective views or opinions that might be expressed in the paper
do not necessarily represent the views of the U.S. Department of Energy
or the United States Government.

### 2021-04-13 doc/sand/pao.tex
Adding SAND document

\if 0
%% Generated by Sphinx.
\def\sphinxdocclass{report}
\documentclass[letterpaper,10pt,english]{sphinxmanual}
\ifdefined\pdfpxdimen
   \let\sphinxpxdimen\pdfpxdimen\else\newdimen\sphinxpxdimen
\fi \sphinxpxdimen=.75bp\relax

\PassOptionsToPackage{warn}{textcomp}
\usepackage[utf8]{inputenc}
\ifdefined\DeclareUnicodeCharacter
% support both utf8 and utf8x syntaxes
  \ifdefined\DeclareUnicodeCharacterAsOptional
    \def\sphinxDUC#1{\DeclareUnicodeCharacter{"#1}}
  \else
    \let\sphinxDUC\DeclareUnicodeCharacter
  \fi
  \sphinxDUC{00A0}{\nobreakspace}
  \sphinxDUC{2500}{\sphinxunichar{2500}}
  \sphinxDUC{2502}{\sphinxunichar{2502}}
  \sphinxDUC{2514}{\sphinxunichar{2514}}
  \sphinxDUC{251C}{\sphinxunichar{251C}}
  \sphinxDUC{2572}{\textbackslash}
\fi
\usepackage{cmap}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amstext}
\usepackage{babel}

\usepackage{times}
\expandafter\ifx\csname T@LGR\endcsname\relax
\else
% LGR was declared as font encoding
  \substitutefont{LGR}{\rmdefault}{cmr}
  \substitutefont{LGR}{\sfdefault}{cmss}
  \substitutefont{LGR}{\ttdefault}{cmtt}
\fi
\expandafter\ifx\csname T@X2\endcsname\relax
  \expandafter\ifx\csname T@T2A\endcsname\relax
  \else
  % T2A was declared as font encoding
    \substitutefont{T2A}{\rmdefault}{cmr}
    \substitutefont{T2A}{\sfdefault}{cmss}
    \substitutefont{T2A}{\ttdefault}{cmtt}
  \fi
\else
% X2 was declared as font encoding
  \substitutefont{X2}{\rmdefault}{cmr}
  \substitutefont{X2}{\sfdefault}{cmss}
  \substitutefont{X2}{\ttdefault}{cmtt}
\fi

\usepackage[Bjarne]{fncychap}
\usepackage{sphinx}

\fvset{fontsize=\small}
\usepackage{geometry}

% Include hyperref last.
\usepackage{hyperref}
% Fix anchor placement for figures with captions.
\usepackage{hypcap}% it must be loaded after hyperref.
% Set up styles of URL: it should be placed after hyperref.
\urlstyle{same}

\usepackage{sphinxmessages}
\setcounter{tocdepth}{0}

\title{PAO}
\date{Apr 13, 2021}
\release{1.0}
\author{PAO Developers}
\newcommand{\sphinxlogo}{\vbox{}}
\renewcommand{\releasename}{Release}
\makeindex
\begin{document}

\pagestyle{empty}
\sphinxmaketitle
\pagestyle{plain}
\sphinxtableofcontents
\pagestyle{normal}
\phantomsection\label{\detokenize{index::doc}}

\subsection{Introduction}

PAO is a Python\sphinxhyphen{}based package for Adversarial Optimization.  The goal of
this package is to provide a general modeling and analysis capability for
bilevel, trilevel and other multilevel optimization forms that express
adversarial dynamics.  PAO integrates two different modeling abstractions:
\begin{enumerate}
\sphinxsetlistlabels{\arabic}{enumi}{enumii}{}{.}%
\item {} 
\sphinxstylestrong{Algebraic models} extend the modeling concepts in the
\sphinxhref{https://github.com/Pyomo/pyomo}{Pyomo} algebraic modeling language
to express problems with an intuitive algebraic syntax.  Thus, we
expect that this modeling abstraction will commonly be used by PAO
end\sphinxhyphen{}users.

\item {} 
\sphinxstylestrong{Compact models} express objective and constraints in a manner
that is typically used to express the mathematical form of these
problems (e.g. using vector and matrix data types).  PAO defines
custom \sphinxstyleemphasis{Multilevel Problem Representations} (MPRs) that simplify the
implementation of solvers for bilevel, trilevel and other multilevel
optimization problems.

\end{enumerate}

\fi

\section{Installation}
\label{\detokenize{installation:installation}}\label{\detokenize{installation::doc}}
PAO currently supports the following versions of Python:
\begin{itemize}
\item {} 
CPython: 3.7, 3.8, 3.9

\end{itemize}

\subsection{Using PIP}
\label{\detokenize{installation:using-pip}}
The standard utility for installing Python packages is \sphinxstylestrong{pip}.
You can install the latest release of PAO by executing the following:

\begin{sphinxVerbatim}[commandchars=\\\{\}]
pip install pao
\end{sphinxVerbatim}

You can also use \sphinxstylestrong{pip} to install from the PAO software repository.
For example, the master branch can be installed as follows:

\begin{sphinxVerbatim}[commandchars=\\\{\}]
python \PYGZhy{}m pip install https://github.com/or\PYGZhy{}fusion/pao.git
\end{sphinxVerbatim}

\subsection{Using CONDA}
\label{\detokenize{installation:using-conda}}
The \sphinxstylestrong{conda} utility can be used to install the latest release of PAO
using the \sphinxcode{\sphinxupquote{conda\sphinxhyphen{}forge}} channel:

\begin{sphinxVerbatim}[commandchars=\\\{\}]
conda install \PYGZhy{}c conda\PYGZhy{}forge pao
\end{sphinxVerbatim}

\subsection{Using GIT}
\label{\detokenize{installation:using-git}}
PAO can be installed by cloning the PAO software repostory and
then directly installing the software.  For example, the master
branch can be installed as follows:

\begin{sphinxVerbatim}[commandchars=\\\{\}]
git clone https://github.com/or\PYGZhy{}fusion/pao.git
cd pao
python setup.py develop
\end{sphinxVerbatim}

\subsection{Conditional Dependencies}
\label{\detokenize{installation:conditional-dependencies}}
Both \sphinxstylestrong{conda} and \sphinxstylestrong{pip} can be used to install the third\sphinxhyphen{}party packages
that are needed to model problems with PAO.  We recommend \sphinxstylestrong{conda}
because it has better support for optimization solver packages.

PAO intrinsically depends on \sphinxhref{https://github.com/Pyomo/pyomo}{Pyomo},
both for the representation of algebraic problems but also for
interfaces to numerical optimizers used by PAO solvers.  \sphinxhref{https://github.com/Pyomo/pyomo}{Pyomo} is installed with PAO, but the Pyomo
website \sphinxcite{pyomoweb} and GitHub site \sphinxcite{pyomogithub} provide additional
resources for installing Pyomo and related software.

PAO and \sphinxhref{https://github.com/Pyomo/pyomo}{Pyomo} have conditional
dependencies on a variety of third\sphinxhyphen{}party packages, including Python
packages like scipy, numpy and optimization solvers.  Optimization solvers
are particularly important, and a commercial optimizer may be needed to
analyze complex, real\sphinxhyphen{}world applications.

The following optimizers are used to test the PAO solvers:
\begin{itemize}
\item {} 
\sphinxhref{https://www.gnu.org/software/glpk/}{glpk} \sphinxhyphen{} An open\sphinxhyphen{}source mixed\sphinxhyphen{}integer linear programming solver

\item {} 
\sphinxhref{https://github.com/coin-or/Cbc}{cbc} \sphinxhyphen{} An open\sphinxhyphen{}source mixed\sphinxhyphen{}integer linear programming solver

\item {} 
\sphinxhref{https://github.com/coin-or/Ipopt}{ipopt} \sphinxhyphen{} An open\sphinxhyphen{}source interior point optimizer for continuous problems

\end{itemize}

Additionally, PAO can interface with optimization solvers at \sphinxhref{https://neos-server.org/neos/}{NEOS}.

\section{Overview}
\label{\detokenize{overview:overview}}\label{\detokenize{overview::doc}}
Many planning situations involve the analysis of a hierarchy of
decision\sphinxhyphen{}makers with competing objectives.  For example, policy decisions
are made at different levels of a government, each of which has a
different objective and decision space.  Similarly, robust planning
against adversaries is often modeled with a 2\sphinxhyphen{}level hierarchy, where
the defensive planner makes decisions that account for adversarial
response.  Multilevel optimization techniques partition control over
decision variables amongst the levels.  Decisions at each level of
the hierarchy may be constrained by decisions at other levels, and the
objectives for each level may account for decisions made at other levels.
The PAO library is designed to express this structure in a manner that is
intuitive to users, and which facilitates the application of appropriate
optimization solvers.  In particular, PAO extends the modeling concepts
in the \sphinxhref{https://github.com/Pyomo/pyomo}{Pyomo} algebraic modeling
language to express problems with an intuitive algebraic syntax.

However, data structures used to represent algebraic models are often
poorly suited for complex numerical algorithms.  Consequently, PAO
supports \sphinxstyleemphasis{distinct} representations for algebraic models (using \sphinxhref{https://github.com/Pyomo/pyomo}{Pyomo}) and compact problem representations
that express objective and constraints using vector and matrix data types.
Currently, PAO includes several \sphinxstyleemphasis{Multilevel Problem Representations}
(MPRs) that represent multilevel optimization problems with an explicit,
compact representation that simplifies the implementation of solvers
for bilevel, trilevel and other multilevel optimization problems.

For example, PAO includes a compact representation for linear bilevel
problems, \sphinxcode{\sphinxupquote{LinearMultilevelProblem}}.  Several solvers have been
developed for problems expressed as a \sphinxcode{\sphinxupquote{LinearMultilevelProblem}},
including the big\sphinxhyphen{}M method proposed by Fortuny\sphinxhyphen{}Amat and McCarl
\sphinxcite{fortunymccarl}.  PAO can similarly express linear bilevel problems in
Pyomo using a \sphinxcode{\sphinxupquote{Submodel}} component, which was previously introduced in
the \sphinxstylestrong{pyomo.bilevel} package \sphinxcite{pyomobookii}.  Further, these Pyomo models
can be automatically converted to the compact \sphinxcode{\sphinxupquote{LinearMultilevelProblem}}
representation and optimized with solvers tailored for that representation.

The use of independent problem representations in this manner has
several implications for PAO.  First, this design facilitates the
development of solvers for algebraic modeling languages like Pyomo
that are intrinsically more robust.  Compact representations like
\sphinxcode{\sphinxupquote{LinearMultilevelProblem}} enable the development of solvers using
natural operations (e.g. matrix\sphinxhyphen{}vector multiplication).  Thus, we expect
these solvers to be more robust and easier to maintain when compared to
solvers developed using Pyomo data structures (e.g. expression trees).
Additionally, the conversion of a Pyomo representation to a compact
representation provides a context for verifying that the Pyomo model
represents the intended problem form.

Second, this design facilitates the development of different problem
representations that may or may not be inter\sphinxhyphen{}operable.  Although PAO
is derived from initial efforts with \sphinxstylestrong{pyomo.bilevel}, it has evolved
from an extension of Pyomo’s modeling capability to be a library that
is synergistic with Pyomo but not strictly dependent on it.

The following sections provide detailed descriptions of these
representations that are illustrated with increasingly complex examples,
including
\begin{itemize}
\item {} 
bilevel

\item {} 
bilevel with multiple lower\sphinxhyphen{}levels

\item {} 
trilevel

\end{itemize}

Additionally, the following sections describe the setup of linear and
quadratic problems, and the transformations that can be applied to them
in PAO.

\begin{sphinxadmonition}{note}{Note:}
We do not restrict the description of PAO representations to
models that PAO can solve. Rather, a goal of this documentation
to illustrate the breadth of the adversarial optimization problems
that can be expressed with PAO, thereby motivating the development
of new solvers for new classes of problems.
\end{sphinxadmonition}

\begin{sphinxadmonition}{note}{Note:}
Although the modeling abstractions in PAO are suitable for both
pessimistic and optimistic multilevel optimization formulations,
PAO currently assumes that all problems are expressed and solved
in an an optimistic form.  Thus, when there multiple solutions in
lower\sphinxhyphen{}level problems, the choice of the solution is determined by
the follower not the leader.  This convention reflects the fact that
solution techniques for pessimistic forms are not well\sphinxhyphen{}developed.
\end{sphinxadmonition}

\section{Simple Examples}
\label{\detokenize{examples:simple-examples}}\label{\detokenize{examples::doc}}
We illustrate the PAO algebraic and compact representations with a series
of simple examples.  Consider the following bilevel problem introduced
by Bard \sphinxcite{bard98} (example 5.1.1):
 \begin{equation*}
 \begin{array}{lll}
 \min_{x \geq 0} & x - 4 y & \\
 \textrm{s.t.} & \min_{y \geq 0} & y\\
 & \textrm{s.t.} & -x -y \leq -3\\
 & & -2 x + y \leq 0\\
 & & 2 x + y \leq 12\\
 & & 3 x - 2 y \leq 4
 \end{array}
 \end{equation*}
This problem has has linear upper\sphinxhyphen{} and lower\sphinxhyphen{}level problems with different
objectives in each level.  Thus, this problem can be represented in
PAO using a Pyomo model representation or the \sphinxcode{\sphinxupquote{LinearMultilevelProblem}}
representation.

\subsection{Using Pyomo}
\label{\detokenize{examples:using-pyomo}}
The following python script defines a bilevel problem in Pyomo:

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{k+kn}{import} \PYG{n+nn}{pyomo}\PYG{n+nn}{.}\PYG{n+nn}{environ} \PYG{k}{as} \PYG{n+nn}{pe}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{k+kn}{from} \PYG{n+nn}{pao}\PYG{n+nn}{.}\PYG{n+nn}{pyomo} \PYG{k+kn}{import} \PYG{o}{*}

\PYG{g+go}{\PYGZsh{} Create a model object}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{ConcreteModel}\PYG{p}{(}\PYG{p}{)}

\PYG{g+go}{\PYGZsh{} Define decision variables}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{x} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Var}\PYG{p}{(}\PYG{n}{bounds}\PYG{o}{=}\PYG{p}{(}\PYG{l+m+mi}{0}\PYG{p}{,}\PYG{k+kc}{None}\PYG{p}{)}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{y} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Var}\PYG{p}{(}\PYG{n}{bounds}\PYG{o}{=}\PYG{p}{(}\PYG{l+m+mi}{0}\PYG{p}{,}\PYG{k+kc}{None}\PYG{p}{)}\PYG{p}{)}

\PYG{g+go}{\PYGZsh{} Define the upper\PYGZhy{}level objective}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{o} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Objective}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=}\PYG{n}{M}\PYG{o}{.}\PYG{n}{x} \PYG{o}{\PYGZhy{}} \PYG{l+m+mi}{4}\PYG{o}{*}\PYG{n}{M}\PYG{o}{.}\PYG{n}{y}\PYG{p}{)}

\PYG{g+go}{\PYGZsh{} Create a SubModel component to declare a lower\PYGZhy{}level problem}
\PYG{g+go}{\PYGZsh{} The variable M.x is fixed in this lower\PYGZhy{}level problem}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L} \PYG{o}{=} \PYG{n}{SubModel}\PYG{p}{(}\PYG{n}{fixed}\PYG{o}{=}\PYG{n}{M}\PYG{o}{.}\PYG{n}{x}\PYG{p}{)}

\PYG{g+go}{\PYGZsh{} Define the lower\PYGZhy{}level objective}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{o} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Objective}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=}\PYG{n}{M}\PYG{o}{.}\PYG{n}{y}\PYG{p}{)}

\PYG{g+go}{\PYGZsh{} Define lower\PYGZhy{}level constraints}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{c1} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Constraint}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=}   \PYG{o}{\PYGZhy{}}\PYG{n}{M}\PYG{o}{.}\PYG{n}{x} \PYG{o}{\PYGZhy{}}   \PYG{n}{M}\PYG{o}{.}\PYG{n}{y} \PYG{o}{\PYGZlt{}}\PYG{o}{=} \PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{3}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{c2} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Constraint}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=} \PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{2}\PYG{o}{*}\PYG{n}{M}\PYG{o}{.}\PYG{n}{x} \PYG{o}{+}   \PYG{n}{M}\PYG{o}{.}\PYG{n}{y} \PYG{o}{\PYGZlt{}}\PYG{o}{=}  \PYG{l+m+mi}{0}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{c3} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Constraint}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=}  \PYG{l+m+mi}{2}\PYG{o}{*}\PYG{n}{M}\PYG{o}{.}\PYG{n}{x} \PYG{o}{+}   \PYG{n}{M}\PYG{o}{.}\PYG{n}{y} \PYG{o}{\PYGZlt{}}\PYG{o}{=} \PYG{l+m+mi}{12}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{c4} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Constraint}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=}  \PYG{l+m+mi}{3}\PYG{o}{*}\PYG{n}{M}\PYG{o}{.}\PYG{n}{x} \PYG{o}{\PYGZhy{}} \PYG{l+m+mi}{2}\PYG{o}{*}\PYG{n}{M}\PYG{o}{.}\PYG{n}{y} \PYG{o}{\PYGZlt{}}\PYG{o}{=}  \PYG{l+m+mi}{4}\PYG{p}{)}

\PYG{g+go}{\PYGZsh{} Create a solver and apply it}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{k}{with} \PYG{n}{Solver}\PYG{p}{(}\PYG{l+s+s1}{\PYGZsq{}}\PYG{l+s+s1}{pao.pyomo.FA}\PYG{l+s+s1}{\PYGZsq{}}\PYG{p}{)} \PYG{k}{as} \PYG{n}{solver}\PYG{p}{:}
\PYG{g+gp}{... }    \PYG{n}{results} \PYG{o}{=} \PYG{n}{solver}\PYG{o}{.}\PYG{n}{solve}\PYG{p}{(}\PYG{n}{M}\PYG{p}{)}

\PYG{g+go}{\PYGZsh{} The final solution is loaded into the model}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n+nb}{print}\PYG{p}{(}\PYG{n}{M}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{value}\PYG{p}{)}
\PYG{g+go}{4.0}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n+nb}{print}\PYG{p}{(}\PYG{n}{M}\PYG{o}{.}\PYG{n}{y}\PYG{o}{.}\PYG{n}{value}\PYG{p}{)}
\PYG{g+go}{4.0}
\end{sphinxVerbatim}

The \sphinxcode{\sphinxupquote{SubModel}} component defines a Pyomo block object within which the
lower\sphinxhyphen{}level problem is declared.  The \sphinxcode{\sphinxupquote{fixed}} option is used to specify
the upper\sphinxhyphen{}level variables whose value is fixed in the lower\sphinxhyphen{}level problem.

The \sphinxcode{\sphinxupquote{pao.pyomo.FA}} uses the method for solving linear bilevel
programs with big\sphinxhyphen{}M relaxations described by Fortuny\sphinxhyphen{}Amat and McCarl
\sphinxcite{fortunymccarl}.  In this example, the default big\sphinxhyphen{}M value is used, but
in general a user may need to explore suitable values for their problem.
By default, the final values of the upper\sphinxhyphen{} and lower\sphinxhyphen{}level variables
are loaded back into the Pyomo model.  The \sphinxcode{\sphinxupquote{results}} object contains
information about the problem, the solver and the solver status.

\subsection{Using LinearMultilevelProblem}
\label{\detokenize{examples:using-linearmultilevelproblem}}
Bilevel linear problems can also be represented using the
\sphinxcode{\sphinxupquote{LinearMultilevelProblem}} class.  This class provides a simple mechanism
for organizing data for variables, objectives and linear constraints.  The following
examples illustrate the use of \sphinxcode{\sphinxupquote{LinearMultilevelProblem}} for Bard’s example 5.1.1 described
above, but this representation can naturally be used to express multi\sphinxhyphen{}level problems as well
as problems with multiple lower\sphinxhyphen{}levels.

\paragraph{Using Numpy and Scipy Data}
\label{\detokenize{examples:using-numpy-and-scipy-data}}
The following python script defines a bilevel problem using \sphinxcode{\sphinxupquote{LinearMultilevelProblem}} with
numpy and scipy data:

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{k+kn}{import} \PYG{n+nn}{numpy} \PYG{k}{as} \PYG{n+nn}{np}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{k+kn}{from} \PYG{n+nn}{scipy}\PYG{n+nn}{.}\PYG{n+nn}{sparse} \PYG{k+kn}{import} \PYG{n}{coo\PYGZus{}matrix}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{k+kn}{from} \PYG{n+nn}{pao}\PYG{n+nn}{.}\PYG{n+nn}{mpr} \PYG{k+kn}{import} \PYG{o}{*}

\PYG{g+go}{\PYGZsh{} Create a model object}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M} \PYG{o}{=} \PYG{n}{LinearMultilevelProblem}\PYG{p}{(}\PYG{p}{)}

\PYG{g+go}{\PYGZsh{} Declare the upper\PYGZhy{} and lower\PYGZhy{}levels, including the number of decision\PYGZhy{}variables}
\PYG{g+go}{\PYGZsh{}  nxR=1 means there will be 1 real\PYGZhy{}valued decision variable}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U} \PYG{o}{=} \PYG{n}{M}\PYG{o}{.}\PYG{n}{add\PYGZus{}upper}\PYG{p}{(}\PYG{n}{nxR}\PYG{o}{=}\PYG{l+m+mi}{1}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L} \PYG{o}{=} \PYG{n}{U}\PYG{o}{.}\PYG{n}{add\PYGZus{}lower}\PYG{p}{(}\PYG{n}{nxR}\PYG{o}{=}\PYG{l+m+mi}{1}\PYG{p}{)}

\PYG{g+go}{\PYGZsh{} Declare the bounds on the decision variables}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{lower\PYGZus{}bounds} \PYG{o}{=} \PYG{n}{np}\PYG{o}{.}\PYG{n}{array}\PYG{p}{(}\PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{]}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{lower\PYGZus{}bounds} \PYG{o}{=} \PYG{n}{np}\PYG{o}{.}\PYG{n}{array}\PYG{p}{(}\PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{]}\PYG{p}{)}

\PYG{g+go}{\PYGZsh{} Declare the upper\PYGZhy{}level objective}
\PYG{g+go}{\PYGZsh{}   U.c[X] is the array of coefficients in the objective for variables in level X}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{c}\PYG{p}{[}\PYG{n}{U}\PYG{p}{]} \PYG{o}{=} \PYG{n}{np}\PYG{o}{.}\PYG{n}{array}\PYG{p}{(}\PYG{p}{[}\PYG{l+m+mi}{1}\PYG{p}{]}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{c}\PYG{p}{[}\PYG{n}{L}\PYG{p}{]} \PYG{o}{=} \PYG{n}{np}\PYG{o}{.}\PYG{n}{array}\PYG{p}{(}\PYG{p}{[}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{4}\PYG{p}{]}\PYG{p}{)}

\PYG{g+go}{\PYGZsh{} Declare the lower\PYGZhy{}level objective, which has no upper\PYGZhy{}level decision\PYGZhy{}variables}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L}\PYG{o}{.}\PYG{n}{c}\PYG{p}{[}\PYG{n}{L}\PYG{p}{]} \PYG{o}{=} \PYG{n}{np}\PYG{o}{.}\PYG{n}{array}\PYG{p}{(}\PYG{p}{[}\PYG{l+m+mi}{1}\PYG{p}{]}\PYG{p}{)}

\PYG{g+go}{\PYGZsh{} Declare the lower\PYGZhy{}level constraints}
\PYG{g+go}{\PYGZsh{}   L.A[X] is the matrix coefficients in the constraints for variables in level X}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L}\PYG{o}{.}\PYG{n}{A}\PYG{p}{[}\PYG{n}{U}\PYG{p}{]} \PYG{o}{=} \PYG{n}{coo\PYGZus{}matrix}\PYG{p}{(}\PYG{p}{(}\PYG{n}{np}\PYG{o}{.}\PYG{n}{array}\PYG{p}{(}\PYG{p}{[}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{1}\PYG{p}{,} \PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{2}\PYG{p}{,} \PYG{l+m+mi}{2}\PYG{p}{,} \PYG{l+m+mi}{3}\PYG{p}{]}\PYG{p}{)}\PYG{p}{,}  \PYG{c+c1}{\PYGZsh{} Coefficients}
\PYG{g+gp}{... }                   \PYG{p}{(}\PYG{n}{np}\PYG{o}{.}\PYG{n}{array}\PYG{p}{(}\PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{,} \PYG{l+m+mi}{1}\PYG{p}{,} \PYG{l+m+mi}{2}\PYG{p}{,} \PYG{l+m+mi}{3}\PYG{p}{]}\PYG{p}{)}\PYG{p}{,}     \PYG{c+c1}{\PYGZsh{} Row indices of matrix entries}
\PYG{g+gp}{... }                    \PYG{n}{np}\PYG{o}{.}\PYG{n}{array}\PYG{p}{(}\PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{,} \PYG{l+m+mi}{0}\PYG{p}{,} \PYG{l+m+mi}{0}\PYG{p}{,} \PYG{l+m+mi}{0}\PYG{p}{]}\PYG{p}{)}\PYG{p}{)}\PYG{p}{)}\PYG{p}{)}   \PYG{c+c1}{\PYGZsh{} Column indices of matrix entries}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L}\PYG{o}{.}\PYG{n}{A}\PYG{p}{[}\PYG{n}{L}\PYG{p}{]} \PYG{o}{=} \PYG{n}{coo\PYGZus{}matrix}\PYG{p}{(}\PYG{p}{(}\PYG{n}{np}\PYG{o}{.}\PYG{n}{array}\PYG{p}{(}\PYG{p}{[}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{1}\PYG{p}{,} \PYG{l+m+mi}{1}\PYG{p}{,} \PYG{l+m+mi}{1}\PYG{p}{,} \PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{2}\PYG{p}{]}\PYG{p}{)}\PYG{p}{,}
\PYG{g+gp}{... }                   \PYG{p}{(}\PYG{n}{np}\PYG{o}{.}\PYG{n}{array}\PYG{p}{(}\PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{,} \PYG{l+m+mi}{1}\PYG{p}{,} \PYG{l+m+mi}{2}\PYG{p}{,} \PYG{l+m+mi}{3}\PYG{p}{]}\PYG{p}{)}\PYG{p}{,}
\PYG{g+gp}{... }                    \PYG{n}{np}\PYG{o}{.}\PYG{n}{array}\PYG{p}{(}\PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{,} \PYG{l+m+mi}{0}\PYG{p}{,} \PYG{l+m+mi}{0}\PYG{p}{,} \PYG{l+m+mi}{0}\PYG{p}{]}\PYG{p}{)}\PYG{p}{)}\PYG{p}{)}\PYG{p}{)}

\PYG{g+go}{\PYGZsh{} Declare the constraint right\PYGZhy{}hand\PYGZhy{}side}
\PYG{g+go}{\PYGZsh{}   By default, constraints are inequalities, so these are upper\PYGZhy{}bounds}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L}\PYG{o}{.}\PYG{n}{b} \PYG{o}{=} \PYG{n}{np}\PYG{o}{.}\PYG{n}{array}\PYG{p}{(}\PYG{p}{[}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{3}\PYG{p}{,} \PYG{l+m+mi}{0}\PYG{p}{,} \PYG{l+m+mi}{12}\PYG{p}{,} \PYG{l+m+mi}{4}\PYG{p}{]}\PYG{p}{)}

\PYG{g+go}{\PYGZsh{} Create a solver and apply it}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{k}{with} \PYG{n}{Solver}\PYG{p}{(}\PYG{l+s+s1}{\PYGZsq{}}\PYG{l+s+s1}{pao.mpr.FA}\PYG{l+s+s1}{\PYGZsq{}}\PYG{p}{)} \PYG{k}{as} \PYG{n}{solver}\PYG{p}{:}
\PYG{g+gp}{... }   \PYG{n}{results} \PYG{o}{=} \PYG{n}{solver}\PYG{o}{.}\PYG{n}{solve}\PYG{p}{(}\PYG{n}{M}\PYG{p}{)}

\PYG{g+go}{\PYGZsh{} The final solution is loaded into the model}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n+nb}{print}\PYG{p}{(}\PYG{n}{U}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{values}\PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{]}\PYG{p}{)}
\PYG{g+go}{4.0}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n+nb}{print}\PYG{p}{(}\PYG{n}{L}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{values}\PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{]}\PYG{p}{)}
\PYG{g+go}{4.0}
\end{sphinxVerbatim}

The \sphinxcode{\sphinxupquote{U}} and \sphinxcode{\sphinxupquote{L}} objects represent the upper\sphinxhyphen{} and lower\sphinxhyphen{}level
respectively.  When declaring these objects, the user specifies the number
of real, integer and binary variables.  The remaining declarations assume
that these variables are used in that order.  Thus, there is a single
declaration for the objective coefficients, \sphinxcode{\sphinxupquote{c}}, which is an array
with values for each of the declared variables.  However, the upper\sphinxhyphen{}
and lower\sphinxhyphen{}level objective coefficients are separately declared for
the upper\sphinxhyphen{} and lower\sphinxhyphen{}level variables by indexing \sphinxcode{\sphinxupquote{c}} with \sphinxcode{\sphinxupquote{U}} and
\sphinxcode{\sphinxupquote{L}} respectively.  This example includes declarations for the upper\sphinxhyphen{}
and lower\sphinxhyphen{}level variable bounds and objective coefficients.  There are no
upper\sphinxhyphen{}level constraints, so only the lower\sphinxhyphen{}level constriants are declared.

Note that the syntax for specifying solvers is analogous to that used
with Pyomo models.  The same solver options are available.  The principle
difference is the specification of the solver name that indicates the
expected type of the model that will be solved.

\paragraph{Using Python Lists and Dictionaries}
\label{\detokenize{examples:using-python-lists-and-dictionaries}}
Although the constraint matrices are dense in this example, the
\sphinxcode{\sphinxupquote{coo\_matrix}} is used to illustrate the general support for sparse data.
The \sphinxcode{\sphinxupquote{LinearMultilevelProblem}} class also supports a simpler syntax
where dense arrays can be specified and Python lists and sparse matrices
can be specified with Python tuple and dictionary objects:

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{k+kn}{from} \PYG{n+nn}{pao}\PYG{n+nn}{.}\PYG{n+nn}{mpr} \PYG{k+kn}{import} \PYG{o}{*}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M} \PYG{o}{=} \PYG{n}{LinearMultilevelProblem}\PYG{p}{(}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U} \PYG{o}{=} \PYG{n}{M}\PYG{o}{.}\PYG{n}{add\PYGZus{}upper}\PYG{p}{(}\PYG{n}{nxR}\PYG{o}{=}\PYG{l+m+mi}{1}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L} \PYG{o}{=} \PYG{n}{U}\PYG{o}{.}\PYG{n}{add\PYGZus{}lower}\PYG{p}{(}\PYG{n}{nxR}\PYG{o}{=}\PYG{l+m+mi}{1}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{lower\PYGZus{}bounds} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{lower\PYGZus{}bounds} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{]}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{c}\PYG{p}{[}\PYG{n}{U}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{1}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{c}\PYG{p}{[}\PYG{n}{L}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{4}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L}\PYG{o}{.}\PYG{n}{c}\PYG{p}{[}\PYG{n}{L}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{1}\PYG{p}{]}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L}\PYG{o}{.}\PYG{n}{A}\PYG{p}{[}\PYG{n}{U}\PYG{p}{]} \PYG{o}{=} \PYG{p}{(}\PYG{l+m+mi}{4}\PYG{p}{,}\PYG{l+m+mi}{1}\PYG{p}{)}\PYG{p}{,} \PYG{p}{\PYGZob{}}\PYG{p}{(}\PYG{l+m+mi}{0}\PYG{p}{,}\PYG{l+m+mi}{0}\PYG{p}{)}\PYG{p}{:}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{1}\PYG{p}{,} \PYG{p}{(}\PYG{l+m+mi}{1}\PYG{p}{,}\PYG{l+m+mi}{0}\PYG{p}{)}\PYG{p}{:}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{2}\PYG{p}{,} \PYG{p}{(}\PYG{l+m+mi}{2}\PYG{p}{,}\PYG{l+m+mi}{0}\PYG{p}{)}\PYG{p}{:}\PYG{l+m+mi}{2}\PYG{p}{,} \PYG{p}{(}\PYG{l+m+mi}{3}\PYG{p}{,}\PYG{l+m+mi}{0}\PYG{p}{)}\PYG{p}{:} \PYG{l+m+mi}{3}\PYG{p}{\PYGZcb{}}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L}\PYG{o}{.}\PYG{n}{A}\PYG{p}{[}\PYG{n}{L}\PYG{p}{]} \PYG{o}{=} \PYG{p}{(}\PYG{l+m+mi}{4}\PYG{p}{,}\PYG{l+m+mi}{1}\PYG{p}{)}\PYG{p}{,} \PYG{p}{\PYGZob{}}\PYG{p}{(}\PYG{l+m+mi}{0}\PYG{p}{,}\PYG{l+m+mi}{0}\PYG{p}{)}\PYG{p}{:}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{1}\PYG{p}{,} \PYG{p}{(}\PYG{l+m+mi}{1}\PYG{p}{,}\PYG{l+m+mi}{0}\PYG{p}{)}\PYG{p}{:} \PYG{l+m+mi}{1}\PYG{p}{,} \PYG{p}{(}\PYG{l+m+mi}{2}\PYG{p}{,}\PYG{l+m+mi}{0}\PYG{p}{)}\PYG{p}{:}\PYG{l+m+mi}{1}\PYG{p}{,} \PYG{p}{(}\PYG{l+m+mi}{3}\PYG{p}{,}\PYG{l+m+mi}{0}\PYG{p}{)}\PYG{p}{:}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{2}\PYG{p}{\PYGZcb{}}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L}\PYG{o}{.}\PYG{n}{b} \PYG{o}{=} \PYG{p}{[}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{3}\PYG{p}{,} \PYG{l+m+mi}{0}\PYG{p}{,} \PYG{l+m+mi}{12}\PYG{p}{,} \PYG{l+m+mi}{4}\PYG{p}{]}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{k}{with} \PYG{n}{Solver}\PYG{p}{(}\PYG{l+s+s1}{\PYGZsq{}}\PYG{l+s+s1}{pao.mpr.FA}\PYG{l+s+s1}{\PYGZsq{}}\PYG{p}{)} \PYG{k}{as} \PYG{n}{solver}\PYG{p}{:}
\PYG{g+gp}{... }   \PYG{n}{results} \PYG{o}{=} \PYG{n}{solver}\PYG{o}{.}\PYG{n}{solve}\PYG{p}{(}\PYG{n}{M}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n+nb}{print}\PYG{p}{(}\PYG{n}{U}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{values}\PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{]}\PYG{p}{)}
\PYG{g+go}{4.0}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n+nb}{print}\PYG{p}{(}\PYG{n}{L}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{values}\PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{]}\PYG{p}{)}
\PYG{g+go}{4.0}
\end{sphinxVerbatim}

When specifying a sparse matrix, a tuple is provided (e.g. for
\sphinxcode{\sphinxupquote{L.A{[}U{]}}}).  The first element is a 2\sphinxhyphen{}tuple that defines the shape
of the matrix, and the second element is a dictionary that defines the
non\sphinxhyphen{}zero values in the sparse matrix.

Similarly, a list\sphinxhyphen{}of\sphinxhyphen{}lists syntax can be used to specify dense matrices:

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{k+kn}{from} \PYG{n+nn}{pao}\PYG{n+nn}{.}\PYG{n+nn}{mpr} \PYG{k+kn}{import} \PYG{o}{*}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M} \PYG{o}{=} \PYG{n}{LinearMultilevelProblem}\PYG{p}{(}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U} \PYG{o}{=} \PYG{n}{M}\PYG{o}{.}\PYG{n}{add\PYGZus{}upper}\PYG{p}{(}\PYG{n}{nxR}\PYG{o}{=}\PYG{l+m+mi}{1}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L} \PYG{o}{=} \PYG{n}{U}\PYG{o}{.}\PYG{n}{add\PYGZus{}lower}\PYG{p}{(}\PYG{n}{nxR}\PYG{o}{=}\PYG{l+m+mi}{1}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{lower\PYGZus{}bounds} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{lower\PYGZus{}bounds} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{]}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{c}\PYG{p}{[}\PYG{n}{U}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{1}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{c}\PYG{p}{[}\PYG{n}{L}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{4}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L}\PYG{o}{.}\PYG{n}{c}\PYG{p}{[}\PYG{n}{L}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{1}\PYG{p}{]}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L}\PYG{o}{.}\PYG{n}{A}\PYG{p}{[}\PYG{n}{U}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{p}{[}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{1}\PYG{p}{]}\PYG{p}{,} \PYG{p}{[}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{2}\PYG{p}{]}\PYG{p}{,} \PYG{p}{[}\PYG{l+m+mi}{2}\PYG{p}{]}\PYG{p}{,} \PYG{p}{[}\PYG{l+m+mi}{3}\PYG{p}{]}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L}\PYG{o}{.}\PYG{n}{A}\PYG{p}{[}\PYG{n}{L}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{p}{[}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{1}\PYG{p}{]}\PYG{p}{,} \PYG{p}{[}\PYG{l+m+mi}{1}\PYG{p}{]}\PYG{p}{,} \PYG{p}{[}\PYG{l+m+mi}{1}\PYG{p}{]}\PYG{p}{,} \PYG{p}{[}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{2}\PYG{p}{]}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L}\PYG{o}{.}\PYG{n}{b} \PYG{o}{=} \PYG{p}{[}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{3}\PYG{p}{,} \PYG{l+m+mi}{0}\PYG{p}{,} \PYG{l+m+mi}{12}\PYG{p}{,} \PYG{l+m+mi}{4}\PYG{p}{]}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{k}{with} \PYG{n}{Solver}\PYG{p}{(}\PYG{l+s+s1}{\PYGZsq{}}\PYG{l+s+s1}{pao.mpr.FA}\PYG{l+s+s1}{\PYGZsq{}}\PYG{p}{)} \PYG{k}{as} \PYG{n}{solver}\PYG{p}{:}
\PYG{g+gp}{... }   \PYG{n}{results} \PYG{o}{=} \PYG{n}{solver}\PYG{o}{.}\PYG{n}{solve}\PYG{p}{(}\PYG{n}{M}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n+nb}{print}\PYG{p}{(}\PYG{n}{U}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{values}\PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{]}\PYG{p}{)}
\PYG{g+go}{4.0}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n+nb}{print}\PYG{p}{(}\PYG{n}{L}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{values}\PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{]}\PYG{p}{)}
\PYG{g+go}{4.0}
\end{sphinxVerbatim}

When native Python data values are used to initialize a
\sphinxcode{\sphinxupquote{LinearMultilevelProblem}}, they are converted into numpy and scipy
data types.  This facilitates the use of \sphinxcode{\sphinxupquote{LinearMultilevelProblem}}
objects for defining numerical solvers using a consistent, convenient
API for numerical operations (e.g. matrix\sphinxhyphen{}vector multiplication).

\section{Pyomo Models}
\label{\detokenize{representations/pyomo:pyomo-models}}\label{\detokenize{representations/pyomo::doc}}
PAO can be used to express linear and quadratic problems in \sphinxhref{https://github.com/Pyomo/pyomo}{Pyomo} using a {\hyperref[\detokenize{reference/pyomo:pao.pyomo.components.SubModel}]{\sphinxcrossref{\sphinxcode{\sphinxupquote{SubModel}}}}}
component, which was previously introduced in the \sphinxstylestrong{pyomo.bilevel}
package \sphinxcite{pyomobookii}.  \sphinxhref{https://github.com/Pyomo/pyomo}{Pyomo}
represents optimization models using an model objects that are
annotated with modeling component objects.  Thus, {\hyperref[\detokenize{reference/pyomo:pao.pyomo.components.SubModel}]{\sphinxcrossref{\sphinxcode{\sphinxupquote{SubModel}}}}}
component is a simple extension of the modeling concepts in \sphinxhref{https://github.com/Pyomo/pyomo}{Pyomo}.

\begin{sphinxadmonition}{hint}{Hint:}
Advanced Pyomo users will realize that the PAO {\hyperref[\detokenize{reference/pyomo:pao.pyomo.components.SubModel}]{\sphinxcrossref{\sphinxcode{\sphinxupquote{SubModel}}}}} component
is a special case of the Pyomo \sphinxcode{\sphinxupquote{Block}} component, which is used to
structure the expression of Pyomo models.
\end{sphinxadmonition}

A {\hyperref[\detokenize{reference/pyomo:pao.pyomo.components.SubModel}]{\sphinxcrossref{\sphinxcode{\sphinxupquote{SubModel}}}}} component creates a context for expressing the
objective and constraints in a lower\sphinxhyphen{}level model.  Pyomo models can
include nested and parallel {\hyperref[\detokenize{reference/pyomo:pao.pyomo.components.SubModel}]{\sphinxcrossref{\sphinxcode{\sphinxupquote{SubModel}}}}} components to express
complex multilevel problems.

\subsection{Bilevel Examples}
\label{\detokenize{representations/pyomo:bilevel-examples}}
Consider the following bilevel problem:
\label{equation:sand:eq-pao1} \begin{equation*}
 \textbf{Model PAO1}
 \begin{array}{ll}
 \min_{x\in[2,6],y} & x + 3 z \\
 \textrm{s.t.} & x + y = 10\\
 & \begin{array}{lll}
   \max_{z \geq 0} & z &\\
   \textrm{s.t.} & x+z &\leq 8\\
   & x + 4 z &\geq 8\\
   & x + 2 z &\leq 13
   \end{array}
 \end{array}
 \end{equation*}
This problem has has linear upper\sphinxhyphen{} and lower\sphinxhyphen{}level problems with different
objectives in each level.

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{ConcreteModel}\PYG{p}{(}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{x} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Var}\PYG{p}{(}\PYG{n}{bounds}\PYG{o}{=}\PYG{p}{(}\PYG{l+m+mi}{2}\PYG{p}{,}\PYG{l+m+mi}{6}\PYG{p}{)}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{y} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Var}\PYG{p}{(}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L} \PYG{o}{=} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{pyomo}\PYG{o}{.}\PYG{n}{SubModel}\PYG{p}{(}\PYG{n}{fixed}\PYG{o}{=}\PYG{p}{[}\PYG{n}{M}\PYG{o}{.}\PYG{n}{x}\PYG{p}{,}\PYG{n}{M}\PYG{o}{.}\PYG{n}{y}\PYG{p}{]}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{z} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Var}\PYG{p}{(}\PYG{n}{bounds}\PYG{o}{=}\PYG{p}{(}\PYG{l+m+mi}{0}\PYG{p}{,}\PYG{k+kc}{None}\PYG{p}{)}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{o} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Objective}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=}\PYG{n}{M}\PYG{o}{.}\PYG{n}{x} \PYG{o}{+} \PYG{l+m+mi}{3}\PYG{o}{*}\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{z}\PYG{p}{,} \PYG{n}{sense}\PYG{o}{=}\PYG{n}{pe}\PYG{o}{.}\PYG{n}{minimize}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{c} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Constraint}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=} \PYG{n}{M}\PYG{o}{.}\PYG{n}{x} \PYG{o}{+} \PYG{n}{M}\PYG{o}{.}\PYG{n}{y} \PYG{o}{==} \PYG{l+m+mi}{10}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{o} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Objective}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=}\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{z}\PYG{p}{,} \PYG{n}{sense}\PYG{o}{=}\PYG{n}{pe}\PYG{o}{.}\PYG{n}{maximize}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{c1} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Constraint}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=} \PYG{n}{M}\PYG{o}{.}\PYG{n}{x} \PYG{o}{+} \PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{z} \PYG{o}{\PYGZlt{}}\PYG{o}{=} \PYG{l+m+mi}{8}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{c2} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Constraint}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=} \PYG{n}{M}\PYG{o}{.}\PYG{n}{x} \PYG{o}{+} \PYG{l+m+mi}{4}\PYG{o}{*}\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{z} \PYG{o}{\PYGZgt{}}\PYG{o}{=} \PYG{l+m+mi}{8}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{c3} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Constraint}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=} \PYG{n}{M}\PYG{o}{.}\PYG{n}{x} \PYG{o}{+} \PYG{l+m+mi}{2}\PYG{o}{*}\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{z} \PYG{o}{\PYGZlt{}}\PYG{o}{=} \PYG{l+m+mi}{13}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{opt} \PYG{o}{=} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{Solver}\PYG{p}{(}\PYG{l+s+s2}{\PYGZdq{}}\PYG{l+s+s2}{pao.pyomo.FA}\PYG{l+s+s2}{\PYGZdq{}}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{results} \PYG{o}{=} \PYG{n}{opt}\PYG{o}{.}\PYG{n}{solve}\PYG{p}{(}\PYG{n}{M}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n+nb}{print}\PYG{p}{(}\PYG{n}{M}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{value}\PYG{p}{,} \PYG{n}{M}\PYG{o}{.}\PYG{n}{y}\PYG{o}{.}\PYG{n}{value}\PYG{p}{,} \PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{z}\PYG{o}{.}\PYG{n}{value}\PYG{p}{)}
\PYG{g+go}{6.0 4.0 2.0}
\end{sphinxVerbatim}

This example illustrates the flexibility of Pyomo representations in PAO:
\begin{itemize}
\item {} 
Each level can express different objectives with different senses

\item {} 
Variables can be bounded or unbounded

\item {} 
Equality and inequality constraints can be expressed

\end{itemize}

The {\hyperref[\detokenize{reference/pyomo:pao.pyomo.components.SubModel}]{\sphinxcrossref{\sphinxcode{\sphinxupquote{SubModel}}}}} component is used to define a logically separate
optimization model that includes variables that are dynamically fixed
by upper\sphinxhyphen{}level problems.  All of the Pyomo objective and constraint
declarations contained in the {\hyperref[\detokenize{reference/pyomo:pao.pyomo.components.SubModel}]{\sphinxcrossref{\sphinxcode{\sphinxupquote{SubModel}}}}} declaration are included
in the sub\sphinxhyphen{}problem that it defines, even if they are nested in Pyomo
\sphinxcode{\sphinxupquote{Block}} components.  The {\hyperref[\detokenize{reference/pyomo:pao.pyomo.components.SubModel}]{\sphinxcrossref{\sphinxcode{\sphinxupquote{SubModel}}}}} component also declares
which variables are fixed in a lower\sphinxhyphen{}level problem.  The value of the
\sphinxtitleref{fixed} argument is a Pyomo variable or a list of variables.  For example,
the following model expresses the upper\sphinxhyphen{}level variables with a single
variable, \sphinxtitleref{M.x}, which is fixed in the {\hyperref[\detokenize{reference/pyomo:pao.pyomo.components.SubModel}]{\sphinxcrossref{\sphinxcode{\sphinxupquote{SubModel}}}}} declaration:

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{ConcreteModel}\PYG{p}{(}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{x} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Var}\PYG{p}{(}\PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{,}\PYG{l+m+mi}{1}\PYG{p}{]}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{x}\PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{]}\PYG{o}{.}\PYG{n}{setlb}\PYG{p}{(}\PYG{l+m+mi}{2}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{x}\PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{]}\PYG{o}{.}\PYG{n}{setub}\PYG{p}{(}\PYG{l+m+mi}{6}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L} \PYG{o}{=} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{pyomo}\PYG{o}{.}\PYG{n}{SubModel}\PYG{p}{(}\PYG{n}{fixed}\PYG{o}{=}\PYG{n}{M}\PYG{o}{.}\PYG{n}{x}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{z} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Var}\PYG{p}{(}\PYG{n}{bounds}\PYG{o}{=}\PYG{p}{(}\PYG{l+m+mi}{0}\PYG{p}{,}\PYG{k+kc}{None}\PYG{p}{)}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{o} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Objective}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=}\PYG{n}{M}\PYG{o}{.}\PYG{n}{x}\PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{]} \PYG{o}{+} \PYG{l+m+mi}{3}\PYG{o}{*}\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{z}\PYG{p}{,} \PYG{n}{sense}\PYG{o}{=}\PYG{n}{pe}\PYG{o}{.}\PYG{n}{minimize}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{c} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Constraint}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=} \PYG{n}{M}\PYG{o}{.}\PYG{n}{x}\PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{]} \PYG{o}{+} \PYG{n}{M}\PYG{o}{.}\PYG{n}{x}\PYG{p}{[}\PYG{l+m+mi}{1}\PYG{p}{]} \PYG{o}{==} \PYG{l+m+mi}{10}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{o} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Objective}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=}\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{z}\PYG{p}{,} \PYG{n}{sense}\PYG{o}{=}\PYG{n}{pe}\PYG{o}{.}\PYG{n}{maximize}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{c1} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Constraint}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=} \PYG{n}{M}\PYG{o}{.}\PYG{n}{x}\PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{]} \PYG{o}{+} \PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{z} \PYG{o}{\PYGZlt{}}\PYG{o}{=} \PYG{l+m+mi}{8}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{c2} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Constraint}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=} \PYG{n}{M}\PYG{o}{.}\PYG{n}{x}\PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{]} \PYG{o}{+} \PYG{l+m+mi}{4}\PYG{o}{*}\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{z} \PYG{o}{\PYGZgt{}}\PYG{o}{=} \PYG{l+m+mi}{8}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{c3} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Constraint}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=} \PYG{n}{M}\PYG{o}{.}\PYG{n}{x}\PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{]} \PYG{o}{+} \PYG{l+m+mi}{2}\PYG{o}{*}\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{z} \PYG{o}{\PYGZlt{}}\PYG{o}{=} \PYG{l+m+mi}{13}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{opt} \PYG{o}{=} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{Solver}\PYG{p}{(}\PYG{l+s+s2}{\PYGZdq{}}\PYG{l+s+s2}{pao.pyomo.FA}\PYG{l+s+s2}{\PYGZdq{}}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{results} \PYG{o}{=} \PYG{n}{opt}\PYG{o}{.}\PYG{n}{solve}\PYG{p}{(}\PYG{n}{M}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n+nb}{print}\PYG{p}{(}\PYG{n}{M}\PYG{o}{.}\PYG{n}{x}\PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{]}\PYG{o}{.}\PYG{n}{value}\PYG{p}{,} \PYG{n}{M}\PYG{o}{.}\PYG{n}{x}\PYG{p}{[}\PYG{l+m+mi}{1}\PYG{p}{]}\PYG{o}{.}\PYG{n}{value}\PYG{p}{,} \PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{z}\PYG{o}{.}\PYG{n}{value}\PYG{p}{)}
\PYG{g+go}{6.0 4.0 2.0}
\end{sphinxVerbatim}

Although a lower\sphinxhyphen{}level problem is logically a separate optimization model,
you cannot use a {\hyperref[\detokenize{reference/pyomo:pao.pyomo.components.SubModel}]{\sphinxcrossref{\sphinxcode{\sphinxupquote{SubModel}}}}} that is defined with a separate Pyomo
model object.  Pyomo implicitly requires that all variables used in
objective and constraint expressions are attributes of the same Pyomo model.
However, the location of variable declarations in a Pyomo model does \sphinxstyleemphasis{not} denote their
use in upper\sphinxhyphen{} or lower\sphinxhyphen{}level problems.  For example, consider the following
model that re\sphinxhyphen{}expresses the previous problem:

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{ConcreteModel}\PYG{p}{(}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{x} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Var}\PYG{p}{(}\PYG{n}{bounds}\PYG{o}{=}\PYG{p}{(}\PYG{l+m+mi}{2}\PYG{p}{,}\PYG{l+m+mi}{6}\PYG{p}{)}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{y} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Var}\PYG{p}{(}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{z} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Var}\PYG{p}{(}\PYG{n}{bounds}\PYG{o}{=}\PYG{p}{(}\PYG{l+m+mi}{0}\PYG{p}{,}\PYG{k+kc}{None}\PYG{p}{)}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{o} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Objective}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=}\PYG{n}{M}\PYG{o}{.}\PYG{n}{x} \PYG{o}{+} \PYG{l+m+mi}{3}\PYG{o}{*}\PYG{n}{M}\PYG{o}{.}\PYG{n}{z}\PYG{p}{,} \PYG{n}{sense}\PYG{o}{=}\PYG{n}{pe}\PYG{o}{.}\PYG{n}{minimize}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{c} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Constraint}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=} \PYG{n}{M}\PYG{o}{.}\PYG{n}{x} \PYG{o}{+} \PYG{n}{M}\PYG{o}{.}\PYG{n}{y} \PYG{o}{==} \PYG{l+m+mi}{10}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L} \PYG{o}{=} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{pyomo}\PYG{o}{.}\PYG{n}{SubModel}\PYG{p}{(}\PYG{n}{fixed}\PYG{o}{=}\PYG{p}{[}\PYG{n}{M}\PYG{o}{.}\PYG{n}{x}\PYG{p}{,}\PYG{n}{M}\PYG{o}{.}\PYG{n}{y}\PYG{p}{]}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{o} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Objective}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=}\PYG{n}{M}\PYG{o}{.}\PYG{n}{z}\PYG{p}{,} \PYG{n}{sense}\PYG{o}{=}\PYG{n}{pe}\PYG{o}{.}\PYG{n}{maximize}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{c1} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Constraint}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=} \PYG{n}{M}\PYG{o}{.}\PYG{n}{x} \PYG{o}{+} \PYG{n}{M}\PYG{o}{.}\PYG{n}{z} \PYG{o}{\PYGZlt{}}\PYG{o}{=} \PYG{l+m+mi}{8}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{c2} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Constraint}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=} \PYG{n}{M}\PYG{o}{.}\PYG{n}{x} \PYG{o}{+} \PYG{l+m+mi}{4}\PYG{o}{*}\PYG{n}{M}\PYG{o}{.}\PYG{n}{z} \PYG{o}{\PYGZgt{}}\PYG{o}{=} \PYG{l+m+mi}{8}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{c3} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Constraint}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=} \PYG{n}{M}\PYG{o}{.}\PYG{n}{x} \PYG{o}{+} \PYG{l+m+mi}{2}\PYG{o}{*}\PYG{n}{M}\PYG{o}{.}\PYG{n}{z} \PYG{o}{\PYGZlt{}}\PYG{o}{=} \PYG{l+m+mi}{13}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{opt} \PYG{o}{=} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{Solver}\PYG{p}{(}\PYG{l+s+s2}{\PYGZdq{}}\PYG{l+s+s2}{pao.pyomo.FA}\PYG{l+s+s2}{\PYGZdq{}}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{results} \PYG{o}{=} \PYG{n}{opt}\PYG{o}{.}\PYG{n}{solve}\PYG{p}{(}\PYG{n}{M}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n+nb}{print}\PYG{p}{(}\PYG{n}{M}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{value}\PYG{p}{,} \PYG{n}{M}\PYG{o}{.}\PYG{n}{y}\PYG{o}{.}\PYG{n}{value}\PYG{p}{,} \PYG{n}{M}\PYG{o}{.}\PYG{n}{z}\PYG{o}{.}\PYG{n}{value}\PYG{p}{)}
\PYG{g+go}{6.0 4.0 2.0}
\end{sphinxVerbatim}

Note that \sphinxstyleemphasis{all} of the decision variables are declared outside of the
{\hyperref[\detokenize{reference/pyomo:pao.pyomo.components.SubModel}]{\sphinxcrossref{\sphinxcode{\sphinxupquote{SubModel}}}}} component, even though the variable \sphinxcode{\sphinxupquote{M.z}} is a
lower\sphinxhyphen{}level variable.  The declarations of {\hyperref[\detokenize{reference/pyomo:pao.pyomo.components.SubModel}]{\sphinxcrossref{\sphinxcode{\sphinxupquote{SubModel}}}}} components
defines the mathematical role of all decision variables in a Pyomo model.
As this example illustrates, the specification of a bilevel problem can
be simplified if all variables are expressed at once.

Finally, we observe that PAO’s Pyomo representation only works with a
subset of the many different modeling components that are supported in
\sphinxhref{https://github.com/Pyomo/pyomo}{Pyomo}:
\begin{itemize}
\item {} 
\sphinxcode{\sphinxupquote{Set}} \sphinxhyphen{} Set declarations

\item {} 
\sphinxcode{\sphinxupquote{Param}} \sphinxhyphen{} Parameter declarations

\item {} 
\sphinxcode{\sphinxupquote{Var}} \sphinxhyphen{} Variable declarations

\item {} 
\sphinxcode{\sphinxupquote{Block}} \sphinxhyphen{} Defines a subset of a model

\item {} 
\sphinxcode{\sphinxupquote{Objective}} \sphinxhyphen{} Define a model objective

\item {} 
\sphinxcode{\sphinxupquote{Constraint}} \sphinxhyphen{} Define model constraints

\end{itemize}

Additional Pyomo modeling components will be added to PAO as motivating
applications arise and as suitable solvers become available.

\subsection{Multilevel Examples}
\label{\detokenize{representations/pyomo:multilevel-examples}}
Multilevel problems can be easily expressed with Pyomo using multiple declarations
of {\hyperref[\detokenize{reference/pyomo:pao.pyomo.components.SubModel}]{\sphinxcrossref{\sphinxcode{\sphinxupquote{SubModel}}}}}.

\paragraph{Multiple Lower Levels}
\label{\detokenize{representations/pyomo:multiple-lower-levels}}
Consider the following bilevel problem that
extends the \sphinxstylestrong{PAO1} model to include two equivalent lower\sphinxhyphen{}levels:
\label{equation:sand:eq-pao2} \begin{equation*}
 \textbf{Model PAO2}
 \begin{array}{ll}
 \min_{x\in[2,6],y} & x + 3 z_1 + 3 z_2 \\
 \textrm{s.t.} & x + y = 10\\
 & \begin{array}{lll}
   \max_{z_1 \geq 0} & z_1 &\\
   \textrm{s.t.} & x+z_1 &\leq 8\\
   & x + 4 z_1 &\geq 8\\
   & x + 2 z_1 &\leq 13\\
   \end{array}\\
 & \begin{array}{lll}
   \max_{z_2 \geq 0} & z_2 &\\
   \textrm{s.t.} & y+z_2 &\leq 8\\
   & y + 4 z_2 &\geq 8\\
   & y + 2 z_2 &\leq 13\\
   \end{array}\\
 \end{array}
 \end{equation*}
The \sphinxstylestrong{PAO2} model can be expressed in Pyomo as follows:

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{ConcreteModel}\PYG{p}{(}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{x} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Var}\PYG{p}{(}\PYG{n}{bounds}\PYG{o}{=}\PYG{p}{(}\PYG{l+m+mi}{2}\PYG{p}{,}\PYG{l+m+mi}{6}\PYG{p}{)}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{y} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Var}\PYG{p}{(}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{z} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Var}\PYG{p}{(}\PYG{p}{[}\PYG{l+m+mi}{1}\PYG{p}{,}\PYG{l+m+mi}{2}\PYG{p}{]}\PYG{p}{,} \PYG{n}{bounds}\PYG{o}{=}\PYG{p}{(}\PYG{l+m+mi}{0}\PYG{p}{,}\PYG{k+kc}{None}\PYG{p}{)}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{o} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Objective}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=}\PYG{n}{M}\PYG{o}{.}\PYG{n}{x} \PYG{o}{+} \PYG{l+m+mi}{3}\PYG{o}{*}\PYG{n}{M}\PYG{o}{.}\PYG{n}{z}\PYG{p}{[}\PYG{l+m+mi}{1}\PYG{p}{]}\PYG{o}{+}\PYG{l+m+mi}{3}\PYG{o}{*}\PYG{n}{M}\PYG{o}{.}\PYG{n}{z}\PYG{p}{[}\PYG{l+m+mi}{2}\PYG{p}{]}\PYG{p}{,} \PYG{n}{sense}\PYG{o}{=}\PYG{n}{pe}\PYG{o}{.}\PYG{n}{minimize}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{c} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Constraint}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=} \PYG{n}{M}\PYG{o}{.}\PYG{n}{x} \PYG{o}{+} \PYG{n}{M}\PYG{o}{.}\PYG{n}{y} \PYG{o}{==} \PYG{l+m+mi}{10}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L1} \PYG{o}{=} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{pyomo}\PYG{o}{.}\PYG{n}{SubModel}\PYG{p}{(}\PYG{n}{fixed}\PYG{o}{=}\PYG{p}{[}\PYG{n}{M}\PYG{o}{.}\PYG{n}{x}\PYG{p}{]}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L1}\PYG{o}{.}\PYG{n}{o} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Objective}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=}\PYG{n}{M}\PYG{o}{.}\PYG{n}{z}\PYG{p}{[}\PYG{l+m+mi}{1}\PYG{p}{]}\PYG{p}{,} \PYG{n}{sense}\PYG{o}{=}\PYG{n}{pe}\PYG{o}{.}\PYG{n}{maximize}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L1}\PYG{o}{.}\PYG{n}{c1} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Constraint}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=} \PYG{n}{M}\PYG{o}{.}\PYG{n}{x} \PYG{o}{+} \PYG{n}{M}\PYG{o}{.}\PYG{n}{z}\PYG{p}{[}\PYG{l+m+mi}{1}\PYG{p}{]} \PYG{o}{\PYGZlt{}}\PYG{o}{=} \PYG{l+m+mi}{8}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L1}\PYG{o}{.}\PYG{n}{c2} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Constraint}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=} \PYG{n}{M}\PYG{o}{.}\PYG{n}{x} \PYG{o}{+} \PYG{l+m+mi}{4}\PYG{o}{*}\PYG{n}{M}\PYG{o}{.}\PYG{n}{z}\PYG{p}{[}\PYG{l+m+mi}{1}\PYG{p}{]} \PYG{o}{\PYGZgt{}}\PYG{o}{=} \PYG{l+m+mi}{8}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L1}\PYG{o}{.}\PYG{n}{c3} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Constraint}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=} \PYG{n}{M}\PYG{o}{.}\PYG{n}{x} \PYG{o}{+} \PYG{l+m+mi}{2}\PYG{o}{*}\PYG{n}{M}\PYG{o}{.}\PYG{n}{z}\PYG{p}{[}\PYG{l+m+mi}{1}\PYG{p}{]} \PYG{o}{\PYGZlt{}}\PYG{o}{=} \PYG{l+m+mi}{13}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L2} \PYG{o}{=} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{pyomo}\PYG{o}{.}\PYG{n}{SubModel}\PYG{p}{(}\PYG{n}{fixed}\PYG{o}{=}\PYG{p}{[}\PYG{n}{M}\PYG{o}{.}\PYG{n}{y}\PYG{p}{]}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L2}\PYG{o}{.}\PYG{n}{o} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Objective}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=}\PYG{n}{M}\PYG{o}{.}\PYG{n}{z}\PYG{p}{[}\PYG{l+m+mi}{2}\PYG{p}{]}\PYG{p}{,} \PYG{n}{sense}\PYG{o}{=}\PYG{n}{pe}\PYG{o}{.}\PYG{n}{maximize}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L2}\PYG{o}{.}\PYG{n}{c1} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Constraint}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=} \PYG{n}{M}\PYG{o}{.}\PYG{n}{y} \PYG{o}{+} \PYG{n}{M}\PYG{o}{.}\PYG{n}{z}\PYG{p}{[}\PYG{l+m+mi}{2}\PYG{p}{]} \PYG{o}{\PYGZlt{}}\PYG{o}{=} \PYG{l+m+mi}{8}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L2}\PYG{o}{.}\PYG{n}{c2} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Constraint}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=} \PYG{n}{M}\PYG{o}{.}\PYG{n}{y} \PYG{o}{+} \PYG{l+m+mi}{4}\PYG{o}{*}\PYG{n}{M}\PYG{o}{.}\PYG{n}{z}\PYG{p}{[}\PYG{l+m+mi}{2}\PYG{p}{]} \PYG{o}{\PYGZgt{}}\PYG{o}{=} \PYG{l+m+mi}{8}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L2}\PYG{o}{.}\PYG{n}{c3} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Constraint}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=} \PYG{n}{M}\PYG{o}{.}\PYG{n}{y} \PYG{o}{+} \PYG{l+m+mi}{2}\PYG{o}{*}\PYG{n}{M}\PYG{o}{.}\PYG{n}{z}\PYG{p}{[}\PYG{l+m+mi}{2}\PYG{p}{]} \PYG{o}{\PYGZlt{}}\PYG{o}{=} \PYG{l+m+mi}{13}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{opt} \PYG{o}{=} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{Solver}\PYG{p}{(}\PYG{l+s+s2}{\PYGZdq{}}\PYG{l+s+s2}{pao.pyomo.FA}\PYG{l+s+s2}{\PYGZdq{}}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{results} \PYG{o}{=} \PYG{n}{opt}\PYG{o}{.}\PYG{n}{solve}\PYG{p}{(}\PYG{n}{M}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n+nb}{print}\PYG{p}{(}\PYG{n}{M}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{value}\PYG{p}{,} \PYG{n}{M}\PYG{o}{.}\PYG{n}{y}\PYG{o}{.}\PYG{n}{value}\PYG{p}{,} \PYG{n}{M}\PYG{o}{.}\PYG{n}{z}\PYG{p}{[}\PYG{l+m+mi}{1}\PYG{p}{]}\PYG{o}{.}\PYG{n}{value}\PYG{p}{,} \PYG{n}{M}\PYG{o}{.}\PYG{n}{z}\PYG{p}{[}\PYG{l+m+mi}{2}\PYG{p}{]}\PYG{o}{.}\PYG{n}{value}\PYG{p}{)}
\PYG{g+go}{2.0 8.0 5.5 0.0}
\end{sphinxVerbatim}

\paragraph{Trilevel Problems}
\label{\detokenize{representations/pyomo:trilevel-problems}}
Trilevel problems can be described with nested declarations of
{\hyperref[\detokenize{reference/pyomo:pao.pyomo.components.SubModel}]{\sphinxcrossref{\sphinxcode{\sphinxupquote{SubModel}}}}} components.  Consider the following trilevel continuous
linear problem described by Anadalingam \sphinxcite{anadalingam}:
\label{equation:sand:eq-anadalingam} \begin{equation*}
 \textbf{Model Anadalingam1988}
 \begin{array}{llll}
 \min_{x_1 \geq 0} & -7 x_1 - 3 x_2 + 4 x_3 \\
 \textrm{s.t.} & \min_{x_2 \geq 0} & -x_2 \\
               & \textrm{s.t.} & \min_{x_3 \in [0,0.5]} & -x_3 \\
               &               & \textrm{s.t.} & x_1 + x_2 + x_3 \leq 3\\
               &               &               & x_1 + x_2 - x_3 \leq 1\\
               &               &               & x_1 + x_2 + x_3 \geq 1\\
               &               &               & -x_1 + x_2 + x_3 \leq 1\\
 \end{array}
 \end{equation*}
This model can be expressed in Pyomo as follows:

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{ConcreteModel}\PYG{p}{(}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{x1} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Var}\PYG{p}{(}\PYG{n}{bounds}\PYG{o}{=}\PYG{p}{(}\PYG{l+m+mi}{0}\PYG{p}{,}\PYG{k+kc}{None}\PYG{p}{)}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{x2} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Var}\PYG{p}{(}\PYG{n}{bounds}\PYG{o}{=}\PYG{p}{(}\PYG{l+m+mi}{0}\PYG{p}{,}\PYG{k+kc}{None}\PYG{p}{)}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{x3} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Var}\PYG{p}{(}\PYG{n}{bounds}\PYG{o}{=}\PYG{p}{(}\PYG{l+m+mi}{0}\PYG{p}{,}\PYG{l+m+mf}{0.5}\PYG{p}{)}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L} \PYG{o}{=} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{pyomo}\PYG{o}{.}\PYG{n}{SubModel}\PYG{p}{(}\PYG{n}{fixed}\PYG{o}{=}\PYG{n}{M}\PYG{o}{.}\PYG{n}{x1}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{B} \PYG{o}{=} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{pyomo}\PYG{o}{.}\PYG{n}{SubModel}\PYG{p}{(}\PYG{n}{fixed}\PYG{o}{=}\PYG{n}{M}\PYG{o}{.}\PYG{n}{x2}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{o} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Objective}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{7}\PYG{o}{*}\PYG{n}{M}\PYG{o}{.}\PYG{n}{x1} \PYG{o}{\PYGZhy{}} \PYG{l+m+mi}{3}\PYG{o}{*}\PYG{n}{M}\PYG{o}{.}\PYG{n}{x2} \PYG{o}{+} \PYG{l+m+mi}{4}\PYG{o}{*}\PYG{n}{M}\PYG{o}{.}\PYG{n}{x3}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{o} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Objective}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=}\PYG{o}{\PYGZhy{}}\PYG{n}{M}\PYG{o}{.}\PYG{n}{x2}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{B}\PYG{o}{.}\PYG{n}{o} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Objective}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=}\PYG{o}{\PYGZhy{}}\PYG{n}{M}\PYG{o}{.}\PYG{n}{x3}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{B}\PYG{o}{.}\PYG{n}{c1} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Constraint}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=}   \PYG{n}{M}\PYG{o}{.}\PYG{n}{x1} \PYG{o}{+} \PYG{n}{M}\PYG{o}{.}\PYG{n}{x2} \PYG{o}{+} \PYG{n}{M}\PYG{o}{.}\PYG{n}{x3} \PYG{o}{\PYGZlt{}}\PYG{o}{=} \PYG{l+m+mi}{3}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{B}\PYG{o}{.}\PYG{n}{c2} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Constraint}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=}   \PYG{n}{M}\PYG{o}{.}\PYG{n}{x1} \PYG{o}{+} \PYG{n}{M}\PYG{o}{.}\PYG{n}{x2} \PYG{o}{\PYGZhy{}} \PYG{n}{M}\PYG{o}{.}\PYG{n}{x3} \PYG{o}{\PYGZlt{}}\PYG{o}{=} \PYG{l+m+mi}{1}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{B}\PYG{o}{.}\PYG{n}{c3} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Constraint}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=}   \PYG{n}{M}\PYG{o}{.}\PYG{n}{x1} \PYG{o}{+} \PYG{n}{M}\PYG{o}{.}\PYG{n}{x2} \PYG{o}{+} \PYG{n}{M}\PYG{o}{.}\PYG{n}{x3} \PYG{o}{\PYGZgt{}}\PYG{o}{=} \PYG{l+m+mi}{1}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{B}\PYG{o}{.}\PYG{n}{c4} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Constraint}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=} \PYG{o}{\PYGZhy{}} \PYG{n}{M}\PYG{o}{.}\PYG{n}{x1} \PYG{o}{+} \PYG{n}{M}\PYG{o}{.}\PYG{n}{x2} \PYG{o}{+} \PYG{n}{M}\PYG{o}{.}\PYG{n}{x3} \PYG{o}{\PYGZlt{}}\PYG{o}{=} \PYG{l+m+mi}{1}\PYG{p}{)}
\end{sphinxVerbatim}

\begin{sphinxadmonition}{note}{Note:}
PAO solvers cannot currently solve trilevel solvers like this,
but an issue has been submitted to add this functionality.
\end{sphinxadmonition}

\paragraph{Bilinear Problems}
\label{\detokenize{representations/pyomo:bilinear-problems}}
PAO models using Pyomo represent general quadratic problems with quadratic
terms in the objective and constraints at each level.  The special
case where bilinear terms arise with an upper\sphinxhyphen{}level binary variable
multiplied with a lower\sphinxhyphen{}level variable is common in many applications.
For this case, the PAO solvers for Pyomo models include an option to
linearize these bilinear terms.

The following models considers a variation of the \sphinxstylestrong{PAO1} model where
binary variables control the expression of lower\sphinxhyphen{}level constraints:
\label{equation:sand:eq-pao3} \begin{equation*}
 \textbf{Model PAO3}
 \begin{array}{ll}
 \min_{x\in[2,6],y,w_1,w_2} & x + 3 z + 5 w_1\\
 \textrm{s.t.} & x + y = 10\\
 & w_1 + w_2 \geq 1\\
 & w_1,w_2 \in \{0,1\}\\
 & \begin{array}{lll}
   \max_{z \geq 0} & z &\\
   \textrm{s.t.} & x+ w_1 z &\leq 8\\
   & x + 4 z &\geq 8\\
   & x + 2 w_2 z &\leq 13
   \end{array}
 \end{array}
 \end{equation*}
The \sphinxstylestrong{PAO3} model can be expressed in Pyomo as follows:

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{ConcreteModel}\PYG{p}{(}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{w} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Var}\PYG{p}{(}\PYG{p}{[}\PYG{l+m+mi}{1}\PYG{p}{,}\PYG{l+m+mi}{2}\PYG{p}{]}\PYG{p}{,} \PYG{n}{within}\PYG{o}{=}\PYG{n}{pe}\PYG{o}{.}\PYG{n}{Binary}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{x} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Var}\PYG{p}{(}\PYG{n}{bounds}\PYG{o}{=}\PYG{p}{(}\PYG{l+m+mi}{2}\PYG{p}{,}\PYG{l+m+mi}{6}\PYG{p}{)}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{y} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Var}\PYG{p}{(}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{z} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Var}\PYG{p}{(}\PYG{n}{bounds}\PYG{o}{=}\PYG{p}{(}\PYG{l+m+mi}{0}\PYG{p}{,}\PYG{k+kc}{None}\PYG{p}{)}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{o} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Objective}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=}\PYG{n}{M}\PYG{o}{.}\PYG{n}{x} \PYG{o}{+} \PYG{l+m+mi}{3}\PYG{o}{*}\PYG{n}{M}\PYG{o}{.}\PYG{n}{z}\PYG{o}{+}\PYG{l+m+mi}{5}\PYG{o}{*}\PYG{n}{M}\PYG{o}{.}\PYG{n}{w}\PYG{p}{[}\PYG{l+m+mi}{1}\PYG{p}{]}\PYG{p}{,} \PYG{n}{sense}\PYG{o}{=}\PYG{n}{pe}\PYG{o}{.}\PYG{n}{minimize}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{c1} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Constraint}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=} \PYG{n}{M}\PYG{o}{.}\PYG{n}{x} \PYG{o}{+} \PYG{n}{M}\PYG{o}{.}\PYG{n}{y} \PYG{o}{==} \PYG{l+m+mi}{10}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{c2} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Constraint}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=} \PYG{n}{M}\PYG{o}{.}\PYG{n}{w}\PYG{p}{[}\PYG{l+m+mi}{1}\PYG{p}{]} \PYG{o}{+} \PYG{n}{M}\PYG{o}{.}\PYG{n}{w}\PYG{p}{[}\PYG{l+m+mi}{2}\PYG{p}{]} \PYG{o}{\PYGZgt{}}\PYG{o}{=} \PYG{l+m+mi}{1}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L} \PYG{o}{=} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{pyomo}\PYG{o}{.}\PYG{n}{SubModel}\PYG{p}{(}\PYG{n}{fixed}\PYG{o}{=}\PYG{p}{[}\PYG{n}{M}\PYG{o}{.}\PYG{n}{x}\PYG{p}{,}\PYG{n}{M}\PYG{o}{.}\PYG{n}{y}\PYG{p}{,}\PYG{n}{M}\PYG{o}{.}\PYG{n}{w}\PYG{p}{]}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{o} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Objective}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=}\PYG{n}{M}\PYG{o}{.}\PYG{n}{z}\PYG{p}{,} \PYG{n}{sense}\PYG{o}{=}\PYG{n}{pe}\PYG{o}{.}\PYG{n}{maximize}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{c1} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Constraint}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=} \PYG{n}{M}\PYG{o}{.}\PYG{n}{x} \PYG{o}{+} \PYG{n}{M}\PYG{o}{.}\PYG{n}{w}\PYG{p}{[}\PYG{l+m+mi}{1}\PYG{p}{]}\PYG{o}{*}\PYG{n}{M}\PYG{o}{.}\PYG{n}{z} \PYG{o}{\PYGZlt{}}\PYG{o}{=} \PYG{l+m+mi}{8}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{c2} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Constraint}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=} \PYG{n}{M}\PYG{o}{.}\PYG{n}{x} \PYG{o}{+} \PYG{l+m+mi}{4}\PYG{o}{*}\PYG{n}{M}\PYG{o}{.}\PYG{n}{z} \PYG{o}{\PYGZgt{}}\PYG{o}{=} \PYG{l+m+mi}{8}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{c3} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Constraint}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=} \PYG{n}{M}\PYG{o}{.}\PYG{n}{x} \PYG{o}{+} \PYG{l+m+mi}{2}\PYG{o}{*}\PYG{n}{M}\PYG{o}{.}\PYG{n}{w}\PYG{p}{[}\PYG{l+m+mi}{2}\PYG{p}{]}\PYG{o}{*}\PYG{n}{M}\PYG{o}{.}\PYG{n}{z} \PYG{o}{\PYGZlt{}}\PYG{o}{=} \PYG{l+m+mi}{13}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{opt} \PYG{o}{=} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{Solver}\PYG{p}{(}\PYG{l+s+s2}{\PYGZdq{}}\PYG{l+s+s2}{pao.pyomo.FA}\PYG{l+s+s2}{\PYGZdq{}}\PYG{p}{,} \PYG{n}{linearize\PYGZus{}bigm}\PYG{o}{=}\PYG{l+m+mi}{100}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{results} \PYG{o}{=} \PYG{n}{opt}\PYG{o}{.}\PYG{n}{solve}\PYG{p}{(}\PYG{n}{M}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n+nb}{print}\PYG{p}{(}\PYG{n}{M}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{value}\PYG{p}{,} \PYG{n}{M}\PYG{o}{.}\PYG{n}{y}\PYG{o}{.}\PYG{n}{value}\PYG{p}{,} \PYG{n}{M}\PYG{o}{.}\PYG{n}{z}\PYG{o}{.}\PYG{n}{value}\PYG{p}{,} \PYG{n}{M}\PYG{o}{.}\PYG{n}{w}\PYG{p}{[}\PYG{l+m+mi}{1}\PYG{p}{]}\PYG{o}{.}\PYG{n}{value}\PYG{p}{,} \PYG{n}{M}\PYG{o}{.}\PYG{n}{w}\PYG{p}{[}\PYG{l+m+mi}{2}\PYG{p}{]}\PYG{o}{.}\PYG{n}{value}\PYG{p}{)}
\PYG{g+go}{6.0 4.0 3.5 0 1}
\end{sphinxVerbatim}

\section{Multilevel Representations}
\label{\detokenize{representations/mpr:multilevel-representations}}\label{\detokenize{representations/mpr::doc}}
PAO includes several \sphinxstyleemphasis{Multilevel Problem Representations} (MPRs)
that represent multilevel optimization problems with an explicit,
compact representation that simplifies the implementation of solvers for
bilevel, trilevel and other multilevel optimization problems.  These MPRs
express objective and constraints using vector and matrix data types.
However, they organize these data types to provide a semantically clear
organization of multilevel problems.  Additionally, the MPRs provide
checks to ensure the consistency of the data within and across levels.

The classes {\hyperref[\detokenize{reference/mpr:pao.mpr.repn.LinearMultilevelProblem}]{\sphinxcrossref{\sphinxcode{\sphinxupquote{LinearMultilevelProblem}}}}}
and {\hyperref[\detokenize{reference/mpr:pao.mpr.repn.QuadraticMultilevelProblem}]{\sphinxcrossref{\sphinxcode{\sphinxupquote{QuadraticMultilevelProblem}}}}} respectively
represent linear and quadratic multilevel problems.  Although
{\hyperref[\detokenize{reference/mpr:pao.mpr.repn.QuadraticMultilevelProblem}]{\sphinxcrossref{\sphinxcode{\sphinxupquote{QuadraticMultilevelProblem}}}}} is a generalization of the
representation in {\hyperref[\detokenize{reference/mpr:pao.mpr.repn.LinearMultilevelProblem}]{\sphinxcrossref{\sphinxcode{\sphinxupquote{LinearMultilevelProblem}}}}}, the use of tailored
representations for different classes of problems clarifies the semantic
context when using them.  For example, this allows for simple error checking
to confirm that a problem is linear.

Currently, all PAO solvers for MPRs support only linear problems, so the
following sections focus on {\hyperref[\detokenize{reference/mpr:pao.mpr.repn.LinearMultilevelProblem}]{\sphinxcrossref{\sphinxcode{\sphinxupquote{LinearMultilevelProblem}}}}}.  However,
we conclude with an example of model transformations that enable the
solution of quadratic problems using {\hyperref[\detokenize{reference/mpr:pao.mpr.repn.QuadraticMultilevelProblem}]{\sphinxcrossref{\sphinxcode{\sphinxupquote{QuadraticMultilevelProblem}}}}}.

\begin{sphinxadmonition}{note}{Note:}
We do not expect many users to directly employ a MPR data
representation for their applications.  Perhaps this would be
desirable if their problem was already represented with matrix and
vector data.  In general, the algebraic representation supported by
Pyomo will be more convenient for large, complex applications.

We expect this representation to be more useful for researchers
developing multilevel solvers, since the MPR representations provide
structure that simplifies the expression of necessary mathematical
operations for these problems.
\end{sphinxadmonition}

\subsection{Linear Bilevel Examples}
\label{\detokenize{representations/mpr:linear-bilevel-examples}}
We consider again the bilevel problem PAO1 \eqref{equation:sand:eq-pao1}.  This problem
has has linear upper\sphinxhyphen{} and lower\sphinxhyphen{}level problems with different objectives
in each level.

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M} \PYG{o}{=} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{mpr}\PYG{o}{.}\PYG{n}{LinearMultilevelProblem}\PYG{p}{(}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U} \PYG{o}{=} \PYG{n}{M}\PYG{o}{.}\PYG{n}{add\PYGZus{}upper}\PYG{p}{(}\PYG{n}{nxR}\PYG{o}{=}\PYG{l+m+mi}{2}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L} \PYG{o}{=} \PYG{n}{U}\PYG{o}{.}\PYG{n}{add\PYGZus{}lower}\PYG{p}{(}\PYG{n}{nxR}\PYG{o}{=}\PYG{l+m+mi}{1}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{lower\PYGZus{}bounds} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{2}\PYG{p}{,} \PYG{n}{np}\PYG{o}{.}\PYG{n}{NINF}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{upper\PYGZus{}bounds} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{6}\PYG{p}{,} \PYG{n}{np}\PYG{o}{.}\PYG{n}{PINF}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{lower\PYGZus{}bounds} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{upper\PYGZus{}bounds} \PYG{o}{=} \PYG{p}{[}\PYG{n}{np}\PYG{o}{.}\PYG{n}{PINF}\PYG{p}{]}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{c}\PYG{p}{[}\PYG{n}{U}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{1}\PYG{p}{,} \PYG{l+m+mi}{0}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{c}\PYG{p}{[}\PYG{n}{L}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{3}\PYG{p}{]}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L}\PYG{o}{.}\PYG{n}{c}\PYG{p}{[}\PYG{n}{L}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{1}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L}\PYG{o}{.}\PYG{n}{maximize} \PYG{o}{=} \PYG{k+kc}{True}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{equalities} \PYG{o}{=} \PYG{k+kc}{True}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{A}\PYG{p}{[}\PYG{n}{U}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{p}{[}\PYG{l+m+mi}{1}\PYG{p}{,} \PYG{l+m+mi}{1}\PYG{p}{]}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{b} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{10}\PYG{p}{]}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L}\PYG{o}{.}\PYG{n}{A}\PYG{p}{[}\PYG{n}{U}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{p}{[} \PYG{l+m+mi}{1}\PYG{p}{,} \PYG{l+m+mi}{0}\PYG{p}{]}\PYG{p}{,}
\PYG{g+gp}{... }         \PYG{p}{[}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{1}\PYG{p}{,} \PYG{l+m+mi}{0}\PYG{p}{]}\PYG{p}{,}
\PYG{g+gp}{... }         \PYG{p}{[} \PYG{l+m+mi}{1}\PYG{p}{,} \PYG{l+m+mi}{0}\PYG{p}{]}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L}\PYG{o}{.}\PYG{n}{A}\PYG{p}{[}\PYG{n}{L}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{p}{[} \PYG{l+m+mi}{1}\PYG{p}{]}\PYG{p}{,}
\PYG{g+gp}{... }         \PYG{p}{[}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{4}\PYG{p}{]}\PYG{p}{,}
\PYG{g+gp}{... }         \PYG{p}{[} \PYG{l+m+mi}{2}\PYG{p}{]}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L}\PYG{o}{.}\PYG{n}{b} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{8}\PYG{p}{,} \PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{8}\PYG{p}{,} \PYG{l+m+mi}{13}\PYG{p}{]}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M}\PYG{o}{.}\PYG{n}{check}\PYG{p}{(}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{opt} \PYG{o}{=} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{Solver}\PYG{p}{(}\PYG{l+s+s2}{\PYGZdq{}}\PYG{l+s+s2}{pao.mpr.FA}\PYG{l+s+s2}{\PYGZdq{}}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{results} \PYG{o}{=} \PYG{n}{opt}\PYG{o}{.}\PYG{n}{solve}\PYG{p}{(}\PYG{n}{M}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n+nb}{print}\PYG{p}{(}\PYG{n}{M}\PYG{o}{.}\PYG{n}{U}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{values}\PYG{p}{)}
\PYG{g+go}{[6.0, 4.0]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n+nb}{print}\PYG{p}{(}\PYG{n}{M}\PYG{o}{.}\PYG{n}{U}\PYG{o}{.}\PYG{n}{LL}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{values}\PYG{p}{)}
\PYG{g+go}{[2.0]}
\end{sphinxVerbatim}

The example illustrates both the flexibility of the MPR representions
in PAO but also the structure they enforce on the multilevel problem
representation.  The upper\sphinxhyphen{}level problem is created by calling the
\sphinxcode{\sphinxupquote{add\_upper()}} method, which takes arguments that specify the
variables at that level:
\begin{itemize}
\item {} 
\sphinxcode{\sphinxupquote{nxR}} \sphinxhyphen{} The number of real variables (Default: 0)

\item {} 
\sphinxcode{\sphinxupquote{nxZ}} \sphinxhyphen{} The number of general integer variables (Default: 0)

\item {} 
\sphinxcode{\sphinxupquote{nxB}} \sphinxhyphen{} The number of binary variables (Default: 0)

\end{itemize}

In each level, the variables are represented as a vector of values,
ordered in this manner.

Similarly, the \sphinxcode{\sphinxupquote{add\_lower()}} method is used to generate a lower\sphinxhyphen{}level
problem from a given level.  Note that this allows for the specification
of arbitrary nesting of levels, since a lower\sphinxhyphen{}level can be defined
relative to any other level in the model.  Additionally, multiple
lower\sphinxhyphen{}levels can be specified for relative to a single level (see below).

The \sphinxcode{\sphinxupquote{add\_upper()}} and \sphinxcode{\sphinxupquote{add\_lower()}} methods return the
corresponding level object, which is used to specify data in the model
later.

For a given level object, \sphinxcode{\sphinxupquote{Z}}, the data \sphinxcode{\sphinxupquote{Z.x}} contains
information about the decision variables.  In particular, the values
\sphinxcode{\sphinxupquote{Z.x.lower\_bounds}} and \sphinxcode{\sphinxupquote{Z.x.upper\_bounds}} can be set with arrays
of numeric values to specify lower\sphinxhyphen{} and upper\sphinxhyphen{}bounds on the decision
variables.  Note that missing lower\sphinxhyphen{} and upper\sphinxhyphen{}bounds are specified with
\sphinxcode{\sphinxupquote{numpy.NINF}} and \sphinxcode{\sphinxupquote{numpy.PINF}} respectively.

The \sphinxcode{\sphinxupquote{Z.c}} data specifies coefficients of the objective function for
this level.  This data is indexed by a level object \sphinxcode{\sphinxupquote{B}} to indicate
the data associated with the variables in \sphinxcode{\sphinxupquote{B}}.  In the example above:
\begin{itemize}
\item {} 
\sphinxcode{\sphinxupquote{U.c{[}U{]}}} is the array of coefficients of the upper\sphinxhyphen{}level objective for the variables in the upper\sphinxhyphen{}level,

\item {} 
\sphinxcode{\sphinxupquote{U.c{[}L{]}}} is the array of coefficients of the upper\sphinxhyphen{}level objective for the variables in the lower\sphinxhyphen{}level, and

\item {} 
\sphinxcode{\sphinxupquote{L.c{[}L{]}}} is the array of coefficients of the lower\sphinxhyphen{}level objective for the variables in the lower\sphinxhyphen{}level.

\end{itemize}

Since \sphinxcode{\sphinxupquote{L.c{[}U{]}}} is not specified, it has a value \sphinxcode{\sphinxupquote{None}} that
indicates that no upper\sphinxhyphen{}level variables have non\sphinxhyphen{}zero coefficients in the
lower\sphinxhyphen{}level objective.  The \sphinxcode{\sphinxupquote{Z.A}} data specifies the matrix coefficients
for the constraints using a similar indexing notation and semantics.

The values \sphinxcode{\sphinxupquote{Z.minimize}} and \sphinxcode{\sphinxupquote{Z.maximize}} can be set to \sphinxcode{\sphinxupquote{True}}
to indicate whether the objective in \sphinxcode{\sphinxupquote{Z}} minimizes or maximizes.
(The default is minimize.)  Similarly the value \sphinxcode{\sphinxupquote{Z.inequalities}}
and \sphinxcode{\sphinxupquote{Z.equalities}} can be set to \sphinxcode{\sphinxupquote{True}} to indicate whether the
constraints in \sphinxcode{\sphinxupquote{Z}} are inequalities or equalities.  (The default
is inequalities.)  Finally, the value \sphinxcode{\sphinxupquote{Z.b}} defines the array of
constraint right\sphinxhyphen{}hand\sphinxhyphen{}side values.

The :meth:\sphinxcode{\sphinxupquote{check}} method provides a convenient sanity check that the
data is defined consistently within each level and between levels.

Note that PAO supports a consistent interface for creating a solver
interface and for applying solvers.  In fact, the user should be
aware that Pyomo and MPR solvers are named in a consistent fashion.
For example, the Pyomo solver \sphinxstylestrong{pao.pyomo.FA} calls the MPR solver
\sphinxstylestrong{pao.mpr.FA} after automatically converting the Pyomo representation
to a :class:\sphinxcode{\sphinxupquote{LinearMultilevelProblem}} representation.  This example
illustrates that values \sphinxcode{\sphinxupquote{Z.x.values}} contains the values of each level
\sphinxcode{\sphinxupquote{Z}} after optimization.

\subsection{Multilevel Examples}
\label{\detokenize{representations/mpr:multilevel-examples}}
Multilevel problems can be easily expressed using the same MPR data
representation.

\paragraph{Multiple Lower Levels}
\label{\detokenize{representations/mpr:multiple-lower-levels}}
We consider again the bilevel problem PAO2 \eqref{equation:sand:eq-pao2}, which has has
multiple lower\sphinxhyphen{}level problems.  The \sphinxstylestrong{PAO2} model can be expressed as
a linear multilevel problem as follows:

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M} \PYG{o}{=} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{mpr}\PYG{o}{.}\PYG{n}{LinearMultilevelProblem}\PYG{p}{(}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U} \PYG{o}{=} \PYG{n}{M}\PYG{o}{.}\PYG{n}{add\PYGZus{}upper}\PYG{p}{(}\PYG{n}{nxR}\PYG{o}{=}\PYG{l+m+mi}{2}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L1} \PYG{o}{=} \PYG{n}{U}\PYG{o}{.}\PYG{n}{add\PYGZus{}lower}\PYG{p}{(}\PYG{n}{nxR}\PYG{o}{=}\PYG{l+m+mi}{1}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L2} \PYG{o}{=} \PYG{n}{U}\PYG{o}{.}\PYG{n}{add\PYGZus{}lower}\PYG{p}{(}\PYG{n}{nxR}\PYG{o}{=}\PYG{l+m+mi}{1}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{lower\PYGZus{}bounds} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{2}\PYG{p}{,} \PYG{n}{np}\PYG{o}{.}\PYG{n}{NINF}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{upper\PYGZus{}bounds} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{6}\PYG{p}{,} \PYG{n}{np}\PYG{o}{.}\PYG{n}{PINF}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{c}\PYG{p}{[}\PYG{n}{U}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{1}\PYG{p}{,} \PYG{l+m+mi}{0}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{c}\PYG{p}{[}\PYG{n}{L1}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{3}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{c}\PYG{p}{[}\PYG{n}{L2}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{3}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{equalities} \PYG{o}{=} \PYG{k+kc}{True}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{A}\PYG{p}{[}\PYG{n}{U}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{p}{[}\PYG{l+m+mi}{1}\PYG{p}{,} \PYG{l+m+mi}{1}\PYG{p}{]}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{b} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{10}\PYG{p}{]}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L1}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{lower\PYGZus{}bounds} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L1}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{upper\PYGZus{}bounds} \PYG{o}{=} \PYG{p}{[}\PYG{n}{np}\PYG{o}{.}\PYG{n}{PINF}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L1}\PYG{o}{.}\PYG{n}{c}\PYG{p}{[}\PYG{n}{L1}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{1}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L1}\PYG{o}{.}\PYG{n}{maximize} \PYG{o}{=} \PYG{k+kc}{True}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L1}\PYG{o}{.}\PYG{n}{A}\PYG{p}{[}\PYG{n}{U}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{p}{[} \PYG{l+m+mi}{1}\PYG{p}{,} \PYG{l+m+mi}{0}\PYG{p}{]}\PYG{p}{,}
\PYG{g+gp}{... }          \PYG{p}{[}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{1}\PYG{p}{,} \PYG{l+m+mi}{0}\PYG{p}{]}\PYG{p}{,}
\PYG{g+gp}{... }          \PYG{p}{[} \PYG{l+m+mi}{1}\PYG{p}{,} \PYG{l+m+mi}{0}\PYG{p}{]}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L1}\PYG{o}{.}\PYG{n}{A}\PYG{p}{[}\PYG{n}{L1}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{p}{[} \PYG{l+m+mi}{1}\PYG{p}{]}\PYG{p}{,}
\PYG{g+gp}{... }           \PYG{p}{[}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{4}\PYG{p}{]}\PYG{p}{,}
\PYG{g+gp}{... }           \PYG{p}{[} \PYG{l+m+mi}{2}\PYG{p}{]}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L1}\PYG{o}{.}\PYG{n}{b} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{8}\PYG{p}{,} \PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{8}\PYG{p}{,} \PYG{l+m+mi}{13}\PYG{p}{]}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L2}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{lower\PYGZus{}bounds} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L2}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{upper\PYGZus{}bounds} \PYG{o}{=} \PYG{p}{[}\PYG{n}{np}\PYG{o}{.}\PYG{n}{PINF}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L2}\PYG{o}{.}\PYG{n}{c}\PYG{p}{[}\PYG{n}{L2}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{1}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L2}\PYG{o}{.}\PYG{n}{maximize} \PYG{o}{=} \PYG{k+kc}{True}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L2}\PYG{o}{.}\PYG{n}{A}\PYG{p}{[}\PYG{n}{U}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{,}  \PYG{l+m+mi}{1}\PYG{p}{]}\PYG{p}{,}
\PYG{g+gp}{... }          \PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{,} \PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{1}\PYG{p}{]}\PYG{p}{,}
\PYG{g+gp}{... }          \PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{,}  \PYG{l+m+mi}{1}\PYG{p}{]}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L2}\PYG{o}{.}\PYG{n}{A}\PYG{p}{[}\PYG{n}{L2}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{p}{[} \PYG{l+m+mi}{1}\PYG{p}{]}\PYG{p}{,}
\PYG{g+gp}{... }           \PYG{p}{[}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{4}\PYG{p}{]}\PYG{p}{,}
\PYG{g+gp}{... }           \PYG{p}{[} \PYG{l+m+mi}{2}\PYG{p}{]}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L2}\PYG{o}{.}\PYG{n}{b} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{8}\PYG{p}{,} \PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{8}\PYG{p}{,} \PYG{l+m+mi}{13}\PYG{p}{]}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{opt} \PYG{o}{=} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{Solver}\PYG{p}{(}\PYG{l+s+s2}{\PYGZdq{}}\PYG{l+s+s2}{pao.mpr.FA}\PYG{l+s+s2}{\PYGZdq{}}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{results} \PYG{o}{=} \PYG{n}{opt}\PYG{o}{.}\PYG{n}{solve}\PYG{p}{(}\PYG{n}{M}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n+nb}{print}\PYG{p}{(}\PYG{n}{U}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{values}\PYG{p}{)}
\PYG{g+go}{[2.0, 8.0]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n+nb}{print}\PYG{p}{(}\PYG{n}{L1}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{values}\PYG{p}{)}
\PYG{g+go}{[5.5]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n+nb}{print}\PYG{p}{(}\PYG{n}{L2}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{values}\PYG{p}{)}
\PYG{g+go}{[0.0]}
\end{sphinxVerbatim}

The declarataion of the two lower level problems is naturally contained
within the data of the \sphinxcode{\sphinxupquote{L1}} and \sphinxcode{\sphinxupquote{L2}} objects.  Further, the
cross\sphinxhyphen{}level interactions are intuitively represented using the index
notation for the objective and constraint data objects.

Note that this more explicit representation clarifies some ambiguity in
the expression of lower\sphinxhyphen{}levels in the Pyomo representation.  The Pyomo
representation of PAO2 only specifies the fixed variables that are
\sphinxstylestrong{used} in each of the two lower\sphinxhyphen{}level problems.  PAO analyzes
the use of decision variables in Pyomo models, and treats \sphinxstyleemphasis{unused}
variables as fixed.  Thus, the Pyomo and MPR representations generate
a consistent interpretation of the variable specifications.  However,
the MPR representation is more explicit in this regard.

\paragraph{Trilevel Problems}
\label{\detokenize{representations/mpr:trilevel-problems}}
We consider again the trilevel problem described by Anadalingam
\eqref{equation:sand:eq-anadalingam}, which can be expressed as a trilevel linear problem
as follows:

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M} \PYG{o}{=} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{mpr}\PYG{o}{.}\PYG{n}{LinearMultilevelProblem}\PYG{p}{(}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U} \PYG{o}{=} \PYG{n}{M}\PYG{o}{.}\PYG{n}{add\PYGZus{}upper}\PYG{p}{(}\PYG{n}{nxR}\PYG{o}{=}\PYG{l+m+mi}{1}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{lower\PYGZus{}bounds} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{]}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L} \PYG{o}{=} \PYG{n}{U}\PYG{o}{.}\PYG{n}{add\PYGZus{}lower}\PYG{p}{(}\PYG{n}{nxR}\PYG{o}{=}\PYG{l+m+mi}{1}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{lower\PYGZus{}bounds} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{]}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{B} \PYG{o}{=} \PYG{n}{L}\PYG{o}{.}\PYG{n}{add\PYGZus{}lower}\PYG{p}{(}\PYG{n}{nxR}\PYG{o}{=}\PYG{l+m+mi}{1}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{B}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{lower\PYGZus{}bounds} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{B}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{upper\PYGZus{}bounds} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mf}{0.5}\PYG{p}{]}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{minimize} \PYG{o}{=} \PYG{k+kc}{True}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{c}\PYG{p}{[}\PYG{n}{U}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{7}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{c}\PYG{p}{[}\PYG{n}{L}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{3}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{c}\PYG{p}{[}\PYG{n}{B}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{4}\PYG{p}{]}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L}\PYG{o}{.}\PYG{n}{minimize} \PYG{o}{=} \PYG{k+kc}{True}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L}\PYG{o}{.}\PYG{n}{c}\PYG{p}{[}\PYG{n}{L}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{1}\PYG{p}{]}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{B}\PYG{o}{.}\PYG{n}{minimize} \PYG{o}{=} \PYG{k+kc}{True}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{B}\PYG{o}{.}\PYG{n}{c}\PYG{p}{[}\PYG{n}{B}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{1}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{B}\PYG{o}{.}\PYG{n}{inequalities} \PYG{o}{=} \PYG{k+kc}{True}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{B}\PYG{o}{.}\PYG{n}{A}\PYG{p}{[}\PYG{n}{U}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{p}{[}\PYG{l+m+mi}{1}\PYG{p}{]}\PYG{p}{,} \PYG{p}{[} \PYG{l+m+mi}{1}\PYG{p}{]}\PYG{p}{,} \PYG{p}{[}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{1}\PYG{p}{]}\PYG{p}{,} \PYG{p}{[}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{1}\PYG{p}{]}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{B}\PYG{o}{.}\PYG{n}{A}\PYG{p}{[}\PYG{n}{L}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{p}{[}\PYG{l+m+mi}{1}\PYG{p}{]}\PYG{p}{,} \PYG{p}{[} \PYG{l+m+mi}{1}\PYG{p}{]}\PYG{p}{,} \PYG{p}{[}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{1}\PYG{p}{]}\PYG{p}{,} \PYG{p}{[} \PYG{l+m+mi}{1}\PYG{p}{]}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{B}\PYG{o}{.}\PYG{n}{A}\PYG{p}{[}\PYG{n}{B}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{p}{[}\PYG{l+m+mi}{1}\PYG{p}{]}\PYG{p}{,} \PYG{p}{[}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{1}\PYG{p}{]}\PYG{p}{,} \PYG{p}{[}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{1}\PYG{p}{]}\PYG{p}{,} \PYG{p}{[} \PYG{l+m+mi}{1}\PYG{p}{]}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{B}\PYG{o}{.}\PYG{n}{b} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{3}\PYG{p}{,}\PYG{l+m+mi}{1}\PYG{p}{,}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{1}\PYG{p}{,}\PYG{l+m+mi}{1}\PYG{p}{]}
\end{sphinxVerbatim}

\paragraph{Bilinear Problems}
\label{\detokenize{representations/mpr:bilinear-problems}}
The {\hyperref[\detokenize{reference/mpr:pao.mpr.repn.QuadraticMultilevelProblem}]{\sphinxcrossref{\sphinxcode{\sphinxupquote{QuadraticMultilevelProblem}}}}} class can represent general
quadratic problems with quadratic terms in the objective and constraints
at each level.  The special case where bilinear terms arise with an
upper\sphinxhyphen{}level binary variable multiplied with a lower\sphinxhyphen{}level variable is
common in many applications.  For this case, PAO provides a function to
linearize these bilinear terms.

We consider again the bilevel problem PAO3 \eqref{equation:sand:eq-pao3}, which is
represented and solved as follows:

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{M} \PYG{o}{=} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{mpr}\PYG{o}{.}\PYG{n}{QuadraticMultilevelProblem}\PYG{p}{(}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U} \PYG{o}{=} \PYG{n}{M}\PYG{o}{.}\PYG{n}{add\PYGZus{}upper}\PYG{p}{(}\PYG{n}{nxR}\PYG{o}{=}\PYG{l+m+mi}{2}\PYG{p}{,} \PYG{n}{nxB}\PYG{o}{=}\PYG{l+m+mi}{2}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L} \PYG{o}{=} \PYG{n}{U}\PYG{o}{.}\PYG{n}{add\PYGZus{}lower}\PYG{p}{(}\PYG{n}{nxR}\PYG{o}{=}\PYG{l+m+mi}{1}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{lower\PYGZus{}bounds} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{2}\PYG{p}{,} \PYG{n}{np}\PYG{o}{.}\PYG{n}{NINF}\PYG{p}{,} \PYG{l+m+mi}{0}\PYG{p}{,} \PYG{l+m+mi}{0}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{upper\PYGZus{}bounds} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{6}\PYG{p}{,} \PYG{n}{np}\PYG{o}{.}\PYG{n}{PINF}\PYG{p}{,} \PYG{l+m+mi}{1}\PYG{p}{,} \PYG{l+m+mi}{1}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{lower\PYGZus{}bounds} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{0}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{upper\PYGZus{}bounds} \PYG{o}{=} \PYG{p}{[}\PYG{n}{np}\PYG{o}{.}\PYG{n}{PINF}\PYG{p}{]}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{c}\PYG{p}{[}\PYG{n}{U}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{1}\PYG{p}{,} \PYG{l+m+mi}{0}\PYG{p}{,} \PYG{l+m+mi}{5}\PYG{p}{,} \PYG{l+m+mi}{0}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{c}\PYG{p}{[}\PYG{n}{L}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{3}\PYG{p}{]}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L}\PYG{o}{.}\PYG{n}{c}\PYG{p}{[}\PYG{n}{L}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{1}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L}\PYG{o}{.}\PYG{n}{maximize} \PYG{o}{=} \PYG{k+kc}{True}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{A}\PYG{p}{[}\PYG{n}{U}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{p}{[} \PYG{l+m+mi}{1}\PYG{p}{,}  \PYG{l+m+mi}{1}\PYG{p}{,}  \PYG{l+m+mi}{0}\PYG{p}{,}  \PYG{l+m+mi}{0}\PYG{p}{]}\PYG{p}{,}
\PYG{g+gp}{... }         \PYG{p}{[}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{1}\PYG{p}{,} \PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{1}\PYG{p}{,}  \PYG{l+m+mi}{0}\PYG{p}{,}  \PYG{l+m+mi}{0}\PYG{p}{]}\PYG{p}{,}
\PYG{g+gp}{... }         \PYG{p}{[} \PYG{l+m+mi}{0}\PYG{p}{,}  \PYG{l+m+mi}{0}\PYG{p}{,} \PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{1}\PYG{p}{,} \PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{1}\PYG{p}{]}
\PYG{g+gp}{... }         \PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{U}\PYG{o}{.}\PYG{n}{b} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{10}\PYG{p}{,} \PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{10}\PYG{p}{,} \PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{1}\PYG{p}{]}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L}\PYG{o}{.}\PYG{n}{A}\PYG{p}{[}\PYG{n}{U}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{p}{[} \PYG{l+m+mi}{1}\PYG{p}{,} \PYG{l+m+mi}{0}\PYG{p}{,} \PYG{l+m+mi}{0}\PYG{p}{,} \PYG{l+m+mi}{0}\PYG{p}{]}\PYG{p}{,}
\PYG{g+gp}{... }         \PYG{p}{[}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{1}\PYG{p}{,} \PYG{l+m+mi}{0}\PYG{p}{,} \PYG{l+m+mi}{0}\PYG{p}{,} \PYG{l+m+mi}{0}\PYG{p}{]}\PYG{p}{,}
\PYG{g+gp}{... }         \PYG{p}{[} \PYG{l+m+mi}{1}\PYG{p}{,} \PYG{l+m+mi}{0}\PYG{p}{,} \PYG{l+m+mi}{0}\PYG{p}{,} \PYG{l+m+mi}{0}\PYG{p}{]}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L}\PYG{o}{.}\PYG{n}{A}\PYG{p}{[}\PYG{n}{L}\PYG{p}{]} \PYG{o}{=} \PYG{p}{[}\PYG{p}{[} \PYG{l+m+mi}{0}\PYG{p}{]}\PYG{p}{,}
\PYG{g+gp}{... }         \PYG{p}{[}\PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{4}\PYG{p}{]}\PYG{p}{,}
\PYG{g+gp}{... }         \PYG{p}{[} \PYG{l+m+mi}{0}\PYG{p}{]}\PYG{p}{]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L}\PYG{o}{.}\PYG{n}{Q}\PYG{p}{[}\PYG{n}{U}\PYG{p}{,}\PYG{n}{L}\PYG{p}{]} \PYG{o}{=} \PYG{p}{(}\PYG{l+m+mi}{3}\PYG{p}{,}\PYG{l+m+mi}{4}\PYG{p}{,}\PYG{l+m+mi}{1}\PYG{p}{)}\PYG{p}{,} \PYG{p}{\PYGZob{}}\PYG{p}{(}\PYG{l+m+mi}{0}\PYG{p}{,}\PYG{l+m+mi}{2}\PYG{p}{,}\PYG{l+m+mi}{0}\PYG{p}{)}\PYG{p}{:}\PYG{l+m+mi}{1}\PYG{p}{,} \PYG{p}{(}\PYG{l+m+mi}{2}\PYG{p}{,}\PYG{l+m+mi}{3}\PYG{p}{,}\PYG{l+m+mi}{0}\PYG{p}{)}\PYG{p}{:}\PYG{l+m+mi}{2}\PYG{p}{\PYGZcb{}}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{L}\PYG{o}{.}\PYG{n}{b} \PYG{o}{=} \PYG{p}{[}\PYG{l+m+mi}{8}\PYG{p}{,} \PYG{o}{\PYGZhy{}}\PYG{l+m+mi}{8}\PYG{p}{,} \PYG{l+m+mi}{13}\PYG{p}{]}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{lmr}\PYG{p}{,} \PYG{n}{soln} \PYG{o}{=} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{mpr}\PYG{o}{.}\PYG{n}{linearize\PYGZus{}bilinear\PYGZus{}terms}\PYG{p}{(}\PYG{n}{M}\PYG{p}{,} \PYG{l+m+mi}{100}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{opt} \PYG{o}{=} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{Solver}\PYG{p}{(}\PYG{l+s+s2}{\PYGZdq{}}\PYG{l+s+s2}{pao.mpr.FA}\PYG{l+s+s2}{\PYGZdq{}}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{results} \PYG{o}{=} \PYG{n}{opt}\PYG{o}{.}\PYG{n}{solve}\PYG{p}{(}\PYG{n}{lmr}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{soln}\PYG{o}{.}\PYG{n}{copy}\PYG{p}{(}\PYG{n}{From}\PYG{o}{=}\PYG{n}{lmr}\PYG{p}{,} \PYG{n}{To}\PYG{o}{=}\PYG{n}{M}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n+nb}{print}\PYG{p}{(}\PYG{n}{U}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{values}\PYG{p}{)}
\PYG{g+go}{[6.0, 4.0, 0, 1]}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n+nb}{print}\PYG{p}{(}\PYG{n}{L}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{values}\PYG{p}{)}
\PYG{g+go}{[3.5]}
\end{sphinxVerbatim}

The data \sphinxcode{\sphinxupquote{L.Q{[}U,L{]}}} specifies the bilinear terms multiplying
variables from level \sphinxcode{\sphinxupquote{U}} with variables from level \sphinxcode{\sphinxupquote{L}}, which
are included in the constraints in level \sphinxcode{\sphinxupquote{L}}.  Note that \sphinxcode{\sphinxupquote{Q}} is a
tensor, which is indexed over the constraints, upper\sphinxhyphen{}level variables
and lower\sphinxhyphen{}level variables.  A similar syntax is used to define bilinear
terms in objectives, \sphinxcode{\sphinxupquote{P}}, though that is represented as a sparse matrix.
Quadratic terms can be specified simply by using the same levels to index
\sphinxcode{\sphinxupquote{Q}} or \sphinxcode{\sphinxupquote{P}}.

Model transformations like :func:\sphinxcode{\sphinxupquote{.linearize\_bilinear\_terms}} are
described in further detail in the next section.  Note that this function
returns both the transformed model as well as a helper class that maps
solutions back to the original model.  This logic facilitates the
automation of model transformations within PAO.

\section{Solvers}
\label{\detokenize{solvers:solvers}}\label{\detokenize{solvers::doc}}
After formulating a multilevel problem, PAO users will generally need to
(1) transform the model to a standard form, and (2) apply an optimizer
to solve the problem.  The examples in the previous sections illustrate
that step (1) is often optional;  PAO automates the applications of
several model transformations, particularly for problems formulated with
Pyomo.  The following section summarizes the solvers available in PAO,
and describes how PAO manages solvers Section \DUrole{xref,std,std-ref}{transformations}
describes model transformations in PAO.

\subsection{Summary of PAO Solvers}
\label{\detokenize{solvers:summary-of-pao-solvers}}
The following summarizes the current solvers available in PAO:
\begin{itemize}
\item {} 
pao.mpr.FA, pao.pyomo.FA
\begin{quote}

PAO solver for Multilevel Problem Representations that define linear
bilevel problems.  Solver uses big\sphinxhyphen{}M relaxations discussed by Fortuny\sphinxhyphen{}
Amat and McCarl (1981).
\end{quote}

\item {} 
pao.mpr.MIBS, pao.pyomo.MIBS
\begin{quote}

PAO solver for Multilevel Problem Representations using the COIN\sphinxhyphen{}OR
MibS solver by Tahernejad, Ralphs, and DeNegre (2020).
\end{quote}

\item {} 
pao.mpr.PCCG, pao.pyomo.PCCG
\begin{quote}

PAO solver for Multilevel Problem Representations that define linear
bilevel problems. Solver uses projected column constraint generation
algorithm described by Yue et al. (2017).
\end{quote}

\item {} 
pao.mpr.REG, pao.pyomo.REG
\begin{quote}

PAO solver for Multilevel Problem Representations that define linear
bilevel problems.  Solver uses regularization discussed by Scheel and
Scholtes (2000) and Ralph and Wright (2004).
\end{quote}

\end{itemize}

The following table summarize key features of the problems these solvers
can be applied to:

\begin{savenotes}\sphinxattablestart
\centering
\begin{tabulary}{\linewidth}[t]{|T|T|T|T|T|T|}
\hline
\sphinxstartmulticolumn{2}%
\begin{varwidth}[t]{\sphinxcolwidth{2}{6}}
\par
\vskip-\baselineskip\vbox{\hbox{\strut}}\end{varwidth}%
\sphinxstopmulticolumn
&\sphinxstartmulticolumn{4}%
\begin{varwidth}[t]{\sphinxcolwidth{4}{6}}
\sphinxstylestrong{Solver}
\par
\vskip-\baselineskip\vbox{\hbox{\strut}}\end{varwidth}%
\sphinxstopmulticolumn
\\
\hline\sphinxstartmulticolumn{2}%
\begin{varwidth}[t]{\sphinxcolwidth{2}{6}}
\sphinxstylestrong{Problem Feature}
\par
\vskip-\baselineskip\vbox{\hbox{\strut}}\end{varwidth}%
\sphinxstopmulticolumn
&
\sphinxstyleemphasis{FA}
&
\sphinxstyleemphasis{REG}
&
\sphinxstyleemphasis{PCCG}
&
\sphinxstyleemphasis{MibS}
\\
\hline\sphinxmultirow{3}{8}{%
\begin{varwidth}[t]{\sphinxcolwidth{1}{6}}
Equation
Structure
\par
\vskip-\baselineskip\vbox{\hbox{\strut}}\end{varwidth}%
}%
&
Linear
&
Y
&
Y
&
Y
&
Y
\\
\cline{2-6}\sphinxtablestrut{8}&
Bilinear
&
Y
&
Y
&
Y
&
Y
\\
\cline{2-6}\sphinxtablestrut{8}&
Nonlinear
&&&&\\
\hline\sphinxmultirow{2}{24}{%
\begin{varwidth}[t]{\sphinxcolwidth{1}{6}}
Upper\sphinxhyphen{}Level
Variables
\par
\vskip-\baselineskip\vbox{\hbox{\strut}}\end{varwidth}%
}%
&
Integer
&
Y
&&
Y
&
Y
\\
\cline{2-6}\sphinxtablestrut{24}&
Real
&
Y
&
Y
&
Y
&
Y
\\
\hline\sphinxmultirow{2}{35}{%
\begin{varwidth}[t]{\sphinxcolwidth{1}{6}}
Lower\sphinxhyphen{}Level
Variables
\par
\vskip-\baselineskip\vbox{\hbox{\strut}}\end{varwidth}%
}%
&
Integer
&&&
Y
&
Y
\\
\cline{2-6}\sphinxtablestrut{35}&
Real
&
Y
&
Y
&
Y
&
Y
\\
\hline\sphinxmultirow{3}{46}{%
\begin{varwidth}[t]{\sphinxcolwidth{1}{6}}
Multilevel
Representation
\par
\vskip-\baselineskip\vbox{\hbox{\strut}}\end{varwidth}%
}%
&
Bilevel
&
Y
&
Y
&
Y
&
Y
\\
\cline{2-6}\sphinxtablestrut{46}&
Trilevel
&&&&\\
\cline{2-6}\sphinxtablestrut{46}&
k\sphinxhyphen{}Bilevel
&
Y
&
Y
&&\\
\hline
\end{tabulary}
\par
\sphinxattableend\end{savenotes}

\begin{sphinxadmonition}{note}{Note:}
The iterface to MibS is a prototype that has not been well\sphinxhyphen{}tested.
This interface will be documented and finalized in an upcomgin next
release of PAO.
\end{sphinxadmonition}

\subsection{The Solver Interface}
\label{\detokenize{solvers:the-solver-interface}}
The \sphinxcode{\sphinxupquote{Solver}} object provides a single interface for setting up
an interface to optimizers in PAO.  This includes \sphinxstyleemphasis{both} PAO solvers for
multilevel optimization problems, but also interfaces to conventional
numerical solvers that are used by PAO solvers.  We illustrate this
distinction with the following example, which optimizes the PAO1
\eqref{equation:sand:eq-pao1} example:

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{c+c1}{\PYGZsh{} Create an interface to the PAO FA solver}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{opt} \PYG{o}{=} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{Solver}\PYG{p}{(}\PYG{l+s+s2}{\PYGZdq{}}\PYG{l+s+s2}{pao.pyomo.FA}\PYG{l+s+s2}{\PYGZdq{}}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{c+c1}{\PYGZsh{} Optimize the model}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{c+c1}{\PYGZsh{} By default, FA uses the glpk MIP solver}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{results} \PYG{o}{=} \PYG{n}{opt}\PYG{o}{.}\PYG{n}{solve}\PYG{p}{(}\PYG{n}{M}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n+nb}{print}\PYG{p}{(}\PYG{n}{M}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{value}\PYG{p}{,} \PYG{n}{M}\PYG{o}{.}\PYG{n}{y}\PYG{o}{.}\PYG{n}{value}\PYG{p}{,} \PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{z}\PYG{o}{.}\PYG{n}{value}\PYG{p}{)}
\PYG{g+go}{6.0 4.0 2.0}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{c+c1}{\PYGZsh{} Create an interface to the PAO FA solver, using cbc}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{opt} \PYG{o}{=} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{Solver}\PYG{p}{(}\PYG{l+s+s2}{\PYGZdq{}}\PYG{l+s+s2}{pao.pyomo.FA}\PYG{l+s+s2}{\PYGZdq{}}\PYG{p}{,} \PYG{n}{mip\PYGZus{}solver}\PYG{o}{=}\PYG{l+s+s2}{\PYGZdq{}}\PYG{l+s+s2}{cbc}\PYG{l+s+s2}{\PYGZdq{}}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{c+c1}{\PYGZsh{} Optimize the model using cbc}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{results} \PYG{o}{=} \PYG{n}{opt}\PYG{o}{.}\PYG{n}{solve}\PYG{p}{(}\PYG{n}{M}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n+nb}{print}\PYG{p}{(}\PYG{n}{M}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{value}\PYG{p}{,} \PYG{n}{M}\PYG{o}{.}\PYG{n}{y}\PYG{o}{.}\PYG{n}{value}\PYG{p}{,} \PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{z}\PYG{o}{.}\PYG{n}{value}\PYG{p}{)}
\PYG{g+go}{6.0 4.0 2.0}
\end{sphinxVerbatim}

The \sphinxcode{\sphinxupquote{Solver}} object is initialized using the solver name followed
by solver\sphinxhyphen{}specific options.  In this case, the FA algorithm accepts
the \sphinxcode{\sphinxupquote{mip\_solver}} option that specifies the mixed\sphinxhyphen{}integer programming
(MIP) solver that is used to solve the MIP that is generated by FA after
reformulating the bilevel problem.  The value of \sphinxcode{\sphinxupquote{mip\_solver}} is itself
an optimizer.  As illustrated here, this option can simply be the string
name of the MIP solver that will be used.  However, the \sphinxcode{\sphinxupquote{Solver}}
object can be used to define a MIP solver interface as well:

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{c+c1}{\PYGZsh{} Create an interface to the cbc MIP solver}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{mip} \PYG{o}{=} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{Solver}\PYG{p}{(}\PYG{l+s+s2}{\PYGZdq{}}\PYG{l+s+s2}{cbc}\PYG{l+s+s2}{\PYGZdq{}}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{c+c1}{\PYGZsh{} Create an interface to the PAO FA solver, using cbc}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{opt} \PYG{o}{=} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{Solver}\PYG{p}{(}\PYG{l+s+s2}{\PYGZdq{}}\PYG{l+s+s2}{pao.pyomo.FA}\PYG{l+s+s2}{\PYGZdq{}}\PYG{p}{,} \PYG{n}{mip\PYGZus{}solver}\PYG{o}{=}\PYG{n}{mip}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{c+c1}{\PYGZsh{} Optimize the model using cbc}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{results} \PYG{o}{=} \PYG{n}{opt}\PYG{o}{.}\PYG{n}{solve}\PYG{p}{(}\PYG{n}{M}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n+nb}{print}\PYG{p}{(}\PYG{n}{M}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{value}\PYG{p}{,} \PYG{n}{M}\PYG{o}{.}\PYG{n}{y}\PYG{o}{.}\PYG{n}{value}\PYG{p}{,} \PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{z}\PYG{o}{.}\PYG{n}{value}\PYG{p}{)}
\PYG{g+go}{6.0 4.0 2.0}
\end{sphinxVerbatim}

This enables the customization of the MIP solver used by FA.  Note that
the \sphinxcode{\sphinxupquote{solve()}} method accepts the same options as \sphinxcode{\sphinxupquote{Solve}}.
This allows for more dynamic specification of solver options:

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{c+c1}{\PYGZsh{} Create an interface to the cbc MIP solver}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{cbc} \PYG{o}{=} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{Solver}\PYG{p}{(}\PYG{l+s+s2}{\PYGZdq{}}\PYG{l+s+s2}{cbc}\PYG{l+s+s2}{\PYGZdq{}}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{c+c1}{\PYGZsh{} Create an interface to the glpk MIP solver}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{glpk} \PYG{o}{=} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{Solver}\PYG{p}{(}\PYG{l+s+s2}{\PYGZdq{}}\PYG{l+s+s2}{glpk}\PYG{l+s+s2}{\PYGZdq{}}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{c+c1}{\PYGZsh{} Create an interface to the PAO FA solver}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{opt} \PYG{o}{=} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{Solver}\PYG{p}{(}\PYG{l+s+s2}{\PYGZdq{}}\PYG{l+s+s2}{pao.pyomo.FA}\PYG{l+s+s2}{\PYGZdq{}}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{c+c1}{\PYGZsh{} Optimize the model using cbc}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{results} \PYG{o}{=} \PYG{n}{opt}\PYG{o}{.}\PYG{n}{solve}\PYG{p}{(}\PYG{n}{M}\PYG{p}{,} \PYG{n}{mip\PYGZus{}solver}\PYG{o}{=}\PYG{n}{cbc}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n+nb}{print}\PYG{p}{(}\PYG{n}{M}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{value}\PYG{p}{,} \PYG{n}{M}\PYG{o}{.}\PYG{n}{y}\PYG{o}{.}\PYG{n}{value}\PYG{p}{,} \PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{z}\PYG{o}{.}\PYG{n}{value}\PYG{p}{)}
\PYG{g+go}{6.0 4.0 2.0}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{c+c1}{\PYGZsh{} Optimize the model using glpk}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{results} \PYG{o}{=} \PYG{n}{opt}\PYG{o}{.}\PYG{n}{solve}\PYG{p}{(}\PYG{n}{M}\PYG{p}{,} \PYG{n}{mip\PYGZus{}solver}\PYG{o}{=}\PYG{n}{glpk}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n+nb}{print}\PYG{p}{(}\PYG{n}{M}\PYG{o}{.}\PYG{n}{x}\PYG{o}{.}\PYG{n}{value}\PYG{p}{,} \PYG{n}{M}\PYG{o}{.}\PYG{n}{y}\PYG{o}{.}\PYG{n}{value}\PYG{p}{,} \PYG{n}{M}\PYG{o}{.}\PYG{n}{L}\PYG{o}{.}\PYG{n}{z}\PYG{o}{.}\PYG{n}{value}\PYG{p}{)}
\PYG{g+go}{6.0 4.0 2.0}
\end{sphinxVerbatim}

\begin{sphinxadmonition}{warning}{Warning:}
The \sphinxcode{\sphinxupquote{solve()}} current passes unknown keyword arguments to the
optimize used by PAO solvers, but this feature will be disabled.
\end{sphinxadmonition}

\paragraph{PAO Solvers}
\label{\detokenize{solvers:pao-solvers}}
Solvers developed in PAO have names that begin with \sphinxcode{\sphinxupquote{pao.}}.
The current set of available PAO solvers can be queried using the
\sphinxcode{\sphinxupquote{Solver}} object:

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{k}{for} \PYG{n}{name} \PYG{o+ow}{in} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{Solver}\PYG{p}{:}
\PYG{g+gp}{... }    \PYG{n+nb}{print}\PYG{p}{(}\PYG{n}{name}\PYG{p}{)}
\PYG{g+go}{pao.mpr.FA}
\PYG{g+go}{pao.mpr.MIBS}
\PYG{g+go}{pao.mpr.PCCG}
\PYG{g+go}{pao.mpr.REG}
\PYG{g+go}{pao.pyomo.FA}
\PYG{g+go}{pao.pyomo.MIBS}
\PYG{g+go}{pao.pyomo.PCCG}
\PYG{g+go}{pao.pyomo.REG}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{pao}\PYG{o}{.}\PYG{n}{Solver}\PYG{o}{.}\PYG{n}{summary}\PYG{p}{(}\PYG{p}{)}
\PYG{g+go}{pao.mpr.FA}
\PYG{g+go}{    PAO solver for Multilevel Problem Representations that define linear}
\PYG{g+go}{    bilevel problems.  Solver uses big\PYGZhy{}M relaxations discussed by Fortuny\PYGZhy{}}
\PYG{g+go}{    Amat and McCarl (1981).}

\PYG{g+go}{pao.mpr.MIBS}
\PYG{g+go}{    PAO solver for Multilevel Problem Representations using the COIN\PYGZhy{}OR}
\PYG{g+go}{    MibS solver by Tahernejad, Ralphs, and DeNegre (2020).}

\PYG{g+go}{pao.mpr.PCCG}
\PYG{g+go}{    PAO solver for Multilevel Problem Representations that define linear}
\PYG{g+go}{    bilevel problems. Solver uses projected column constraint generation}
\PYG{g+go}{    algorithm described by Yue et al. (2017).}

\PYG{g+go}{pao.mpr.REG}
\PYG{g+go}{    PAO solver for Multilevel Problem Representations that define linear}
\PYG{g+go}{    bilevel problems.  Solver uses regularization discussed by Scheel and}
\PYG{g+go}{    Scholtes (2000) and Ralph and Wright (2004).}

\PYG{g+go}{pao.pyomo.FA}
\PYG{g+go}{    PAO solver for Pyomo models that define linear and bilinear bilevel}
\PYG{g+go}{    problems.  Solver uses big\PYGZhy{}M relaxations discussed by Fortuny\PYGZhy{}Amat and}
\PYG{g+go}{    McCarl (1981).}

\PYG{g+go}{pao.pyomo.MIBS}
\PYG{g+go}{    PAO solver for Multilevel Problem Representations using the COIN\PYGZhy{}OR}
\PYG{g+go}{    MibS solver by Tahernejad, Ralphs, and DeNegre (2020).}

\PYG{g+go}{pao.pyomo.PCCG}
\PYG{g+go}{    PAO solver for Pyomo models that define linear and bilinear bilevel}
\PYG{g+go}{    problems.  Solver uses projected column constraint generation}
\PYG{g+go}{    algorithm described by Yue et al. (2017)}

\PYG{g+go}{pao.pyomo.REG}
\PYG{g+go}{    PAO solver for Pyomo models that define linear and bilinear bilevel}
\PYG{g+go}{    problems.  Solver uses regularization discussed by Scheel and Scholtes}
\PYG{g+go}{    (2000) and Ralph and Wright (2004).}
\end{sphinxVerbatim}

The \sphinxcode{\sphinxupquote{solve()}} method includes documentation describing the keyword
arguments for a specific solver.  For example:

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{opt} \PYG{o}{=} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{Solver}\PYG{p}{(}\PYG{l+s+s2}{\PYGZdq{}}\PYG{l+s+s2}{pao.pyomo.FA}\PYG{l+s+s2}{\PYGZdq{}}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{help}\PYG{p}{(}\PYG{n}{opt}\PYG{o}{.}\PYG{n}{solve}\PYG{p}{)}
\PYG{g+go}{Help on method solve in module pao.pyomo.solvers.mpr\PYGZus{}solvers:}

\PYG{g+go}{solve(model, **options) method of pao.pyomo.solvers.mpr\PYGZus{}solvers.PyomoSubmodelSolver\PYGZus{}FA instance}
\PYG{g+go}{    Executes the solver and loads the solution into the model.}

\PYG{g+go}{    Parameters}
\PYG{g+go}{    \PYGZhy{}\PYGZhy{}\PYGZhy{}\PYGZhy{}\PYGZhy{}\PYGZhy{}\PYGZhy{}\PYGZhy{}\PYGZhy{}\PYGZhy{}}
\PYG{g+go}{    model}
\PYG{g+go}{        The model that is being optimized.}
\PYG{g+go}{    options}
\PYG{g+go}{        Keyword options that are used to configure the solver.}

\PYG{g+go}{    Keyword Arguments}
\PYG{g+go}{    \PYGZhy{}\PYGZhy{}\PYGZhy{}\PYGZhy{}\PYGZhy{}\PYGZhy{}\PYGZhy{}\PYGZhy{}\PYGZhy{}\PYGZhy{}\PYGZhy{}\PYGZhy{}\PYGZhy{}\PYGZhy{}\PYGZhy{}\PYGZhy{}\PYGZhy{}}
\PYG{g+go}{    tee}
\PYG{g+go}{      If True, then solver output is streamed to stdout. (default is False)}
\PYG{g+go}{    load\PYGZus{}solutions}
\PYG{g+go}{      If True, then the finale solution is loaded into the model. (default is True)}
\PYG{g+go}{    linearize\PYGZus{}bigm}
\PYG{g+go}{      The name of the big\PYGZhy{}M value used to linearize bilinear terms.  If this is not specified, then the solver will throw an error if bilinear terms exist in the model.}
\PYG{g+go}{    mip\PYGZus{}solver}
\PYG{g+go}{      The MIP solver used by FA.  (default is glpk)}

\PYG{g+go}{    Returns}
\PYG{g+go}{    \PYGZhy{}\PYGZhy{}\PYGZhy{}\PYGZhy{}\PYGZhy{}\PYGZhy{}\PYGZhy{}}
\PYG{g+go}{    Results}
\PYG{g+go}{        A summary of the optimization results.}
\end{sphinxVerbatim}

The \sphinxcode{\sphinxupquote{solve()}} method returns a results object that contains
data about the optimization process.  In particular, this object
contains information about the termination conditions for the solver.
The \sphinxcode{\sphinxupquote{check\_optimal\_termination()}} method can be used confirm that the
termination condition indicates that an optimal solution was found.  For example:

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{nlp} \PYG{o}{=} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{Solver}\PYG{p}{(}\PYG{l+s+s1}{\PYGZsq{}}\PYG{l+s+s1}{ipopt}\PYG{l+s+s1}{\PYGZsq{}}\PYG{p}{,} \PYG{n}{print\PYGZus{}level}\PYG{o}{=}\PYG{l+m+mi}{3}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{opt} \PYG{o}{=} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{Solver}\PYG{p}{(}\PYG{l+s+s1}{\PYGZsq{}}\PYG{l+s+s1}{pao.pyomo.REG}\PYG{l+s+s1}{\PYGZsq{}}\PYG{p}{,} \PYG{n}{nlp\PYGZus{}solver}\PYG{o}{=}\PYG{n}{nlp}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{results} \PYG{o}{=} \PYG{n}{opt}\PYG{o}{.}\PYG{n}{solve}\PYG{p}{(}\PYG{n}{M}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n+nb}{print}\PYG{p}{(}\PYG{n}{results}\PYG{o}{.}\PYG{n}{solver}\PYG{o}{.}\PYG{n}{termination\PYGZus{}condition}\PYG{p}{)}
\PYG{g+go}{TerminationCondition.optimal}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{results}\PYG{o}{.}\PYG{n}{check\PYGZus{}optimal\PYGZus{}termination}\PYG{p}{(}\PYG{p}{)}
\PYG{g+go}{True}
\end{sphinxVerbatim}

\paragraph{Pyomo Solvers}
\label{\detokenize{solvers:pyomo-solvers}}
The \sphinxcode{\sphinxupquote{Solver}} object also provides a convenient interface to
conventional numerical solvers.  Currently, solver objects constructed
by \sphinxcode{\sphinxupquote{Solver}} are simple wrappers around Pyomo optimization
solver objects.  This interface supports two types of solver
interfaces: (1) solvers that execute locally, and (2) solvers that execute
on remote servers.

When optimizating a \sphinxstylestrong{Pyomo} model, solver parameters can be setup
both when the solver interface is created and when a model is optimized.
For example:

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{c+c1}{\PYGZsh{} This is a nonlinear toy problem modeled with Pyomo}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{NLP} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{ConcreteModel}\PYG{p}{(}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{A} \PYG{o}{=} \PYG{n+nb}{list}\PYG{p}{(}\PYG{n+nb}{range}\PYG{p}{(}\PYG{l+m+mi}{10}\PYG{p}{)}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{NLP}\PYG{o}{.}\PYG{n}{x} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Var}\PYG{p}{(}\PYG{n}{A}\PYG{p}{,} \PYG{n}{bounds}\PYG{o}{=}\PYG{p}{(}\PYG{l+m+mi}{0}\PYG{p}{,}\PYG{k+kc}{None}\PYG{p}{)}\PYG{p}{,} \PYG{n}{initialize}\PYG{o}{=}\PYG{l+m+mi}{1}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{NLP}\PYG{o}{.}\PYG{n}{o} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Objective}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=}\PYG{n+nb}{sum}\PYG{p}{(}\PYG{n}{pe}\PYG{o}{.}\PYG{n}{sin}\PYG{p}{(}\PYG{p}{(}\PYG{n}{i}\PYG{o}{+}\PYG{l+m+mi}{1}\PYG{p}{)}\PYG{o}{*}\PYG{n}{NLP}\PYG{o}{.}\PYG{n}{x}\PYG{p}{[}\PYG{n}{i}\PYG{p}{]}\PYG{p}{)} \PYG{k}{for} \PYG{n}{i} \PYG{o+ow}{in} \PYG{n}{A}\PYG{p}{)}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{NLP}\PYG{o}{.}\PYG{n}{c} \PYG{o}{=} \PYG{n}{pe}\PYG{o}{.}\PYG{n}{Constraint}\PYG{p}{(}\PYG{n}{expr}\PYG{o}{=}\PYG{n+nb}{sum}\PYG{p}{(}\PYG{n}{NLP}\PYG{o}{.}\PYG{n}{x}\PYG{p}{[}\PYG{n}{i}\PYG{p}{]} \PYG{k}{for} \PYG{n}{i} \PYG{o+ow}{in} \PYG{n}{A}\PYG{p}{)} \PYG{o}{\PYGZgt{}}\PYG{o}{=} \PYG{l+m+mi}{1}\PYG{p}{)}

\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{nlp} \PYG{o}{=} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{Solver}\PYG{p}{(}\PYG{l+s+s1}{\PYGZsq{}}\PYG{l+s+s1}{ipopt}\PYG{l+s+s1}{\PYGZsq{}}\PYG{p}{,} \PYG{n}{print\PYGZus{}level}\PYG{o}{=}\PYG{l+m+mi}{3}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{c+c1}{\PYGZsh{} Apply ipopt with print level 3}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{results} \PYG{o}{=} \PYG{n}{nlp}\PYG{o}{.}\PYG{n}{solve}\PYG{p}{(}\PYG{n}{NLP}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{c+c1}{\PYGZsh{} Override the default print level to using 5}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{results} \PYG{o}{=} \PYG{n}{nlp}\PYG{o}{.}\PYG{n}{solve}\PYG{p}{(}\PYG{n}{NLP}\PYG{p}{,} \PYG{n}{print\PYGZus{}level}\PYG{o}{=}\PYG{l+m+mi}{5}\PYG{p}{)}
\end{sphinxVerbatim}

However, PAO users will typically setup solver parameters when the
Pyomo solver is initially created:

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{nlp} \PYG{o}{=} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{Solver}\PYG{p}{(}\PYG{l+s+s1}{\PYGZsq{}}\PYG{l+s+s1}{ipopt}\PYG{l+s+s1}{\PYGZsq{}}\PYG{p}{,} \PYG{n}{print\PYGZus{}level}\PYG{o}{=}\PYG{l+m+mi}{3}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{opt} \PYG{o}{=} \PYG{n}{pao}\PYG{o}{.}\PYG{n}{Solver}\PYG{p}{(}\PYG{l+s+s1}{\PYGZsq{}}\PYG{l+s+s1}{pao.pyomo.REG}\PYG{l+s+s1}{\PYGZsq{}}\PYG{p}{,} \PYG{n}{nlp\PYGZus{}solver}\PYG{o}{=}\PYG{n}{nlp}\PYG{p}{)}
\PYG{g+gp}{\PYGZgt{}\PYGZgt{}\PYGZgt{} }\PYG{n}{results} \PYG{o}{=} \PYG{n}{opt}\PYG{o}{.}\PYG{n}{solve}\PYG{p}{(}\PYG{n}{M}\PYG{p}{)}
\end{sphinxVerbatim}

When executing locally, the \sphinxcode{\sphinxupquote{executable}} option can be used
to explicitly specify the path to the executable that is used by this solver.
This is helpful in contexts where Pyomo is not automatically finding the \sphinxstyleemphasis{correct}
optimizer executable in a user’s shell environment.

When executing on a remote server, the \sphinxcode{\sphinxupquote{server}} is used to
specify the server that is used.  Currently, only the \sphinxcode{\sphinxupquote{neos}} server is
supported, which allows the user to perform optimization at NEOS \sphinxcite{neos}.
The NEOS server requires a user to specify a valid email address:

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYGZgt{}\PYGZgt{}\PYGZgt{} nlp = pao.Solver(\PYGZsq{}ipopt\PYGZsq{}, server=\PYGZsq{}neos\PYGZsq{}, email=\PYGZsq{}pao@gmail.com\PYGZsq{})
\PYGZgt{}\PYGZgt{}\PYGZgt{} opt = pao.Solver(\PYGZsq{}pao.pyomo.REG\PYGZsq{}, nlp\PYGZus{}solver=nlp)
\PYGZgt{}\PYGZgt{}\PYGZgt{} results = opt.solve(M)
\end{sphinxVerbatim}

\begin{sphinxadmonition}{warning}{Warning:}
There is no common reference for solver\sphinxhyphen{}specific parameters for the
solvers available in Pyomo.  These are generally documented with
solver documentation, and users should expect to contact solver
developers to learn about these.
\end{sphinxadmonition}

\section{Model Transformations}
\label{\detokenize{xfrm:model-transformations}}\label{\detokenize{xfrm:transformations}}\label{\detokenize{xfrm::doc}}
PAO includes a variety of functions that transform models, which generally
are applied as follows:

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYGZgt{}\PYGZgt{}\PYGZgt{} new\PYGZus{}model, soln = transform\PYGZus{}function(old\PYGZus{}model)
\end{sphinxVerbatim}

Here, the function \sphinxcode{\sphinxupquote{transform\_func}} generates the model \sphinxcode{\sphinxupquote{new\_model}} from the model \sphinxcode{\sphinxupquote{old\_model}}.  The
object \sphinxcode{\sphinxupquote{soln}} is used to map a solution back to the old model:

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYGZgt{}\PYGZgt{}\PYGZgt{} soln.copy(From=new\PYGZus{}model, To=old\PYGZus{}model)
\end{sphinxVerbatim}

The following transformation functions are documented and suitable for use by end\sphinxhyphen{}users:
\begin{itemize}
\item {} 
{\hyperref[\detokenize{reference/pyomo:pao.pyomo.convert.convert_pyomo2MultilevelProblem}]{\sphinxcrossref{\sphinxcode{\sphinxupquote{pao.pyomo.convert.convert\_pyomo2MultilevelProblem()}}}}}
\begin{quote}

This function generates a {\hyperref[\detokenize{reference/mpr:pao.mpr.repn.LinearMultilevelProblem}]{\sphinxcrossref{\sphinxcode{\sphinxupquote{LinearMultilevelProblem}}}}} or
{\hyperref[\detokenize{reference/mpr:pao.mpr.repn.QuadraticMultilevelProblem}]{\sphinxcrossref{\sphinxcode{\sphinxupquote{QuadraticMultilevelProblem}}}}} from a Pyomo model.  By default,
all constrints in the MPR representation are inequalities.
\end{quote}

\item {} 
{\hyperref[\detokenize{reference/mpr:pao.mpr.convert_repn.linearize_bilinear_terms}]{\sphinxcrossref{\sphinxcode{\sphinxupquote{pao.mpr.convert\_repn.linearize\_bilinear\_terms()}}}}}
\begin{quote}

This function generates a {\hyperref[\detokenize{reference/mpr:pao.mpr.repn.LinearMultilevelProblem}]{\sphinxcrossref{\sphinxcode{\sphinxupquote{LinearMultilevelProblem}}}}} from a {\hyperref[\detokenize{reference/mpr:pao.mpr.repn.QuadraticMultilevelProblem}]{\sphinxcrossref{\sphinxcode{\sphinxupquote{QuadraticMultilevelProblem}}}}}
that only contains bilinear terms.  This transformation currently is limited to
MPRs that only contain inequality constraints.
\end{quote}

\item {} 
{\hyperref[\detokenize{reference/mpr:pao.mpr.convert_repn.convert_to_standard_form}]{\sphinxcrossref{\sphinxcode{\sphinxupquote{pao.mpr.convert\_repn.convert\_to\_standard\_form()}}}}}
\begin{quote}

This function generates an equivalent linear multilevel representation for which all
variables are non\sphinxhyphen{}negative and all constraints have the same form (inequalities or equalities).
This simplifies the formulation of solvers, which typically assume a standard form
for subproblems.
\end{quote}

\end{itemize}

\section{Library Reference}
\label{\detokenize{reference:library-reference}}\label{\detokenize{reference::doc}}
The following classes and functions represent the core functionality
in PAO:

\subsection{Solver API}
\label{\detokenize{reference/solverapi:solver-api}}\label{\detokenize{reference/solverapi::doc}}\index{SolverAPI (class in pao.common.solver)@\spxentry{SolverAPI}\spxextra{class in pao.common.solver}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.SolverAPI}}\pysigline{\sphinxbfcode{\sphinxupquote{class }}\sphinxbfcode{\sphinxupquote{SolverAPI}}}
The base class for all PAO solvers.

The SolverAPI class defines a consistent API for optimization solvers.
\index{\_\_bool\_\_() (SolverAPI method)@\spxentry{\_\_bool\_\_()}\spxextra{SolverAPI method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.SolverAPI.__bool__}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{\_\_bool\_\_}}}{}{}
Raises an error because this class cannot be interpreted with a boolean.
\begin{quote}\begin{description}
\item[{Raises}] \leavevmode
\sphinxstyleliteralstrong{\sphinxupquote{RuntimeError}} \textendash{} Casting a solver to bool is not allowed.

\end{description}\end{quote}

\end{fulllineitems}

\index{\_\_enter\_\_() (SolverAPI method)@\spxentry{\_\_enter\_\_()}\spxextra{SolverAPI method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.SolverAPI.__enter__}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{\_\_enter\_\_}}}{}{}
Setup a solver and return it within a context manager.
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
Return a reference to \sphinxstylestrong{self}.

\item[{Return type}] \leavevmode
{\hyperref[\detokenize{reference/solverapi:pao.common.solver.SolverAPI}]{\sphinxcrossref{SolverAPI}}}

\end{description}\end{quote}

\end{fulllineitems}

\index{\_\_exit\_\_() (SolverAPI method)@\spxentry{\_\_exit\_\_()}\spxextra{SolverAPI method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.SolverAPI.__exit__}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{\_\_exit\_\_}}}{\emph{\DUrole{n}{t}}, \emph{\DUrole{n}{v}}, \emph{\DUrole{n}{traceback}}}{}
Cleanup the solver at the end of a context manager.

\end{fulllineitems}

\index{available() (SolverAPI method)@\spxentry{available()}\spxextra{SolverAPI method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.SolverAPI.available}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{available}}}{}{}
Returns a bool indicating if the solver can be executed.

The default behavior is to always return \sphinxtitleref{True}, but this method
can be overloaded in a subclass to support solver\sphinxhyphen{}specific logic
(e.g.  to confirm that a solver license is available).
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
This method returns True if the solver can be executed.

\item[{Return type}] \leavevmode
bool

\end{description}\end{quote}

\end{fulllineitems}

\index{is\_persistent() (SolverAPI method)@\spxentry{is\_persistent()}\spxextra{SolverAPI method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.SolverAPI.is_persistent}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{is\_persistent}}}{}{}
Returns True if the solver is persistent.

Persistent solvers maintain the model representation in memory,
which enables performance optimization when a problem is resolved
after changing initial conditions or tweaking model parameters.

The default is to return False, but this method can be overloaded
in a subclass to support solver\sphinxhyphen{}specific logic.
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
This method returns True if the solver is persistent.

\item[{Return type}] \leavevmode
bool

\end{description}\end{quote}

\end{fulllineitems}

\index{solve() (SolverAPI method)@\spxentry{solve()}\spxextra{SolverAPI method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.SolverAPI.solve}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{abstract }}\sphinxbfcode{\sphinxupquote{solve}}}{\emph{\DUrole{n}{model}}, \emph{\DUrole{o}{**}\DUrole{n}{options}}}{}
Executes the solver and loads the solution into the model.
\begin{quote}\begin{description}
\item[{Parameters}] \leavevmode\begin{itemize}
\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{model}} \textendash{} The model that is being optimized.

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{options}} \textendash{} Keyword options that are used to configure the solver.

\end{itemize}

\item[{Keyword Arguments}] \leavevmode\begin{itemize}
\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{tee}} \textendash{} If True, then solver output is streamed to stdout. (default is False)

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{load\_solutions}} \textendash{} If True, then the finale solution is loaded into the model. (default is True)

\end{itemize}

\item[{Returns}] \leavevmode
A summary of the optimization results.

\item[{Return type}] \leavevmode
{\hyperref[\detokenize{reference/solverapi:pao.common.solver.Results}]{\sphinxcrossref{Results}}}

\end{description}\end{quote}

\end{fulllineitems}

\index{valid\_license() (SolverAPI method)@\spxentry{valid\_license()}\spxextra{SolverAPI method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.SolverAPI.valid_license}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{valid\_license}}}{}{}
Returns a bool indicating if the solver has a valid license.

The default behavior is to always return \sphinxtitleref{True}, but this method
can be overloaded in a subclass to support solver\sphinxhyphen{}specific logic
(e.g.  to check the solver license).
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
This method returns True if the solver license is valid.

\item[{Return type}] \leavevmode
bool

\end{description}\end{quote}

\end{fulllineitems}

\index{version() (SolverAPI method)@\spxentry{version()}\spxextra{SolverAPI method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.SolverAPI.version}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{version}}}{}{}
Returns a tuple that describes the solver version.

The return value is a tuple of strings.  A typical format is (major, minor, patch), but this
is not required. The default behavior is to return an empty tuple.
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
The solver version.

\item[{Return type}] \leavevmode
tuple

\end{description}\end{quote}

\end{fulllineitems}

\end{fulllineitems}

\index{SolverFactory (class in pao.common.solver)@\spxentry{SolverFactory}\spxextra{class in pao.common.solver}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.SolverFactory}}\pysigline{\sphinxbfcode{\sphinxupquote{class }}\sphinxbfcode{\sphinxupquote{SolverFactory}}}
A class that manages a registry of solvers.

A solver factory manages a registry that enables
solvers to be created by name.
\index{\_\_call\_\_() (SolverFactory method)@\spxentry{\_\_call\_\_()}\spxextra{SolverFactory method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.SolverFactory.__call__}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{\_\_call\_\_}}}{\emph{\DUrole{n}{name}}, \emph{\DUrole{o}{**}\DUrole{n}{options}}}{}
Constructs the specified solver.

This method creates a class instance for the solver that is specified.
\begin{quote}\begin{description}
\item[{Parameters}] \leavevmode\begin{itemize}
\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{name}} (\sphinxstyleliteralemphasis{\sphinxupquote{str}}) \textendash{} The name of a solver

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{options}} \textendash{} Keyword options that are used to configure the solver.

\end{itemize}

\item[{Returns}] \leavevmode
A solver class instance for the solver that is specified.

\item[{Return type}] \leavevmode
{\hyperref[\detokenize{reference/solverapi:pao.common.solver.SolverAPI}]{\sphinxcrossref{SolverAPI}}}

\end{description}\end{quote}

\end{fulllineitems}

\index{\_\_iter\_\_() (SolverFactory method)@\spxentry{\_\_iter\_\_()}\spxextra{SolverFactory method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.SolverFactory.__iter__}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{\_\_iter\_\_}}}{}{}~\begin{quote}\begin{description}
\item[{Yields}] \leavevmode
\sphinxstyleemphasis{string} \textendash{} Yields the next solver name that has been registered

\end{description}\end{quote}

\end{fulllineitems}

\index{description() (SolverFactory method)@\spxentry{description()}\spxextra{SolverFactory method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.SolverFactory.description}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{description}}}{\emph{\DUrole{n}{name}}}{}
Returns the description of the specified solver.
\begin{quote}\begin{description}
\item[{Parameters}] \leavevmode
\sphinxstyleliteralstrong{\sphinxupquote{name}} (\sphinxstyleliteralemphasis{\sphinxupquote{str}}) \textendash{} The name of a solver

\item[{Returns}] \leavevmode
A short description of the specified solver

\item[{Return type}] \leavevmode
str

\end{description}\end{quote}

\end{fulllineitems}

\index{register() (SolverFactory method)@\spxentry{register()}\spxextra{SolverFactory method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.SolverFactory.register}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{register}}}{\emph{\DUrole{n}{cls}\DUrole{o}{=}\DUrole{default_value}{None}}, \emph{\DUrole{o}{*}}, \emph{\DUrole{n}{name}\DUrole{o}{=}\DUrole{default_value}{None}}, \emph{\DUrole{n}{doc}\DUrole{o}{=}\DUrole{default_value}{None}}}{}
Register a solver with the specified name.
\begin{quote}\begin{description}
\item[{Parameters}] \leavevmode\begin{itemize}
\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{cls}} \textendash{} Class type for the solver

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{name}} (\sphinxstyleliteralemphasis{\sphinxupquote{str}}) \textendash{} Unique name of the solver

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{doc}} (\sphinxstyleliteralemphasis{\sphinxupquote{str}}) \textendash{} Short description of the solver

\end{itemize}

\item[{Returns}] \leavevmode
If the \sphinxstylestrong{cls} parameter is None, then a class
decorator function
is returned that can be used to register a solver.

\item[{Return type}] \leavevmode
decorator

\end{description}\end{quote}

\end{fulllineitems}

\index{summary() (SolverFactory method)@\spxentry{summary()}\spxextra{SolverFactory method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.SolverFactory.summary}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{summary}}}{}{}
Print a summary of all solvers.

\end{fulllineitems}

\end{fulllineitems}

\index{Results (class in pao.common.solver)@\spxentry{Results}\spxextra{class in pao.common.solver}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.Results}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{class }}\sphinxbfcode{\sphinxupquote{Results}}}{\emph{\DUrole{n}{found\_feasible\_solution}\DUrole{o}{=}\DUrole{default_value}{None}}}{}
The results object.
\index{\_\_str\_\_() (Results method)@\spxentry{\_\_str\_\_()}\spxextra{Results method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.Results.__str__}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{\_\_str\_\_}}}{}{}
Generate a string summary of this results object.
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
A string summarizing the data in this object.

\item[{Return type}] \leavevmode
str

\end{description}\end{quote}

\end{fulllineitems}

\index{check\_optimal\_termination() (Results method)@\spxentry{check\_optimal\_termination()}\spxextra{Results method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.Results.check_optimal_termination}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{check\_optimal\_termination}}}{}{}
This function returns True if the termination condition for the solver
is ‘optimal’, ‘locallyOptimal’, or ‘globallyOptimal’, and the status is ‘ok’
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode

\item[{Return type}] \leavevmode
bool

\end{description}\end{quote}

\end{fulllineitems}

\index{found\_feasible\_solution() (Results method)@\spxentry{found\_feasible\_solution()}\spxextra{Results method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.Results.found_feasible_solution}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{found\_feasible\_solution}}}{}{}~\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
True if at least one feasible solution was found. False otherwise.

\item[{Return type}] \leavevmode
bool

\end{description}\end{quote}

\end{fulllineitems}

\index{load\_from() (Results method)@\spxentry{load\_from()}\spxextra{Results method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.Results.load_from}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{load\_from}}}{\emph{\DUrole{n}{model}}}{}
Load solution from the model.

When completed, this results object contains only one solution, which corresponds
to the solution in the model.

\end{fulllineitems}

\index{store\_to() (Results method)@\spxentry{store\_to()}\spxextra{Results method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.Results.store_to}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{store\_to}}}{\emph{\DUrole{n}{model}}, \emph{\DUrole{n}{i}\DUrole{o}{=}\DUrole{default_value}{0}}}{}
Store the solution in this object into the given model.

A results object may contain one or more solutions. This method
copies the \sphinxstylestrong{i}\sphinxhyphen{}th solution into the model.
\begin{quote}\begin{description}
\item[{Parameters}] \leavevmode
\sphinxstyleliteralstrong{\sphinxupquote{i}} (\sphinxstyleliteralemphasis{\sphinxupquote{int}}) \textendash{} The index of the solution copied into the model.

\end{description}\end{quote}

\end{fulllineitems}

\end{fulllineitems}

\index{TerminationCondition (class in pao.common.solver)@\spxentry{TerminationCondition}\spxextra{class in pao.common.solver}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.TerminationCondition}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{class }}\sphinxbfcode{\sphinxupquote{TerminationCondition}}}{\emph{\DUrole{n}{value}}}{}
The TerminationCondition class defines a enumeration of optimization termination conditions.
\index{error (TerminationCondition attribute)@\spxentry{error}\spxextra{TerminationCondition attribute}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.TerminationCondition.error}}\pysigline{\sphinxbfcode{\sphinxupquote{error}}\sphinxbfcode{\sphinxupquote{ = 11}}}
The solver exited due to an error

\end{fulllineitems}

\index{globallyOptimal (TerminationCondition attribute)@\spxentry{globallyOptimal}\spxextra{TerminationCondition attribute}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.TerminationCondition.globallyOptimal}}\pysigline{\sphinxbfcode{\sphinxupquote{globallyOptimal}}\sphinxbfcode{\sphinxupquote{ = 7}}}
The solver exited with a locally optimal solution

\end{fulllineitems}

\index{infeasible (TerminationCondition attribute)@\spxentry{infeasible}\spxextra{TerminationCondition attribute}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.TerminationCondition.infeasible}}\pysigline{\sphinxbfcode{\sphinxupquote{infeasible}}\sphinxbfcode{\sphinxupquote{ = 9}}}
The solver exited because the problem is infeasible

\end{fulllineitems}

\index{infeasibleOrUnbounded (TerminationCondition attribute)@\spxentry{infeasibleOrUnbounded}\spxextra{TerminationCondition attribute}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.TerminationCondition.infeasibleOrUnbounded}}\pysigline{\sphinxbfcode{\sphinxupquote{infeasibleOrUnbounded}}\sphinxbfcode{\sphinxupquote{ = 10}}}
The solver exited because the problem is either infeasible or unbounded

\end{fulllineitems}

\index{interrupted (TerminationCondition attribute)@\spxentry{interrupted}\spxextra{TerminationCondition attribute}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.TerminationCondition.interrupted}}\pysigline{\sphinxbfcode{\sphinxupquote{interrupted}}\sphinxbfcode{\sphinxupquote{ = 12}}}
The solver exited because it was interrupted

\end{fulllineitems}

\index{licensingProblems (TerminationCondition attribute)@\spxentry{licensingProblems}\spxextra{TerminationCondition attribute}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.TerminationCondition.licensingProblems}}\pysigline{\sphinxbfcode{\sphinxupquote{licensingProblems}}\sphinxbfcode{\sphinxupquote{ = 13}}}
The solver exited due to licensing problems

\end{fulllineitems}

\index{locallyOptimal (TerminationCondition attribute)@\spxentry{locallyOptimal}\spxextra{TerminationCondition attribute}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.TerminationCondition.locallyOptimal}}\pysigline{\sphinxbfcode{\sphinxupquote{locallyOptimal}}\sphinxbfcode{\sphinxupquote{ = 6}}}
The solver exited with a locally optimal solution

\end{fulllineitems}

\index{maxIterations (TerminationCondition attribute)@\spxentry{maxIterations}\spxextra{TerminationCondition attribute}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.TerminationCondition.maxIterations}}\pysigline{\sphinxbfcode{\sphinxupquote{maxIterations}}\sphinxbfcode{\sphinxupquote{ = 2}}}
The solver exited due to an iteration limit

\end{fulllineitems}

\index{maxTimeLimit (TerminationCondition attribute)@\spxentry{maxTimeLimit}\spxextra{TerminationCondition attribute}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.TerminationCondition.maxTimeLimit}}\pysigline{\sphinxbfcode{\sphinxupquote{maxTimeLimit}}\sphinxbfcode{\sphinxupquote{ = 1}}}
The solver exited due to a time limit

\end{fulllineitems}

\index{minStepLength (TerminationCondition attribute)@\spxentry{minStepLength}\spxextra{TerminationCondition attribute}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.TerminationCondition.minStepLength}}\pysigline{\sphinxbfcode{\sphinxupquote{minStepLength}}\sphinxbfcode{\sphinxupquote{ = 4}}}
The solver exited due to a minimum step length

\end{fulllineitems}

\index{objectiveLimit (TerminationCondition attribute)@\spxentry{objectiveLimit}\spxextra{TerminationCondition attribute}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.TerminationCondition.objectiveLimit}}\pysigline{\sphinxbfcode{\sphinxupquote{objectiveLimit}}\sphinxbfcode{\sphinxupquote{ = 3}}}
The solver exited due to an objective limit

\end{fulllineitems}

\index{optimal (TerminationCondition attribute)@\spxentry{optimal}\spxextra{TerminationCondition attribute}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.TerminationCondition.optimal}}\pysigline{\sphinxbfcode{\sphinxupquote{optimal}}\sphinxbfcode{\sphinxupquote{ = 5}}}
The solver exited with an optimal solution

\end{fulllineitems}

\index{unbounded (TerminationCondition attribute)@\spxentry{unbounded}\spxextra{TerminationCondition attribute}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.TerminationCondition.unbounded}}\pysigline{\sphinxbfcode{\sphinxupquote{unbounded}}\sphinxbfcode{\sphinxupquote{ = 8}}}
The solver exited because the problem is unbounded

\end{fulllineitems}

\index{unknown (TerminationCondition attribute)@\spxentry{unknown}\spxextra{TerminationCondition attribute}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/solverapi:pao.common.solver.TerminationCondition.unknown}}\pysigline{\sphinxbfcode{\sphinxupquote{unknown}}\sphinxbfcode{\sphinxupquote{ = 0}}}
unknown serves as both a default value, and it is used when no other enum member makes sense

\end{fulllineitems}

\end{fulllineitems}

\subsection{Pyomo Models}
\label{\detokenize{reference/pyomo:pyomo-models}}\label{\detokenize{reference/pyomo::doc}}

\paragraph{Pyomo Representation}
\label{\detokenize{reference/pyomo:pyomo-representation}}\index{SubModel (class in pao.pyomo.components)@\spxentry{SubModel}\spxextra{class in pao.pyomo.components}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/pyomo:pao.pyomo.components.SubModel}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{class }}\sphinxbfcode{\sphinxupquote{SubModel}}}{\emph{\DUrole{o}{*}\DUrole{n}{args}}, \emph{\DUrole{o}{**}\DUrole{n}{kwds}}}{}
This Pyomo model component defines a sub\sphinxhyphen{}model in a multi\sphinxhyphen{}level problem.

Pyomo models can include nested and parallel SubModel components to express
complex multi\sphinxhyphen{}level problems.
\index{\_\_init\_\_() (SubModel method)@\spxentry{\_\_init\_\_()}\spxextra{SubModel method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/pyomo:pao.pyomo.components.SubModel.__init__}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{\_\_init\_\_}}}{\emph{\DUrole{o}{*}\DUrole{n}{args}}, \emph{\DUrole{o}{**}\DUrole{n}{kwargs}}}{}
The constructor for SubModel components.
\begin{quote}\begin{description}
\item[{Keyword Arguments}] \leavevmode
\sphinxstyleliteralstrong{\sphinxupquote{fixed}} \textendash{} A Pyomo variable or a list of Pyomo variables that are optimized by upper\sphinxhyphen{}levels in the model.

\item[{Parameters}] \leavevmode\begin{itemize}
\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{*args}} \textendash{} Arguments passed to the SimpleBlock base class.

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{**kwargs}} \textendash{} Other keyword arguments passed to the SimpleBlock base class.

\end{itemize}

\end{description}\end{quote}

\end{fulllineitems}

\end{fulllineitems}

\paragraph{Model Transformations}
\label{\detokenize{reference/pyomo:model-transformations}}\index{convert\_pyomo2MultilevelProblem() (in module pao.pyomo.convert)@\spxentry{convert\_pyomo2MultilevelProblem()}\spxextra{in module pao.pyomo.convert}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/pyomo:pao.pyomo.convert.convert_pyomo2MultilevelProblem}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{convert\_pyomo2MultilevelProblem}}}{\emph{\DUrole{n}{model}}, \emph{\DUrole{o}{*}}, \emph{\DUrole{n}{determinism}\DUrole{o}{=}\DUrole{default_value}{1}}, \emph{\DUrole{n}{inequalities}\DUrole{o}{=}\DUrole{default_value}{True}}, \emph{\DUrole{n}{linear}\DUrole{o}{=}\DUrole{default_value}{None}}}{}
Traverse the model an generate a LinearMultilevelProblem or
QuadraticMultilevelProblem.  Generate errors if this problem cannot
be represented in this form.
\begin{quote}\begin{description}
\item[{Parameters}] \leavevmode\begin{itemize}
\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{model}} \textendash{} A Pyomo model object.

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{determinism}} (\sphinxstyleliteralemphasis{\sphinxupquote{int}}\sphinxstyleliteralemphasis{\sphinxupquote{, }}\sphinxstyleliteralemphasis{\sphinxupquote{Default: 1}}) \textendash{} 
Indicates whether the traversal of \sphinxstylestrong{model} is
ordered.  Valid values are:
\begin{itemize}
\item {} 
0 \sphinxhyphen{} Unordered traversal of \sphinxstylestrong{model}

\item {} 
1 \sphinxhyphen{} Ordered traversal of component indices in \sphinxstylestrong{model}

\item {} 
2 \sphinxhyphen{} Ordered traversal of components by name in \sphinxstylestrong{model}

\end{itemize}

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{linear}} (\sphinxstyleliteralemphasis{\sphinxupquote{bool}}) \textendash{} A flag that indicates whether the expected model representation is linear (True) or
quadratic (False).  If not specified, then no error checking is done to confirm
whether the model is linear or quadratic.

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{inequalities}} (\sphinxstyleliteralemphasis{\sphinxupquote{bool}}\sphinxstyleliteralemphasis{\sphinxupquote{, }}\sphinxstyleliteralemphasis{\sphinxupquote{Default: True}}) \textendash{} If True, then the multilevel problem object represents all
constraints as less\sphinxhyphen{}than\sphinxhyphen{}or\sphinxhyphen{}equal inequalities.  Otherwise,
the multilevel problem represents all constraints as equalities.

\end{itemize}

\item[{Returns}] \leavevmode
This object corresponds to the problem in \sphinxstylestrong{model}.

\item[{Return type}] \leavevmode
{\hyperref[\detokenize{reference/mpr:pao.mpr.repn.LinearMultilevelProblem}]{\sphinxcrossref{LinearMultilevelProblem}}} or {\hyperref[\detokenize{reference/mpr:pao.mpr.repn.QuadraticMultilevelProblem}]{\sphinxcrossref{QuadraticMultilevelProblem}}}

\end{description}\end{quote}

\end{fulllineitems}

\paragraph{PAO Solvers}
\label{\detokenize{reference/pyomo:pao-solvers}}\index{PyomoSubmodelSolver\_FA (class in pao.pyomo.solvers.mpr\_solvers)@\spxentry{PyomoSubmodelSolver\_FA}\spxextra{class in pao.pyomo.solvers.mpr\_solvers}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/pyomo:pao.pyomo.solvers.mpr_solvers.PyomoSubmodelSolver_FA}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{class }}\sphinxbfcode{\sphinxupquote{PyomoSubmodelSolver\_FA}}}{\emph{\DUrole{o}{**}\DUrole{n}{kwds}}}{}
PAO FA solver for Pyomo models: pao.pyomo.FA

This solver converts the Pyomo model to a LinearBilevelProblem and
calls the pao.mpr.FA solver.
\index{available() (PyomoSubmodelSolver\_FA method)@\spxentry{available()}\spxextra{PyomoSubmodelSolver\_FA method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/pyomo:pao.pyomo.solvers.mpr_solvers.PyomoSubmodelSolver_FA.available}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{available}}}{}{}
Returns a bool indicating if the solver can be executed.

The default behavior is to always return \sphinxtitleref{True}, but this method
can be overloaded in a subclass to support solver\sphinxhyphen{}specific logic
(e.g.  to confirm that a solver license is available).
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
This method returns True if the solver can be executed.

\item[{Return type}] \leavevmode
bool

\end{description}\end{quote}

\end{fulllineitems}

\index{is\_persistent() (PyomoSubmodelSolver\_FA method)@\spxentry{is\_persistent()}\spxextra{PyomoSubmodelSolver\_FA method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/pyomo:pao.pyomo.solvers.mpr_solvers.PyomoSubmodelSolver_FA.is_persistent}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{is\_persistent}}}{}{}
Returns True if the solver is persistent.

Persistent solvers maintain the model representation in memory,
which enables performance optimization when a problem is resolved
after changing initial conditions or tweaking model parameters.

The default is to return False, but this method can be overloaded
in a subclass to support solver\sphinxhyphen{}specific logic.
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
This method returns True if the solver is persistent.

\item[{Return type}] \leavevmode
bool

\end{description}\end{quote}

\end{fulllineitems}

\index{solve() (PyomoSubmodelSolver\_FA method)@\spxentry{solve()}\spxextra{PyomoSubmodelSolver\_FA method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/pyomo:pao.pyomo.solvers.mpr_solvers.PyomoSubmodelSolver_FA.solve}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{solve}}}{\emph{\DUrole{n}{model}}, \emph{\DUrole{o}{**}\DUrole{n}{options}}}{}
Executes the solver and loads the solution into the model.
\begin{quote}\begin{description}
\item[{Parameters}] \leavevmode\begin{itemize}
\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{model}} \textendash{} The model that is being optimized.

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{options}} \textendash{} Keyword options that are used to configure the solver.

\end{itemize}

\item[{Keyword Arguments}] \leavevmode\begin{itemize}
\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{tee}} \textendash{} If True, then solver output is streamed to stdout. (default is False)

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{load\_solutions}} \textendash{} If True, then the finale solution is loaded into the model. (default is True)

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{linearize\_bigm}} \textendash{} The name of the big\sphinxhyphen{}M value used to linearize bilinear terms.  If this is not specified, then the solver will throw an error if bilinear terms exist in the model.

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{mip\_solver}} \textendash{} The MIP solver used by FA.  (default is glpk)

\end{itemize}

\item[{Returns}] \leavevmode
A summary of the optimization results.

\item[{Return type}] \leavevmode
{\hyperref[\detokenize{reference/solverapi:pao.common.solver.Results}]{\sphinxcrossref{Results}}}

\end{description}\end{quote}

\end{fulllineitems}

\index{valid\_license() (PyomoSubmodelSolver\_FA method)@\spxentry{valid\_license()}\spxextra{PyomoSubmodelSolver\_FA method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/pyomo:pao.pyomo.solvers.mpr_solvers.PyomoSubmodelSolver_FA.valid_license}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{valid\_license}}}{}{}
Returns a bool indicating if the solver has a valid license.

The default behavior is to always return \sphinxtitleref{True}, but this method
can be overloaded in a subclass to support solver\sphinxhyphen{}specific logic
(e.g.  to check the solver license).
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
This method returns True if the solver license is valid.

\item[{Return type}] \leavevmode
bool

\end{description}\end{quote}

\end{fulllineitems}

\index{version() (PyomoSubmodelSolver\_FA method)@\spxentry{version()}\spxextra{PyomoSubmodelSolver\_FA method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/pyomo:pao.pyomo.solvers.mpr_solvers.PyomoSubmodelSolver_FA.version}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{version}}}{}{}
Returns a tuple that describes the solver version.

The return value is a tuple of strings.  A typical format is (major, minor, patch), but this
is not required. The default behavior is to return an empty tuple.
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
The solver version.

\item[{Return type}] \leavevmode
tuple

\end{description}\end{quote}

\end{fulllineitems}

\end{fulllineitems}

\index{PyomoSubmodelSolver\_MIBS (class in pao.pyomo.solvers.mpr\_solvers)@\spxentry{PyomoSubmodelSolver\_MIBS}\spxextra{class in pao.pyomo.solvers.mpr\_solvers}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/pyomo:pao.pyomo.solvers.mpr_solvers.PyomoSubmodelSolver_MIBS}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{class }}\sphinxbfcode{\sphinxupquote{PyomoSubmodelSolver\_MIBS}}}{\emph{\DUrole{o}{**}\DUrole{n}{kwds}}}{}
PAO MibS solver for Pyomo models: pao.pyomo.MIBS

This solver converts the Pyomo model to a LinearBilevelProblem and
calls the pao.mpr.MIBS solver.
\index{available() (PyomoSubmodelSolver\_MIBS method)@\spxentry{available()}\spxextra{PyomoSubmodelSolver\_MIBS method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/pyomo:pao.pyomo.solvers.mpr_solvers.PyomoSubmodelSolver_MIBS.available}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{available}}}{}{}
Returns a bool indicating if the solver can be executed.

The default behavior is to always return \sphinxtitleref{True}, but this method
can be overloaded in a subclass to support solver\sphinxhyphen{}specific logic
(e.g.  to confirm that a solver license is available).
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
This method returns True if the solver can be executed.

\item[{Return type}] \leavevmode
bool

\end{description}\end{quote}

\end{fulllineitems}

\index{is\_persistent() (PyomoSubmodelSolver\_MIBS method)@\spxentry{is\_persistent()}\spxextra{PyomoSubmodelSolver\_MIBS method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/pyomo:pao.pyomo.solvers.mpr_solvers.PyomoSubmodelSolver_MIBS.is_persistent}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{is\_persistent}}}{}{}
Returns True if the solver is persistent.

Persistent solvers maintain the model representation in memory,
which enables performance optimization when a problem is resolved
after changing initial conditions or tweaking model parameters.

The default is to return False, but this method can be overloaded
in a subclass to support solver\sphinxhyphen{}specific logic.
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
This method returns True if the solver is persistent.

\item[{Return type}] \leavevmode
bool

\end{description}\end{quote}

\end{fulllineitems}

\index{solve() (PyomoSubmodelSolver\_MIBS method)@\spxentry{solve()}\spxextra{PyomoSubmodelSolver\_MIBS method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/pyomo:pao.pyomo.solvers.mpr_solvers.PyomoSubmodelSolver_MIBS.solve}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{solve}}}{\emph{\DUrole{n}{model}}, \emph{\DUrole{o}{**}\DUrole{n}{options}}}{}
Executes the solver and loads the solution into the model.
\begin{quote}\begin{description}
\item[{Parameters}] \leavevmode\begin{itemize}
\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{model}} \textendash{} The model that is being optimized.

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{options}} \textendash{} Keyword options that are used to configure the solver.

\end{itemize}

\item[{Keyword Arguments}] \leavevmode\begin{itemize}
\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{tee}} \textendash{} If True, then solver output is streamed to stdout. (default is False)

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{load\_solutions}} \textendash{} If True, then the finale solution is loaded into the model. (default is True)

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{linearize\_bigm}} \textendash{} The name of the big\sphinxhyphen{}M value used to linearize bilinear terms.  If this is not specified, then the solver will throw an error if bilinear terms exist in the model.

\end{itemize}

\item[{Returns}] \leavevmode
A summary of the optimization results.

\item[{Return type}] \leavevmode
{\hyperref[\detokenize{reference/solverapi:pao.common.solver.Results}]{\sphinxcrossref{Results}}}

\end{description}\end{quote}

\end{fulllineitems}

\index{valid\_license() (PyomoSubmodelSolver\_MIBS method)@\spxentry{valid\_license()}\spxextra{PyomoSubmodelSolver\_MIBS method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/pyomo:pao.pyomo.solvers.mpr_solvers.PyomoSubmodelSolver_MIBS.valid_license}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{valid\_license}}}{}{}
Returns a bool indicating if the solver has a valid license.

The default behavior is to always return \sphinxtitleref{True}, but this method
can be overloaded in a subclass to support solver\sphinxhyphen{}specific logic
(e.g.  to check the solver license).
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
This method returns True if the solver license is valid.

\item[{Return type}] \leavevmode
bool

\end{description}\end{quote}

\end{fulllineitems}

\index{version() (PyomoSubmodelSolver\_MIBS method)@\spxentry{version()}\spxextra{PyomoSubmodelSolver\_MIBS method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/pyomo:pao.pyomo.solvers.mpr_solvers.PyomoSubmodelSolver_MIBS.version}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{version}}}{}{}
Returns a tuple that describes the solver version.

The return value is a tuple of strings.  A typical format is (major, minor, patch), but this
is not required. The default behavior is to return an empty tuple.
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
The solver version.

\item[{Return type}] \leavevmode
tuple

\end{description}\end{quote}

\end{fulllineitems}

\end{fulllineitems}

\index{PyomoSubmodelSolver\_PCCG (class in pao.pyomo.solvers.mpr\_solvers)@\spxentry{PyomoSubmodelSolver\_PCCG}\spxextra{class in pao.pyomo.solvers.mpr\_solvers}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/pyomo:pao.pyomo.solvers.mpr_solvers.PyomoSubmodelSolver_PCCG}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{class }}\sphinxbfcode{\sphinxupquote{PyomoSubmodelSolver\_PCCG}}}{\emph{\DUrole{o}{**}\DUrole{n}{kwds}}}{}
PAO PCCG solver for Pyomo models: pao.pyomo.PCCG

This solver converts the Pyomo model to a LinearBilevelProblem and
calls the pao.mpr.PCCG solver.
\index{available() (PyomoSubmodelSolver\_PCCG method)@\spxentry{available()}\spxextra{PyomoSubmodelSolver\_PCCG method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/pyomo:pao.pyomo.solvers.mpr_solvers.PyomoSubmodelSolver_PCCG.available}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{available}}}{}{}
Returns a bool indicating if the solver can be executed.

The default behavior is to always return \sphinxtitleref{True}, but this method
can be overloaded in a subclass to support solver\sphinxhyphen{}specific logic
(e.g.  to confirm that a solver license is available).
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
This method returns True if the solver can be executed.

\item[{Return type}] \leavevmode
bool

\end{description}\end{quote}

\end{fulllineitems}

\index{is\_persistent() (PyomoSubmodelSolver\_PCCG method)@\spxentry{is\_persistent()}\spxextra{PyomoSubmodelSolver\_PCCG method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/pyomo:pao.pyomo.solvers.mpr_solvers.PyomoSubmodelSolver_PCCG.is_persistent}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{is\_persistent}}}{}{}
Returns True if the solver is persistent.

Persistent solvers maintain the model representation in memory,
which enables performance optimization when a problem is resolved
after changing initial conditions or tweaking model parameters.

The default is to return False, but this method can be overloaded
in a subclass to support solver\sphinxhyphen{}specific logic.
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
This method returns True if the solver is persistent.

\item[{Return type}] \leavevmode
bool

\end{description}\end{quote}

\end{fulllineitems}

\index{solve() (PyomoSubmodelSolver\_PCCG method)@\spxentry{solve()}\spxextra{PyomoSubmodelSolver\_PCCG method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/pyomo:pao.pyomo.solvers.mpr_solvers.PyomoSubmodelSolver_PCCG.solve}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{solve}}}{\emph{\DUrole{n}{model}}, \emph{\DUrole{o}{**}\DUrole{n}{options}}}{}
Executes the solver and loads the solution into the model.
\begin{quote}\begin{description}
\item[{Parameters}] \leavevmode\begin{itemize}
\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{model}} \textendash{} The model that is being optimized.

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{options}} \textendash{} Keyword options that are used to configure the solver.

\end{itemize}

\item[{Keyword Arguments}] \leavevmode\begin{itemize}
\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{tee}} \textendash{} If True, then solver output is streamed to stdout. (default is False)

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{load\_solutions}} \textendash{} If True, then the finale solution is loaded into the model. (default is True)

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{linearize\_bigm}} \textendash{} The name of the big\sphinxhyphen{}M value used to linearize bilinear terms.  If this is not specified, then the solver will throw an error if bilinear terms exist in the model.

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{mip\_solver}} \textendash{} The MIP solver used by PCCG.  (default is cbc)

\end{itemize}

\item[{Returns}] \leavevmode
A summary of the optimization results.

\item[{Return type}] \leavevmode
{\hyperref[\detokenize{reference/solverapi:pao.common.solver.Results}]{\sphinxcrossref{Results}}}

\end{description}\end{quote}

\end{fulllineitems}

\index{valid\_license() (PyomoSubmodelSolver\_PCCG method)@\spxentry{valid\_license()}\spxextra{PyomoSubmodelSolver\_PCCG method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/pyomo:pao.pyomo.solvers.mpr_solvers.PyomoSubmodelSolver_PCCG.valid_license}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{valid\_license}}}{}{}
Returns a bool indicating if the solver has a valid license.

The default behavior is to always return \sphinxtitleref{True}, but this method
can be overloaded in a subclass to support solver\sphinxhyphen{}specific logic
(e.g.  to check the solver license).
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
This method returns True if the solver license is valid.

\item[{Return type}] \leavevmode
bool

\end{description}\end{quote}

\end{fulllineitems}

\index{version() (PyomoSubmodelSolver\_PCCG method)@\spxentry{version()}\spxextra{PyomoSubmodelSolver\_PCCG method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/pyomo:pao.pyomo.solvers.mpr_solvers.PyomoSubmodelSolver_PCCG.version}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{version}}}{}{}
Returns a tuple that describes the solver version.

The return value is a tuple of strings.  A typical format is (major, minor, patch), but this
is not required. The default behavior is to return an empty tuple.
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
The solver version.

\item[{Return type}] \leavevmode
tuple

\end{description}\end{quote}

\end{fulllineitems}

\end{fulllineitems}

\index{PyomoSubmodelSolver\_REG (class in pao.pyomo.solvers.mpr\_solvers)@\spxentry{PyomoSubmodelSolver\_REG}\spxextra{class in pao.pyomo.solvers.mpr\_solvers}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/pyomo:pao.pyomo.solvers.mpr_solvers.PyomoSubmodelSolver_REG}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{class }}\sphinxbfcode{\sphinxupquote{PyomoSubmodelSolver\_REG}}}{\emph{\DUrole{o}{**}\DUrole{n}{kwds}}}{}
PAO REG solver for Pyomo models: pao.pyomo.REG

This solver converts the Pyomo model to a LinearBilevelProblem and
calls the pao.mpr.REG solver.
\index{available() (PyomoSubmodelSolver\_REG method)@\spxentry{available()}\spxextra{PyomoSubmodelSolver\_REG method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/pyomo:pao.pyomo.solvers.mpr_solvers.PyomoSubmodelSolver_REG.available}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{available}}}{}{}
Returns a bool indicating if the solver can be executed.

The default behavior is to always return \sphinxtitleref{True}, but this method
can be overloaded in a subclass to support solver\sphinxhyphen{}specific logic
(e.g.  to confirm that a solver license is available).
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
This method returns True if the solver can be executed.

\item[{Return type}] \leavevmode
bool

\end{description}\end{quote}

\end{fulllineitems}

\index{is\_persistent() (PyomoSubmodelSolver\_REG method)@\spxentry{is\_persistent()}\spxextra{PyomoSubmodelSolver\_REG method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/pyomo:pao.pyomo.solvers.mpr_solvers.PyomoSubmodelSolver_REG.is_persistent}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{is\_persistent}}}{}{}
Returns True if the solver is persistent.

Persistent solvers maintain the model representation in memory,
which enables performance optimization when a problem is resolved
after changing initial conditions or tweaking model parameters.

The default is to return False, but this method can be overloaded
in a subclass to support solver\sphinxhyphen{}specific logic.
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
This method returns True if the solver is persistent.

\item[{Return type}] \leavevmode
bool

\end{description}\end{quote}

\end{fulllineitems}

\index{solve() (PyomoSubmodelSolver\_REG method)@\spxentry{solve()}\spxextra{PyomoSubmodelSolver\_REG method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/pyomo:pao.pyomo.solvers.mpr_solvers.PyomoSubmodelSolver_REG.solve}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{solve}}}{\emph{\DUrole{n}{model}}, \emph{\DUrole{o}{**}\DUrole{n}{options}}}{}
Executes the solver and loads the solution into the model.
\begin{quote}\begin{description}
\item[{Parameters}] \leavevmode\begin{itemize}
\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{model}} \textendash{} The model that is being optimized.

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{options}} \textendash{} Keyword options that are used to configure the solver.

\end{itemize}

\item[{Keyword Arguments}] \leavevmode\begin{itemize}
\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{tee}} \textendash{} If True, then solver output is streamed to stdout. (default is False)

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{load\_solutions}} \textendash{} If True, then the finale solution is loaded into the model. (default is True)

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{linearize\_bigm}} \textendash{} The name of the big\sphinxhyphen{}M value used to linearize bilinear terms.  If this is not specified, then the solver will throw an error if bilinear terms exist in the model.

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{nlp\_solver}} \textendash{} The NLP solver used by REG.  (default is ipopt)

\end{itemize}

\item[{Returns}] \leavevmode
A summary of the optimization results.

\item[{Return type}] \leavevmode
{\hyperref[\detokenize{reference/solverapi:pao.common.solver.Results}]{\sphinxcrossref{Results}}}

\end{description}\end{quote}

\end{fulllineitems}

\index{valid\_license() (PyomoSubmodelSolver\_REG method)@\spxentry{valid\_license()}\spxextra{PyomoSubmodelSolver\_REG method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/pyomo:pao.pyomo.solvers.mpr_solvers.PyomoSubmodelSolver_REG.valid_license}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{valid\_license}}}{}{}
Returns a bool indicating if the solver has a valid license.

The default behavior is to always return \sphinxtitleref{True}, but this method
can be overloaded in a subclass to support solver\sphinxhyphen{}specific logic
(e.g.  to check the solver license).
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
This method returns True if the solver license is valid.

\item[{Return type}] \leavevmode
bool

\end{description}\end{quote}

\end{fulllineitems}

\index{version() (PyomoSubmodelSolver\_REG method)@\spxentry{version()}\spxextra{PyomoSubmodelSolver\_REG method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/pyomo:pao.pyomo.solvers.mpr_solvers.PyomoSubmodelSolver_REG.version}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{version}}}{}{}
Returns a tuple that describes the solver version.

The return value is a tuple of strings.  A typical format is (major, minor, patch), but this
is not required. The default behavior is to return an empty tuple.
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
The solver version.

\item[{Return type}] \leavevmode
tuple

\end{description}\end{quote}

\end{fulllineitems}

\end{fulllineitems}

\subsection{Compact Models}
\label{\detokenize{reference/mpr:compact-models}}\label{\detokenize{reference/mpr::doc}}

\paragraph{LinearMultilevelProblem Representation}
\label{\detokenize{reference/mpr:linearmultilevelproblem-representation}}\index{LinearMultilevelProblem (class in pao.mpr.repn)@\spxentry{LinearMultilevelProblem}\spxextra{class in pao.mpr.repn}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/mpr:pao.mpr.repn.LinearMultilevelProblem}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{class }}\sphinxbfcode{\sphinxupquote{LinearMultilevelProblem}}}{\emph{\DUrole{n}{name}\DUrole{o}{=}\DUrole{default_value}{None}}}{}~
\begin{sphinxVerbatim}[commandchars=\\\{\}]
For bilevel problems, let:

  U   = LinearMultilevelProblem.U
  L   = U.L
  x   = [U.x, L.x]\PYGZsq{}       \PYGZsh{} dense column vector
  U.c = [U.c[U], U.c[L]]\PYGZsq{} \PYGZsh{} dense column vector
  L.c = [L.c[U], L.c[L]]\PYGZsq{} \PYGZsh{} dense column vector
  U.A = [U.A[U], U.A[L]]  \PYGZsh{} sparse matrix
  L.A = [L.A[U], L.A[L]]  \PYGZsh{} sparse matrix

Then we have:

  min\PYGZus{}\PYGZob{}U.x\PYGZcb{}   U.c\PYGZsq{} * x + U.d
  s.t.        U.A  * x       \PYGZlt{}= U.b                  \PYGZsh{} Or ==

          where L.x satisifies

              min\PYGZus{}\PYGZob{}L.x\PYGZcb{}   L.c\PYGZsq{} * x + L.d
              s.t.        L.A  * x       \PYGZlt{}= L.b      \PYGZsh{} Or ==
\end{sphinxVerbatim}

\end{fulllineitems}

\index{QuadraticMultilevelProblem (class in pao.mpr.repn)@\spxentry{QuadraticMultilevelProblem}\spxextra{class in pao.mpr.repn}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/mpr:pao.mpr.repn.QuadraticMultilevelProblem}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{class }}\sphinxbfcode{\sphinxupquote{QuadraticMultilevelProblem}}}{\emph{\DUrole{n}{name}\DUrole{o}{=}\DUrole{default_value}{None}}, \emph{\DUrole{n}{bilinear}\DUrole{o}{=}\DUrole{default_value}{False}}}{}~
\begin{sphinxVerbatim}[commandchars=\\\{\}]
For bilevel problems, let:

  U   = QuadraticMultilevelProblem.U
  L   = U.L
  x   = [U.x, L.x]\PYGZsq{}           \PYGZsh{} dense column vector
  U.c = [U.c[U], U.c[L]]\PYGZsq{}     \PYGZsh{} dense column vector
  U.P = [U.P[U,U], U.P[U,L]]  \PYGZsh{} sparse matrix
        [0,        U.P[L,L]]
  U.A = [U.A[U], U.A[L]]      \PYGZsh{} sparse matrix
  U.Q = [U.Q[U,U], U.Q[U,L]]  \PYGZsh{} sparse matrix
        [0,        U.Q[L,L]]
  L.c = [L.c[U], L.c[L]]\PYGZsq{}     \PYGZsh{} dense column vector
  L.P = [L.P[U,U], L.P[U,L]]  \PYGZsh{} sparse matrix
        [0,        L.P[L,L]]
  L.A = [L.A[U], L.A[L]]      \PYGZsh{} sparse matrix
  L.Q = [L.Q[U,U], L.Q[U,L]]  \PYGZsh{} sparse matrix
        [0,        L.Q[L,L]]

Then we have:

  min\PYGZus{}\PYGZob{}U.x\PYGZcb{}   U.c\PYGZsq{} * x + x\PYGZsq{} * U.P * x + U.d
  s.t.        U.A  * x + x\PYGZsq{} * U.Q * x       \PYGZlt{}= U.b                 \PYGZsh{} Or ==

          where L.x satisifies

              min\PYGZus{}\PYGZob{}L.x\PYGZcb{}   L.c\PYGZsq{} * x + x\PYGZsq{} * L.P * x + L.d
              s.t.        L.A  * x + x\PYGZsq{} * L.Q * x       \PYGZlt{}= L.b     \PYGZsh{} Or ==
\end{sphinxVerbatim}

\end{fulllineitems}

\paragraph{Model Transformations}
\label{\detokenize{reference/mpr:model-transformations}}\index{convert\_to\_standard\_form() (in module pao.mpr.convert\_repn)@\spxentry{convert\_to\_standard\_form()}\spxextra{in module pao.mpr.convert\_repn}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/mpr:pao.mpr.convert_repn.convert_to_standard_form}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{convert\_to\_standard\_form}}}{\emph{\DUrole{n}{M}}, \emph{\DUrole{n}{inequalities}\DUrole{o}{=}\DUrole{default_value}{False}}}{}
Normalize the LinearMultilevelProblem into a standard form.

This function copies the multilevel problem, \sphinxstylestrong{M}, and transforms
the problem such that
\begin{enumerate}
\sphinxsetlistlabels{\arabic}{enumi}{enumii}{}{.}%
\item {} 
Each real variable x is nonnegative (x \textgreater{}= 0)

\item {} 
Constraints have the specified form (e.g. all equalities or all inequalities)

\item {} 
Each level is a minimization problem

\end{enumerate}
\begin{quote}\begin{description}
\item[{Parameters}] \leavevmode\begin{itemize}
\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{M}} ({\hyperref[\detokenize{reference/mpr:pao.mpr.repn.LinearMultilevelProblem}]{\sphinxcrossref{\sphinxstyleliteralemphasis{\sphinxupquote{LinearMultilevelProblem}}}}}) \textendash{} The model that is being normalized

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{inequalities}} (\sphinxstyleliteralemphasis{\sphinxupquote{bool}}\sphinxstyleliteralemphasis{\sphinxupquote{, }}\sphinxstyleliteralemphasis{\sphinxupquote{Default: False}}) \textendash{} If this is True, then the normalized form has inequality constraints.  Otherwise, the normalized
form has equality constraints.

\end{itemize}

\item[{Returns}] \leavevmode
A normalized version of the input model

\item[{Return type}] \leavevmode
{\hyperref[\detokenize{reference/mpr:pao.mpr.repn.LinearMultilevelProblem}]{\sphinxcrossref{LinearMultilevelProblem}}}

\end{description}\end{quote}

\end{fulllineitems}

\index{linearize\_bilinear\_terms() (in module pao.mpr.convert\_repn)@\spxentry{linearize\_bilinear\_terms()}\spxextra{in module pao.mpr.convert\_repn}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/mpr:pao.mpr.convert_repn.linearize_bilinear_terms}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{linearize\_bilinear\_terms}}}{\emph{\DUrole{n}{M}}, \emph{\DUrole{n}{bigM}}}{}
Generate a linear multilevel problem from a quadratic multilevel
problem by linearizing bilinear terms.

This function copies the linear terms in the multilevel problem,
\sphinxstylestrong{M}, and replaces bilinear terms with a new variable.  This
transformation only applies when at least one of the variables in
each bilinear term is binary or integer.
\begin{quote}\begin{description}
\item[{Parameters}] \leavevmode\begin{itemize}
\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{M}} ({\hyperref[\detokenize{reference/mpr:pao.mpr.repn.QuadraticMultilevelProblem}]{\sphinxcrossref{\sphinxstyleliteralemphasis{\sphinxupquote{QuadraticMultilevelProblem}}}}}) \textendash{} The model that is being linearized

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{bigM}} (\sphinxstyleliteralemphasis{\sphinxupquote{float}}) \textendash{} The big\sphinxhyphen{}M value used to linearize nonlinear terms

\end{itemize}

\item[{Returns}] \leavevmode
A linearized version of the input model

\item[{Return type}] \leavevmode
{\hyperref[\detokenize{reference/mpr:pao.mpr.repn.LinearMultilevelProblem}]{\sphinxcrossref{LinearMultilevelProblem}}}

\end{description}\end{quote}

\end{fulllineitems}

\paragraph{PAO Solvers}
\label{\detokenize{reference/mpr:pao-solvers}}\index{LinearMultilevelSolver\_FA (class in pao.mpr.solvers.fa)@\spxentry{LinearMultilevelSolver\_FA}\spxextra{class in pao.mpr.solvers.fa}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/mpr:pao.mpr.solvers.fa.LinearMultilevelSolver_FA}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{class }}\sphinxbfcode{\sphinxupquote{LinearMultilevelSolver\_FA}}}{\emph{\DUrole{o}{**}\DUrole{n}{kwds}}}{}
PAO FA solver for linear MPRs: pao.mpr.FA

This solver replaces lower\sphinxhyphen{}level problems using the KKT conditions and
calls a MIP solver to solve the reformulated problem.
\index{available() (LinearMultilevelSolver\_FA method)@\spxentry{available()}\spxextra{LinearMultilevelSolver\_FA method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/mpr:pao.mpr.solvers.fa.LinearMultilevelSolver_FA.available}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{available}}}{}{}
Returns a bool indicating if the solver can be executed.

The default behavior is to always return \sphinxtitleref{True}, but this method
can be overloaded in a subclass to support solver\sphinxhyphen{}specific logic
(e.g.  to confirm that a solver license is available).
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
This method returns True if the solver can be executed.

\item[{Return type}] \leavevmode
bool

\end{description}\end{quote}

\end{fulllineitems}

\index{is\_persistent() (LinearMultilevelSolver\_FA method)@\spxentry{is\_persistent()}\spxextra{LinearMultilevelSolver\_FA method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/mpr:pao.mpr.solvers.fa.LinearMultilevelSolver_FA.is_persistent}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{is\_persistent}}}{}{}
Returns True if the solver is persistent.

Persistent solvers maintain the model representation in memory,
which enables performance optimization when a problem is resolved
after changing initial conditions or tweaking model parameters.

The default is to return False, but this method can be overloaded
in a subclass to support solver\sphinxhyphen{}specific logic.
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
This method returns True if the solver is persistent.

\item[{Return type}] \leavevmode
bool

\end{description}\end{quote}

\end{fulllineitems}

\index{solve() (LinearMultilevelSolver\_FA method)@\spxentry{solve()}\spxextra{LinearMultilevelSolver\_FA method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/mpr:pao.mpr.solvers.fa.LinearMultilevelSolver_FA.solve}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{solve}}}{\emph{\DUrole{n}{model}}, \emph{\DUrole{o}{**}\DUrole{n}{options}}}{}
Executes the solver and loads the solution into the model.
\begin{quote}\begin{description}
\item[{Parameters}] \leavevmode\begin{itemize}
\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{model}} \textendash{} The model that is being optimized.

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{options}} \textendash{} Keyword options that are used to configure the solver.

\end{itemize}

\item[{Keyword Arguments}] \leavevmode\begin{itemize}
\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{tee}} \textendash{} If True, then solver output is streamed to stdout. (default is False)

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{load\_solutions}} \textendash{} If True, then the finale solution is loaded into the model. (default is True)

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{mip\_solver}} \textendash{} The MIP solver used by FA.  (default is glpk)

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{bigm}} \textendash{} The big\sphinxhyphen{}M value used to enforce complementarity conditions.  (default is 1e5)

\end{itemize}

\item[{Returns}] \leavevmode
A summary of the optimization results.

\item[{Return type}] \leavevmode
{\hyperref[\detokenize{reference/solverapi:pao.common.solver.Results}]{\sphinxcrossref{Results}}}

\end{description}\end{quote}

\end{fulllineitems}

\index{valid\_license() (LinearMultilevelSolver\_FA method)@\spxentry{valid\_license()}\spxextra{LinearMultilevelSolver\_FA method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/mpr:pao.mpr.solvers.fa.LinearMultilevelSolver_FA.valid_license}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{valid\_license}}}{}{}
Returns a bool indicating if the solver has a valid license.

The default behavior is to always return \sphinxtitleref{True}, but this method
can be overloaded in a subclass to support solver\sphinxhyphen{}specific logic
(e.g.  to check the solver license).
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
This method returns True if the solver license is valid.

\item[{Return type}] \leavevmode
bool

\end{description}\end{quote}

\end{fulllineitems}

\index{version() (LinearMultilevelSolver\_FA method)@\spxentry{version()}\spxextra{LinearMultilevelSolver\_FA method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/mpr:pao.mpr.solvers.fa.LinearMultilevelSolver_FA.version}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{version}}}{}{}
Returns a tuple that describes the solver version.

The return value is a tuple of strings.  A typical format is (major, minor, patch), but this
is not required. The default behavior is to return an empty tuple.
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
The solver version.

\item[{Return type}] \leavevmode
tuple

\end{description}\end{quote}

\end{fulllineitems}

\end{fulllineitems}

\index{LinearMultilevelSolver\_MIBS (class in pao.mpr.solvers.mibs)@\spxentry{LinearMultilevelSolver\_MIBS}\spxextra{class in pao.mpr.solvers.mibs}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/mpr:pao.mpr.solvers.mibs.LinearMultilevelSolver_MIBS}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{class }}\sphinxbfcode{\sphinxupquote{LinearMultilevelSolver\_MIBS}}}{\emph{\DUrole{o}{**}\DUrole{n}{kwds}}}{}
PAO MibS solver for linear MPRs: pao.mpr.MIBS
\index{available() (LinearMultilevelSolver\_MIBS method)@\spxentry{available()}\spxextra{LinearMultilevelSolver\_MIBS method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/mpr:pao.mpr.solvers.mibs.LinearMultilevelSolver_MIBS.available}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{available}}}{}{}
Returns a bool indicating if the solver can be executed.

The default behavior is to always return \sphinxtitleref{True}, but this method
can be overloaded in a subclass to support solver\sphinxhyphen{}specific logic
(e.g.  to confirm that a solver license is available).
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
This method returns True if the solver can be executed.

\item[{Return type}] \leavevmode
bool

\end{description}\end{quote}

\end{fulllineitems}

\index{create\_mibs\_model() (LinearMultilevelSolver\_MIBS method)@\spxentry{create\_mibs\_model()}\spxextra{LinearMultilevelSolver\_MIBS method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/mpr:pao.mpr.solvers.mibs.LinearMultilevelSolver_MIBS.create_mibs_model}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{create\_mibs\_model}}}{\emph{\DUrole{n}{repn}}, \emph{\DUrole{n}{mps\_filename}}, \emph{\DUrole{n}{aux\_filename}}}{}
TODO \sphinxhyphen{} Document this transformation

\end{fulllineitems}

\index{is\_persistent() (LinearMultilevelSolver\_MIBS method)@\spxentry{is\_persistent()}\spxextra{LinearMultilevelSolver\_MIBS method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/mpr:pao.mpr.solvers.mibs.LinearMultilevelSolver_MIBS.is_persistent}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{is\_persistent}}}{}{}
Returns True if the solver is persistent.

Persistent solvers maintain the model representation in memory,
which enables performance optimization when a problem is resolved
after changing initial conditions or tweaking model parameters.

The default is to return False, but this method can be overloaded
in a subclass to support solver\sphinxhyphen{}specific logic.
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
This method returns True if the solver is persistent.

\item[{Return type}] \leavevmode
bool

\end{description}\end{quote}

\end{fulllineitems}

\index{solve() (LinearMultilevelSolver\_MIBS method)@\spxentry{solve()}\spxextra{LinearMultilevelSolver\_MIBS method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/mpr:pao.mpr.solvers.mibs.LinearMultilevelSolver_MIBS.solve}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{solve}}}{\emph{\DUrole{n}{model}}, \emph{\DUrole{o}{**}\DUrole{n}{options}}}{}
Executes the solver and loads the solution into the model.
\begin{quote}\begin{description}
\item[{Parameters}] \leavevmode\begin{itemize}
\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{model}} \textendash{} The model that is being optimized.

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{options}} \textendash{} Keyword options that are used to configure the solver.

\end{itemize}

\item[{Keyword Arguments}] \leavevmode\begin{itemize}
\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{tee}} \textendash{} If True, then solver output is streamed to stdout. (default is False)

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{load\_solutions}} \textendash{} If True, then the finale solution is loaded into the model. (default is True)

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{executable}} \textendash{} The executable used for MibS.  (default is mibs)

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{param\_file}} \textendash{} The parameter file used to configure MibS.  (default is None)

\end{itemize}

\item[{Returns}] \leavevmode
A summary of the optimization results.

\item[{Return type}] \leavevmode
{\hyperref[\detokenize{reference/solverapi:pao.common.solver.Results}]{\sphinxcrossref{Results}}}

\end{description}\end{quote}

\end{fulllineitems}

\index{valid\_license() (LinearMultilevelSolver\_MIBS method)@\spxentry{valid\_license()}\spxextra{LinearMultilevelSolver\_MIBS method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/mpr:pao.mpr.solvers.mibs.LinearMultilevelSolver_MIBS.valid_license}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{valid\_license}}}{}{}
Returns a bool indicating if the solver has a valid license.

The default behavior is to always return \sphinxtitleref{True}, but this method
can be overloaded in a subclass to support solver\sphinxhyphen{}specific logic
(e.g.  to check the solver license).
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
This method returns True if the solver license is valid.

\item[{Return type}] \leavevmode
bool

\end{description}\end{quote}

\end{fulllineitems}

\index{version() (LinearMultilevelSolver\_MIBS method)@\spxentry{version()}\spxextra{LinearMultilevelSolver\_MIBS method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/mpr:pao.mpr.solvers.mibs.LinearMultilevelSolver_MIBS.version}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{version}}}{}{}
Returns a tuple that describes the solver version.

The return value is a tuple of strings.  A typical format is (major, minor, patch), but this
is not required. The default behavior is to return an empty tuple.
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
The solver version.

\item[{Return type}] \leavevmode
tuple

\end{description}\end{quote}

\end{fulllineitems}

\end{fulllineitems}

\index{LinearMultilevelSolver\_PCCG (class in pao.mpr.solvers.pccg)@\spxentry{LinearMultilevelSolver\_PCCG}\spxextra{class in pao.mpr.solvers.pccg}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/mpr:pao.mpr.solvers.pccg.LinearMultilevelSolver_PCCG}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{class }}\sphinxbfcode{\sphinxupquote{LinearMultilevelSolver\_PCCG}}}{\emph{\DUrole{o}{**}\DUrole{n}{kwds}}}{}
PAO PCCG solver for linear MPRs: pao.mpr.PCCG

This solver iteratively adds constraints to tighten a relaxation of the lower\sphinxhyphen{}level problem.
\index{available() (LinearMultilevelSolver\_PCCG method)@\spxentry{available()}\spxextra{LinearMultilevelSolver\_PCCG method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/mpr:pao.mpr.solvers.pccg.LinearMultilevelSolver_PCCG.available}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{available}}}{}{}
Returns a bool indicating if the solver can be executed.

The default behavior is to always return \sphinxtitleref{True}, but this method
can be overloaded in a subclass to support solver\sphinxhyphen{}specific logic
(e.g.  to confirm that a solver license is available).
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
This method returns True if the solver can be executed.

\item[{Return type}] \leavevmode
bool

\end{description}\end{quote}

\end{fulllineitems}

\index{is\_persistent() (LinearMultilevelSolver\_PCCG method)@\spxentry{is\_persistent()}\spxextra{LinearMultilevelSolver\_PCCG method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/mpr:pao.mpr.solvers.pccg.LinearMultilevelSolver_PCCG.is_persistent}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{is\_persistent}}}{}{}
Returns True if the solver is persistent.

Persistent solvers maintain the model representation in memory,
which enables performance optimization when a problem is resolved
after changing initial conditions or tweaking model parameters.

The default is to return False, but this method can be overloaded
in a subclass to support solver\sphinxhyphen{}specific logic.
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
This method returns True if the solver is persistent.

\item[{Return type}] \leavevmode
bool

\end{description}\end{quote}

\end{fulllineitems}

\index{solve() (LinearMultilevelSolver\_PCCG method)@\spxentry{solve()}\spxextra{LinearMultilevelSolver\_PCCG method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/mpr:pao.mpr.solvers.pccg.LinearMultilevelSolver_PCCG.solve}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{solve}}}{\emph{\DUrole{n}{mpr}}, \emph{\DUrole{n}{options}\DUrole{o}{=}\DUrole{default_value}{None}}, \emph{\DUrole{o}{**}\DUrole{n}{config\_options}}}{}
Executes the solver and loads the solution into the model.
\begin{quote}\begin{description}
\item[{Parameters}] \leavevmode\begin{itemize}
\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{model}} \textendash{} The model that is being optimized.

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{options}} \textendash{} Keyword options that are used to configure the solver.

\end{itemize}

\item[{Keyword Arguments}] \leavevmode\begin{itemize}
\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{tee}} \textendash{} If True, then solver output is streamed to stdout. (default is False)

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{load\_solutions}} \textendash{} If True, then the finale solution is loaded into the model. (default is True)

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{mip\_solver}} \textendash{} The MIP solver used by PCCG.  (default is cbc)

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{bigm}} \textendash{} The big\sphinxhyphen{}M value used to enforce complementarity conditions.       (default is 1e6)

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{epsilon}} \textendash{} Parameter used in disjunction approximation. (default is 1e\sphinxhyphen{}4)

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{atol}} \textendash{} Convergence tolerance for |UB\sphinxhyphen{}LB|. (default is 1e\sphinxhyphen{}8)

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{rtol}} \textendash{} Convergence tolerance for |UB\sphinxhyphen{}LB|. (default is 1e\sphinxhyphen{}8)

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{maxit}} \textendash{} Maximum number of iterations. (default is None)

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{quiet}} \textendash{} If False, then enable verbose solver output. (default is True)

\end{itemize}

\item[{Returns}] \leavevmode
A summary of the optimization results.

\item[{Return type}] \leavevmode
{\hyperref[\detokenize{reference/solverapi:pao.common.solver.Results}]{\sphinxcrossref{Results}}}

\end{description}\end{quote}

\end{fulllineitems}

\index{valid\_license() (LinearMultilevelSolver\_PCCG method)@\spxentry{valid\_license()}\spxextra{LinearMultilevelSolver\_PCCG method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/mpr:pao.mpr.solvers.pccg.LinearMultilevelSolver_PCCG.valid_license}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{valid\_license}}}{}{}
Returns a bool indicating if the solver has a valid license.

The default behavior is to always return \sphinxtitleref{True}, but this method
can be overloaded in a subclass to support solver\sphinxhyphen{}specific logic
(e.g.  to check the solver license).
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
This method returns True if the solver license is valid.

\item[{Return type}] \leavevmode
bool

\end{description}\end{quote}

\end{fulllineitems}

\index{version() (LinearMultilevelSolver\_PCCG method)@\spxentry{version()}\spxextra{LinearMultilevelSolver\_PCCG method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/mpr:pao.mpr.solvers.pccg.LinearMultilevelSolver_PCCG.version}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{version}}}{}{}
Returns a tuple that describes the solver version.

The return value is a tuple of strings.  A typical format is (major, minor, patch), but this
is not required. The default behavior is to return an empty tuple.
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
The solver version.

\item[{Return type}] \leavevmode
tuple

\end{description}\end{quote}

\end{fulllineitems}

\end{fulllineitems}

\index{LinearMultilevelSolver\_REG (class in pao.mpr.solvers.reg)@\spxentry{LinearMultilevelSolver\_REG}\spxextra{class in pao.mpr.solvers.reg}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/mpr:pao.mpr.solvers.reg.LinearMultilevelSolver_REG}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{class }}\sphinxbfcode{\sphinxupquote{LinearMultilevelSolver\_REG}}}{\emph{\DUrole{o}{**}\DUrole{n}{kwds}}}{}
PAO REG solver for linear MPRs: pao.mpr.REG

This solver replaces lower\sphinxhyphen{}level problems using the KKT conditions and
calls a NLP solver to solve the reformulated problem.
\index{available() (LinearMultilevelSolver\_REG method)@\spxentry{available()}\spxextra{LinearMultilevelSolver\_REG method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/mpr:pao.mpr.solvers.reg.LinearMultilevelSolver_REG.available}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{available}}}{}{}
Returns a bool indicating if the solver can be executed.

The default behavior is to always return \sphinxtitleref{True}, but this method
can be overloaded in a subclass to support solver\sphinxhyphen{}specific logic
(e.g.  to confirm that a solver license is available).
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
This method returns True if the solver can be executed.

\item[{Return type}] \leavevmode
bool

\end{description}\end{quote}

\end{fulllineitems}

\index{is\_persistent() (LinearMultilevelSolver\_REG method)@\spxentry{is\_persistent()}\spxextra{LinearMultilevelSolver\_REG method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/mpr:pao.mpr.solvers.reg.LinearMultilevelSolver_REG.is_persistent}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{is\_persistent}}}{}{}
Returns True if the solver is persistent.

Persistent solvers maintain the model representation in memory,
which enables performance optimization when a problem is resolved
after changing initial conditions or tweaking model parameters.

The default is to return False, but this method can be overloaded
in a subclass to support solver\sphinxhyphen{}specific logic.
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
This method returns True if the solver is persistent.

\item[{Return type}] \leavevmode
bool

\end{description}\end{quote}

\end{fulllineitems}

\index{solve() (LinearMultilevelSolver\_REG method)@\spxentry{solve()}\spxextra{LinearMultilevelSolver\_REG method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/mpr:pao.mpr.solvers.reg.LinearMultilevelSolver_REG.solve}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{solve}}}{\emph{\DUrole{n}{model}}, \emph{\DUrole{o}{**}\DUrole{n}{options}}}{}
Executes the solver and loads the solution into the model.
\begin{quote}\begin{description}
\item[{Parameters}] \leavevmode\begin{itemize}
\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{model}} \textendash{} The model that is being optimized.

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{options}} \textendash{} Keyword options that are used to configure the solver.

\end{itemize}

\item[{Keyword Arguments}] \leavevmode\begin{itemize}
\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{tee}} \textendash{} If True, then solver output is streamed to stdout. (default is False)

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{load\_solutions}} \textendash{} If True, then the finale solution is loaded into the model. (default is True)

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{nlp\_solver}} \textendash{} The NLP solver used by REG.  (default is ipopt)

\item {} 
\sphinxstyleliteralstrong{\sphinxupquote{rho}} \textendash{} The tolerance for constraints that enforce complementarity conditions.  (default is 1e\sphinxhyphen{}7)

\end{itemize}

\item[{Returns}] \leavevmode
A summary of the optimization results.

\item[{Return type}] \leavevmode
{\hyperref[\detokenize{reference/solverapi:pao.common.solver.Results}]{\sphinxcrossref{Results}}}

\end{description}\end{quote}

\end{fulllineitems}

\index{valid\_license() (LinearMultilevelSolver\_REG method)@\spxentry{valid\_license()}\spxextra{LinearMultilevelSolver\_REG method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/mpr:pao.mpr.solvers.reg.LinearMultilevelSolver_REG.valid_license}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{valid\_license}}}{}{}
Returns a bool indicating if the solver has a valid license.

The default behavior is to always return \sphinxtitleref{True}, but this method
can be overloaded in a subclass to support solver\sphinxhyphen{}specific logic
(e.g.  to check the solver license).
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
This method returns True if the solver license is valid.

\item[{Return type}] \leavevmode
bool

\end{description}\end{quote}

\end{fulllineitems}

\index{version() (LinearMultilevelSolver\_REG method)@\spxentry{version()}\spxextra{LinearMultilevelSolver\_REG method}}

\begin{fulllineitems}
\phantomsection\label{\detokenize{reference/mpr:pao.mpr.solvers.reg.LinearMultilevelSolver_REG.version}}\pysiglinewithargsret{\sphinxbfcode{\sphinxupquote{version}}}{}{}
Returns a tuple that describes the solver version.

The return value is a tuple of strings.  A typical format is (major, minor, patch), but this
is not required. The default behavior is to return an empty tuple.
\begin{quote}\begin{description}
\item[{Returns}] \leavevmode
The solver version.

\item[{Return type}] \leavevmode
tuple

\end{description}\end{quote}

\end{fulllineitems}

\end{fulllineitems}

\begin{sphinxadmonition}{warning}{Warning:}
The logic in \sphinxcode{\sphinxupquote{pao.duality}} is currently disabled.  There are known errors in this code
that will be resolved by re\sphinxhyphen{}implementing it using the logic in \sphinxcode{\sphinxupquote{pao.mpr}}.
\end{sphinxadmonition}

\if 0

\section{Bibliography}
\label{\detokenize{bibliography:bibliography}}\label{\detokenize{bibliography::doc}}

\section{Indices and tables}
\label{\detokenize{index:indices-and-tables}}\begin{itemize}
\item {} 
\DUrole{xref,std,std-ref}{genindex}

\item {} 
\DUrole{xref,std,std-ref}{modindex}

\item {} 
\DUrole{xref,std,std-ref}{search}

\end{itemize}

\subsection{PAO Resources}
\label{\detokenize{index:pao-resources}}
PAO development is hosted at GitHub:
\begin{itemize}
\item {} 
\sphinxurl{https://github.com/or-fusion/pao}

\end{itemize}

The OR\sphinxhyphen{}Fusion GitHub organization is used to coordinate installation of
PAO with other OR\sphinxhyphen{}related capabilities:
\begin{itemize}
\item {} 
\sphinxurl{https://github.com/or-fusion}

\end{itemize}

Ask a question on StackOverflow:
\begin{itemize}
\item {} 
\sphinxurl{https://stackoverflow.com/questions/ask?tags=pyomo}

\end{itemize}
\fi

\begin{sphinxthebibliography}{FortunyM}

\bibitem[1]{anadalingam}
G. Anandalingam, A mathematical programming model of decentralized multi\sphinxhyphen{}level systems.
J. Oper.Res. Soc.39(11), 1021\textendash{}1033 (1988)

\bibitem[2]{bard98}
J. F. Bard, Practical bilevel optimization: Algorithms and
applications. Kluwer Academic Publishers, 1998.

\bibitem[3]{fortunymccarl}
J. Fortuny\sphinxhyphen{}Amat and B. McCarl, A representation and economic interpretation of a two\sphinxhyphen{}level programming problem,
The Journal of the Operations Research Society. 32(9), 783\sphinxhyphen{}792, 1981.

\bibitem[4]{neos}
\sphinxurl{https://neos-server.org/neos/}

\bibitem[5]{pyomobookii}
W. E. Hart, C. D. Laird,
J.\sphinxhyphen{}P. Watson, D. L. Woodruff, G. A. Hackebeil, B. L. Nicholson,
J. D. Siirola. Pyomo \sphinxhyphen{} Optimization Modeling in Python,
2nd Edition.  Springer Optimization and Its
Applications, Vol 67.  Springer, 2017.

\bibitem[6]{pyomoweb}
\sphinxurl{http://www.pyomo.org/}

\bibitem[7]{pyomogithub}
\sphinxurl{https://github.com/Pyomo/pyomo}
\end{sphinxthebibliography}

\if 0
\renewcommand{\indexname}{Index}
\printindex
\end{document}
\fi

### 2021-05-20 pao/common/solver.py
Misc updates to account for API changes in Pyomo 6.0

@@ -13,6 +13,12 @@ from pyutilib.misc import Options
 
 import pyomo.opt.parallel.manager
 import pyomo.environ as pe
+using_ScalarBlock = False
+try:
+    getattr(pe, "ScalarBlock")
+    using_ScalarBlock = True
+except:
+    pass
 from pyomo.common.config import ConfigValue, ConfigBlock, add_docstring_list
 from pyomo.neos.kestrel import kestrelAMPL
 
@@ -274,7 +280,7 @@ class PyomoSolver(SolverAPI):
         return self.solver.available(exception_flag=False)
 
     def solve(self, model, **options):
-        assert (isinstance(model, pe.Model) or isinstance(model, pe.SimpleBlock)), "The Pyomo solver '%s' cannot solve a model of type %s" % (self.name, str(type(model)))
+        assert (isinstance(model, pe.Model) or (using_ScalarBlock and isinstance(model, pe.ScalarBlock)) or (not using_ScalarBlock and isinstance(model, pe.SimpleBlock))), "The Pyomo solver '%s' cannot solve a model of type %s" % (self.name, str(type(model)))
         tmp_config = self.config()
         tmp_options = self._update_config(options, config=tmp_config, validate_options=False)
         if tmp_config['executable'] is not None:

### 2021-05-20 doc/sand/pao.tex

@@
 \begin{sphinxadmonition}{note}{Note:}
-PAO solvers cannot currently solve trilevel solvers like this,
+PAO solvers cannot currently solve trilevel problems like this,
 but an issue has been submitted to add this functionality.
@@
 
+In general, it may be difficult to determine a valid value of $M$.   Thus, this transformation may
+result in a restriction or relaxation of the original problem (depending on where the big-M values
+are introduced).
+
 \section{Multilevel Representations}
@@
 Pyomo.  The following section summarizes the solvers available in PAO,
-and describes how PAO manages solvers Section \DUrole{xref,std,std-ref}{transformations}
+and describes how PAO manages solvers. Section \DUrole{xref,std,std-ref}{transformations}
 describes model transformations in PAO.
@@
 The iterface to MibS is a prototype that has not been well\sphinxhyphen{}tested.
-This interface will be documented and finalized in an upcomgin next
+This interface will be documented and finalized in an upcoming
 release of PAO.
@@
 The \sphinxcode{\sphinxupquote{solve()}} current passes unknown keyword arguments to the
-optimize used by PAO solvers, but this feature will be disabled.
+optimizer used by PAO solvers, but this feature will be disabled.
 \end{sphinxadmonition}
@@
 {\hyperref[\detokenize{reference/mpr:pao.mpr.repn.QuadraticMultilevelProblem}]{\sphinxcrossref{\sphinxcode{\sphinxupquote{QuadraticMultilevelProblem}}}}} from a Pyomo model.  By default,
-all constrints in the MPR representation are inequalities.
+all constraints in the MPR representation are inequalities.
 \end{quote}
@@
 variables are non\sphinxhyphen{}negative and all constraints have the same form (inequalities or equalities).
-This simplifies the formulation of solvers, which typically assume a standard form
+This simplifies the implementation of solvers, which typically assume a standard form
 for subproblems.

### 2021-05-24 doc/sand/pao.tex

@@
  \end{equation*}
-This problem has has linear upper\sphinxhyphen{} and lower\sphinxhyphen{}level problems with different
+This problem has linear upper\sphinxhyphen{} and lower\sphinxhyphen{}level problems with different
 objectives in each level.  Thus, this problem can be represented in
@@
  \end{equation*}
-This problem has has linear upper\sphinxhyphen{} and lower\sphinxhyphen{}level problems with different
+This problem has linear upper\sphinxhyphen{} and lower\sphinxhyphen{}level problems with different
 objectives in each level.
@@
 We consider again the bilevel problem PAO1 \eqref{equation:sand:eq-pao1}.  This problem
-has has linear upper\sphinxhyphen{} and lower\sphinxhyphen{}level problems with different objectives
+has linear upper\sphinxhyphen{} and lower\sphinxhyphen{}level problems with different objectives
 in each level.
